---
title: "Predicted vs CiiDER readout — Clone C1 vs Control C4"
author: "XuanThangNguyen"
date: "`r Sys.Date()`"
output: pdf
---
# Goal

Compare the **predicted** TF binding sites engineered into clone **C1** (sheet `catTFRE_design`)
against the **CiiDER readout** of motifs detected in C1 versus the control clone
**C4** (sheet `CiiDER_results`). Quantify how well the design is recovered, test whether
designed TFs are preferentially enriched, and visualize the comparison.

CiiDER reports binding-site counts as ordinal bins:

| symbol | sites    | ordinal |
|--------|----------|---------|
| Na     | 0        | 0       |
| \*     | 1–10     | 1       |
| \*\*   | 10–20    | 2       |
| \*\*\* | 20–50    | 3       |
| \*\*\*\*| > 50    | 4       |

# Packages


```{r packages, include=FALSE}
# Set directory paths
root_dir <- rprojroot::find_root(rprojroot::has_dir("catTFRE_MS_FinalRound"))
analysis_dir <- file.path(root_dir, "catTFRE_MS_FinalRound")
plots_dir <- file.path(analysis_dir, "catTFRE_CiiDER")
# Create plots_dir if it doesn't exist
if (!dir.exists(plots_dir)) {dir.create(plots_dir, recursive = TRUE)}


require_packages <- c("readxl","dplyr","tidyr","stringr","purrr","ggplot2", "ggalluvial", "patchwork","scales","openxlsx")
to_install <- setdiff(require_packages, rownames(installed.packages()))
if (length(to_install)) install.packages(to_install, repos = "https://cloud.r-project.org")
invisible(lapply(require_packages, library, character.only = TRUE))
theme_set(theme_classic(base_size = 10) +
          theme(plot.title = element_text(face = "bold")))
```

# Load and clean

```{r load}
# catTFRE_design starts the table on the second row (first row is the supplemental caption)
catTFRE_design <- read_excel(file.path(analysis_dir, "catTFRE_CiiDER.xlsx"), sheet = "S1A", skip = 1) %>%
  rename_with(~ str_trim(.x)) %>%
  mutate(across(where(is.character), str_trim))

# CiiDER_results starts the table on the third row (caption + legend above)
CiiDER_results <- read_excel(file.path(analysis_dir, "catTFRE_CiiDER.xlsx"), sheet = "S1B", skip = 2) %>%
  rename_with(~ str_trim(.x)) %>%
  mutate(across(where(is.character), str_trim))

glimpse(catTFRE_design)
glimpse(CiiDER_results)
head(CiiDER_results)
```

```{r encode}
star_ord <- c("NA" = 0, "*" = 1, "**" = 2, "***" = 3, "****" = 4)
star_lab <- c("NA","*","**","***","****")

# Normalize the star strings before lookup. Handles every variation we've seen:
#   - leading / trailing / internal whitespace ("Na ", "Na\t", " * ")
#   - any casing of NA literal (Na, NA, nA, na) and true R NA
#   - empty strings, "0", and "-" are treated as NA-like
#   - factor inputs (coerced to character first)
# Anything that doesn't match one of {NA, *, **, ***, ****} after this is left untouched
normalize_star <- function(x) {
  x <- as.character(x)                               # factor / logical → character
  x <- str_replace_all(x, "\\s+", "")                # strip ALL whitespace
  na_like <- is.na(x) |
             toupper(x) %in% c("NA","N/A","NAN") |
             x %in% c("", "0", "-", ".")
  x[na_like] <- "NA"
  x
}

CiiDER_results <- CiiDER_results %>%
  mutate(Clone_C1 = normalize_star(Clone_C1),
         Clone_C4 = normalize_star(Clone_C4),
         C1_ord   = unname(star_ord[Clone_C1]),
         C4_ord   = unname(star_ord[Clone_C4]),
         delta    = C1_ord - C4_ord)

# every value must have mapped to 0–4. If any are NA, the original string didn't match NA/*/**/***/**** will be stop
bad <- CiiDER_results %>% filter(is.na(C1_ord) | is.na(C4_ord))
if (nrow(bad) > 0) { print(unique(c(bad$Clone_C1, bad$Clone_C4)))
  stop("Unmapped star strings found — extend normalize_star() / star_ord.")
}

# Show the mapping so you can eyeball that Na → 0 (and nothing else slipped in).
cat("C1 mapping (raw symbol → ordinal → row count):\n")
CiiDER_results %>% dplyr::count(Clone_C1, C1_ord) %>% arrange(C1_ord) %>% print()
cat("\nC4 mapping:\n")
CiiDER_results %>% dplyr::count(Clone_C4, C4_ord) %>% arrange(C4_ord) %>% print()

# Split dimer / paralog names so we can match on any component.
# Strips parentheticals: JUND(Var.2) → JUND.
# Drops all-digit TFs defensively: 
# Examples handled: "ARNT::HIF1A", "KLF16--SP3", "JUND(Var.2)",
#                   "HOXA13--HOXD13", "NKX3-2" (kept whole).
split_name <- function(x) {
  x %>%
    str_replace_all("\\(.*?\\)", "") %>%
    str_split("::|--") %>%
    map(~ toupper(str_trim(.x))) %>%
    map(~ .x[nzchar(.x) & !grepl("^[0-9]+$", .x)])
}
catTFRE_design$parts <- split_name(catTFRE_design$TF)
CiiDER_results$parts <- split_name(CiiDER_results$TF)

# Diagnostic — eyeball that every TF is a real TF and the count matches what you expect. 
designed_TFs_preview <- sort(unique(unlist(catTFRE_design$parts)))
cat(sprintf("Designed TFs: %d unique\n", length(designed_TFs_preview)))
cat("All TFs:\n")
print(designed_TFs_preview)
stopifnot("Phantom digit-only TF detected — check S1A renaming" =
          !any(grepl("^[0-9]+$", designed_TFs_preview)))

designed_TFs <- unique(unlist(catTFRE_design$parts))
# vapply avoid any tidy-evaluation surprises.
CiiDER_results$Designed <- vapply(CiiDER_results$parts,
                       function(x) any(x %in% designed_TFs),
                       logical(1))

# Quick sanity check: not everything should be flagged as designed.
stopifnot("Designed flag must vary CiiDER_results" =
          length(unique(CiiDER_results$Designed)) == 2)
```

# Per-designed-TF recovery
```{r recovery}
lookup_best <- function(query_parts, db = CiiDER_results) {
  hit <- vapply(db$parts, function(x) any(x %in% query_parts), logical(1))
  if (!any(hit))
    return(tibble(CiiDER  = FALSE,
                  C1_ord     = 0L, C4_ord = 0L, delta = 0L,
                  best_match = NA_character_))
  hits <- db[hit, , drop = FALSE]
  best <- hits[which.max(hits$C1_ord), , drop = FALSE]
  tibble(CiiDER  = TRUE,
         C1_ord     = as.integer(best$C1_ord),
         C4_ord     = as.integer(best$C4_ord),
         delta      = as.integer(best$C1_ord - best$C4_ord),
         best_match = best$TF)
}

ud <- catTFRE_design %>%
  distinct(TF, .keep_all = TRUE) %>%
  mutate(rec = map(parts, lookup_best)) %>%
  unnest_wider(rec) %>%
  mutate(C1_bin           = star_lab[C1_ord + 1L],
         C4_bin           = star_lab[C4_ord + 1L],
         Detected_C4      = C4_ord >= 1,
         Detected_C1      = C1_ord >= 1,
         Strong_C1        = C1_ord >= 3,
         Strong_C4        = C4_ord >= 3,
         `Enriched_C1>C4` = delta  > 0)

# scoping bug returning a constant for every TF).
stopifnot("C1_ord must vary across designed TFs" = length(unique(ud$C1_ord)) > 1,
          "CiiDER must vary across designed TFs" = length(unique(ud$CiiDER)) >= 1 && !all(ud$CiiDER & ud$C1_ord == max(ud$C1_ord)))

# Show how every designed TF was classified.
cat("Designed-TF C1 bin breakdown:\n")
ud %>%
  dplyr::count(C1_bin) %>%
  mutate(C1_bin = factor(C1_bin, levels = star_lab)) %>%
  arrange(C1_bin) %>%
  print()

cat("\nDetection summary:\n")
ud %>%
  summarise(n_total      = n(),
            n_CiiDER  = sum(CiiDER),
            n_detected   = sum(Detected_C4),       # C4 >= *
            n_detected   = sum(Detected_C1),       # C1 >= *
            n_strong_C4  = sum(Strong_C4),         # C1 >= ***
            n_strong_C1  = sum(Strong_C1),         # C1 >= ***
            n_enriched   = sum(`Enriched_C1>C4`)) %>%
  print()

ud %>% head(10)
```




# Statistics

```{r stats}
mw_alt <- wilcox.test(CiiDER_results$C1_ord[CiiDER_results$Designed],
                      CiiDER_results$C1_ord[!CiiDER_results$Designed],
                      alternative = "greater")

wsr <- wilcox.test(CiiDER_results$C1_ord[CiiDER_results$Designed],
                   CiiDER_results$C4_ord[CiiDER_results$Designed],
                   paired = TRUE, alternative = "greater")

TF_lookup_best <- function(tok, db = CiiDER_results) {
  hit <- vapply(db$parts, function(x) tok %in% x, logical(1))
  if (!any(hit)) return(tibble(in_CiiDER = FALSE, C1_ord = 0L, C4_ord = 0L,
                               best_match = NA_character_))
  hits <- db[hit, , drop = FALSE]
  best <- hits[which.max(hits$C1_ord), , drop = FALSE]
  tibble(in_CiiDER  = TRUE,
         C1_ord     = as.integer(best$C1_ord),
         C4_ord     = as.integer(best$C4_ord),
         best_match = best$TF)
}

ud_TF <- tibble(TF = sort(unique(unlist(catTFRE_design$parts)))) %>%
  mutate(rec = map(TF, TF_lookup_best)) %>%
  unnest_wider(rec) %>%
  mutate(C1_bin           = star_lab[C1_ord + 1L],
         C4_bin           = star_lab[C4_ord + 1L],
         Detected_C4      = C4_ord >= 1,
         Detected_C1      = C1_ord >= 1,
         Strong_C4        = C4_ord >= 3,
         Strong_C1        = C1_ord >= 3,
         `Enriched_C1>C4` = (C1_ord - C4_ord) > 0)

cat(sprintf("TF-level denominator: %d unique TFs (vs %d S1A labels)\n",
            nrow(ud_TF), nrow(ud)))

# --- Design-level (TF-level) contingency tables --------------------------
# Designed row: UNIQUE designed TFs (n=126).
# Not-designed row: CiiDER motifs whose parts share NO TF with any design.
n_des_C1_sig    <- sum(ud_TF$Detected_C1)
n_des_C1_notsig <- sum(!ud_TF$Detected_C1)
n_nd_C1_sig     <- sum(!CiiDER_results$Designed & CiiDER_results$C1_ord >= 1)
n_nd_C1_notsig  <- sum(!CiiDER_results$Designed & CiiDER_results$C1_ord == 0)
tab_C1 <- matrix(c(n_des_C1_sig,    n_des_C1_notsig,
                   n_nd_C1_sig,     n_nd_C1_notsig),
                 nrow = 2, byrow = TRUE,
                 dimnames = list(Designed = c("Designed (unique TFs)","Not designed"),
                                 C1_sig   = c("Detected (≥*)","Not detected (NA)")))
fish_C1    <- fisher.test(tab_C1, alternative = "greater")
recall_C1  <- n_des_C1_sig / (n_des_C1_sig + n_des_C1_notsig)
nd_rate_C1 <- n_nd_C1_sig / (n_nd_C1_sig + n_nd_C1_notsig)

n_des_C4_sig    <- sum(ud_TF$Detected_C4)
n_des_C4_notsig <- sum(!ud_TF$Detected_C4)
n_nd_C4_sig     <- sum(!CiiDER_results$Designed & CiiDER_results$C4_ord >= 1)
n_nd_C4_notsig  <- sum(!CiiDER_results$Designed & CiiDER_results$C4_ord == 0)
tab_C4 <- matrix(c(n_des_C4_sig,    n_des_C4_notsig,
                   n_nd_C4_sig,     n_nd_C4_notsig),
                 nrow = 2, byrow = TRUE,
                 dimnames = list(Designed = c("Designed (unique TFs)","Not designed"),
                                 C4_sig   = c("Detected (≥*)","Not detected (NA)")))
fish_C4   <- fisher.test(tab_C4, alternative = "greater")
recall_C4 <- n_des_C4_sig / (n_des_C4_sig + n_des_C4_notsig)

cat("\nTF-level contingency C1:\n"); print(tab_C1)
cat("\nTF-level contingency C4:\n"); print(tab_C4)

stats_tbl <- tibble(
  Metric = c("Designed TFS (unique, denominator)",
             "  └ S1A labels (for reference)",
             "Detected in C4 (TF best ≥*)",
             "Detected in C1 (TF best ≥*)",
             "Strong in C4 (TF best ≥***)",
             "Strong in C1 (TF best ≥***)",
             "Enriched C1 > C4 (TF-level)",
             sprintf("Recall C1 = detected / %d TFs", nrow(ud_TF)),
             sprintf("Recall C4 = detected / %d TFs", nrow(ud_TF)),
             "Non-designed CiiDER motifs sig in C1 (rate)",
             "Mann-Whitney p (designed > non-designed C1)",
             "Wilcoxon paired p (designed C1 > C4)",
             "Fisher OR  C1 (TF-level)",
             "Fisher p   C1",
             "Fisher OR  C4 (TF-level)",
             "Fisher p   C4"),
  Value = c(nrow(ud_TF),
            nrow(ud),
            sum(ud_TF$Detected_C4),
            sum(ud_TF$Detected_C1),
            sum(ud_TF$Strong_C4),
            sum(ud_TF$Strong_C1),
            sum(ud_TF$`Enriched_C1>C4`),
            sprintf("%.2f", recall_C1),
            sprintf("%.2f", recall_C4),
            sprintf("%d / %d (%.1f%%)", n_nd_C1_sig,
                    n_nd_C1_sig + n_nd_C1_notsig, nd_rate_C1 * 100),
            signif(mw_alt$p.value, 3),
            signif(wsr$p.value, 3),
            sprintf("%.2f", fish_C1$estimate),
            signif(fish_C1$p.value, 3),
            sprintf("%.2f", fish_C4$estimate),
            signif(fish_C4$p.value, 3)))
stats_tbl
```

# Plots
## A. Recovery C1/C4
```{r panelA}
casc <- tibble(
  cat = factor(c("Designed","CiiDER\ntable","Detected\n(C4≥*)","Detected\n(C1≥*)",
                 "Strong\n(C4≥***)","Strong\n(C1≥***)","Enriched\nC1>C4"),
               levels = c("Designed","CiiDER\ntable","Detected\n(C4≥*)","Detected\n(C1≥*)",
                          "Strong\n(C4≥***)","Strong\n(C1≥***)","Enriched\nC1>C4")),
  n   = c(nrow(ud_TF),
          sum(ud_TF$in_CiiDER),
          sum(ud_TF$Detected_C4),
          sum(ud_TF$Detected_C1),
          sum(ud_TF$Strong_C4),
          sum(ud_TF$Strong_C1),
          sum(ud_TF$`Enriched_C1>C4`))) %>%
  mutate(pct = n / nrow(ud_TF) * 100,
         lbl = sprintf("%d\n(%.0f%%)", n, pct))

pA <- ggplot(casc, aes(cat, n, fill = cat)) +
  geom_col(color = "black", linewidth = 0.3, show.legend = FALSE) +
  geom_text(aes(label = lbl), vjust = -0.2, size = 3) +
  scale_fill_manual(values = c("#444444","#888888","#2ca02c","#2ca02c",
                               "#1f77b4","#1f77b4","#d62728")) +
  ylim(0, nrow(ud_TF) * 1.2) +
  labs(title    = "A. Recovery of designed TFs (TF-level)",
       subtitle = sprintf("Denominator = %d unique TFs (after splitting %d S1A labels' dimers)",
                          nrow(ud_TF), nrow(ud)),
       x = NULL, y = "Designed TFs (n)") +
  theme(plot.subtitle = element_text(size = 8, color = "grey25"))
ggsave(pA, file = file.path(plots_dir, "catTFRE_design_CiiDER_comparison.pdf"), height = 4, width = 4.5)
```

## B. C1 distribution: designed vs non-designed
```{r panelB}
# IMPORTANT: the Designed group is now TF-level (ud_TF, n=126). The Non-designed group stays at CiiDER row level (n=571) 
n_des    <- nrow(ud_TF)                          # 126
n_nondes <- sum(!CiiDER_results$Designed)         # 571

# Designed side: one row per unique designed TF, using its best matching
# CiiDER row's bin (already stored in ud_TF$C4_ord / C1_ord).
des_long <- bind_rows(ud_TF %>% transmute(ord = C4_ord, Clone = "C4"),
                      ud_TF %>% transmute(ord = C1_ord, Clone = "C1")) %>% mutate(Designed = TRUE)

# Non-designed side: each CiiDER motif row that shares no TF with any design.
nondes_long <- bind_rows(CiiDER_results %>% filter(!Designed) %>% transmute(ord = C4_ord, Clone = "C4"),
                         CiiDER_results %>% filter(!Designed) %>% transmute(ord = C1_ord, Clone = "C1")) %>% 
               mutate(Designed = FALSE)

distB <- bind_rows(des_long, nondes_long) %>%
  dplyr::count(Clone, Designed, ord) %>%
  group_by(Clone, Designed) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup() %>%
  mutate(group   = ifelse(Designed,
                          sprintf("Designed\n(n=%d)", n_des),
                          sprintf("Non-designed\n(n=%d)", n_nondes)),
         bin_lab = factor(star_lab[ord + 1L], levels = star_lab),
         Clone   = factor(Clone, levels = c("C4", "C1")))   # C4 left, C1 right

pB <- ggplot(distB, aes(group, pct, fill = bin_lab)) +
  geom_col(color = "white", linewidth = 0.3) +
  geom_text(aes(label = ifelse(pct > 4, n, "")),
            position = position_stack(vjust = 0.5),
            size = 2.6, color = "white") +
  facet_wrap(~ Clone, nrow = 1) +
  scale_fill_manual(values = c("#dddddd","#bcd2e8","#7ba9d6","#3a7cbe","#0d3b66"),
                    name = "Bin") +
  labs(title    = "B. CiiDER bin distribution (designed vs non-designed)",
       subtitle = sprintf("Designed = %d unique TFs · \nNon-designed = %d CiiDER motifs ·\n Mann-Whitney p = %.1e",
                          n_des, n_nondes,
                          wilcox.test(ud_TF$C1_ord,
                                      CiiDER_results$C1_ord[!CiiDER_results$Designed],
                                      alternative = "greater")$p.value),
       x = NULL, y = "% of TFs") +
  theme_classic(base_size = 10) +
  theme(strip.background = element_rect(fill = "grey92", color = NA),
        strip.text       = element_text(face = "bold", size = 10),
        plot.title       = element_text(face = "bold", size = 11),
        plot.subtitle    = element_text(size = 8.5, color = "grey25"),
        legend.position  = "right")
pB
ggsave(pB, file = file.path(plots_dir,"catTFRE_C1_C4_design_CiiDER_designed_nonDesigned.pdf"), height = 4, width = 4.5)

```

## C. C1 vs C4 slope chart for designed TFs

```{r panelC}
# === TF-level alluvial ===================================================
# Now uses ud_TF (one row per UNIQUE designed TF TF, n=126),
flow_src <- ud_TF

# Step 1: aggregate transitions so each ribbon = ONE merged stream
flow_df <- flow_src %>%
  dplyr::count(C4_bin, C1_bin, name = "n_TFs") %>%
  mutate(C4_bin    = factor(C4_bin, levels = star_lab),
         C1_bin    = factor(C1_bin, levels = star_lab),
         direction = case_when(as.integer(C1_bin) > as.integer(C4_bin) ~ "C1 > C4 (gained)",
                               as.integer(C1_bin) < as.integer(C4_bin) ~ "C1 < C4 (lost)",
                               TRUE                                    ~ "C1 = C4 (no change)"))
 
# Step 2: pre-compute stratum label positions so we can place "Na (n=50)"
# OUTSIDE the bars (avoids the n=3 collision you saw).
make_strat_labels <- function(bin_col, axis_x) {
  flow_src %>%
    dplyr::count(bin = .data[[bin_col]]) %>%
    mutate(bin = factor(bin, levels = star_lab)) %>%
    complete(bin, fill = list(n = 0)) %>%
    arrange(desc(bin)) %>%                 # **** first → bottom of plot
    mutate(cum_top = cumsum(n),
           y       = cum_top - n / 2,
           x       = axis_x) %>%
    filter(n > 0)
}
strat_lab <- bind_rows(make_strat_labels("C4_bin", 1) %>% mutate(hjust = 1.15),  # nudge left
                       make_strat_labels("C1_bin", 2) %>% mutate(hjust = -0.15))  # nudge right

 
pC <- ggplot(flow_df, aes(axis1 = C4_bin, axis2 = C1_bin, y = n_TFs)) +
  geom_alluvium(aes(fill = direction), alpha = 0.78,
                width = 0.18, knot.pos = 0.40,
                color = "white", linewidth = 0.3) +
  geom_stratum(width = 0.18, fill = "grey92", color = "grey40", linewidth = 0.4) +
  # External labels (bin + count) — placed via x/y, NOT stat="stratum"
  geom_text(data        = strat_lab,
            aes(x = x, y = y, label = sprintf("%s (n=%d)", bin, n), hjust = hjust),
            size = 2.7, inherit.aes = FALSE) +
  scale_x_continuous(breaks = c(1, 2),
                     labels = c("C4\n(control)", "C1\n(designed)"),
                     limits = c(0.45, 2.55), expand = c(0, 0)) +
  scale_fill_manual(values = c("C1 > C4 (gained)"    = "#d62728",
                               "C1 < C4 (lost)"      = "#1f77b4",
                               "C1 = C4 (no change)" = "#bdbdbd"),
                    name = NULL,
                    breaks = c("C1 > C4 (gained)",
                               "C1 = C4 (no change)",
                               "C1 < C4 (lost)")) +
  labs(title    = "C. Shift in CiiDER bin (C4 → C1)",
       subtitle = sprintf("%d / %d designed TFs gained signal  Wilcoxon p = %.1e",
                          sum(flow_src$`Enriched_C1>C4`),
                          nrow(flow_src), wsr$p.value),
       x = NULL, y = NULL) +
  theme_classic(base_size = 10) +
  theme(legend.position    = "bottom",
        legend.direction   = "vertical",
        legend.text        = element_text(size = 8),
        legend.key.size    = unit(0.45, "cm"),
        legend.margin      = margin(t = 2, b = 0),
        plot.title         = element_text(face = "bold", size = 11, margin = margin(b = 2)),
        plot.subtitle      = element_text(size = 8.5, color = "grey25", margin = margin(b = 6)),
        axis.line          = element_blank(),
        axis.ticks         = element_blank(),
        axis.text.y        = element_blank(),
        axis.text.x        = element_text(face = "bold", size = 9, margin = margin(t = 4)))
pC
# Companion transition table — backbone of the plot.
trans_tbl <- flow_src %>%
  dplyr::count(C4_bin, C1_bin, name = "n_TFs") %>%
  mutate(C4_bin = factor(C4_bin, levels = star_lab),
         C1_bin = factor(C1_bin, levels = star_lab),
         pct    = sprintf("%.1f%%", n_TFs / nrow(flow_src) * 100)) %>%
  arrange(C4_bin, C1_bin)
trans_tbl
ggsave(pC, file = file.path(plots_dir, "catTFRE_design_CiiDER_C1_C4.pdf"), height = 4, width = 5)
```


```{r}
flow_src <- ud_TF
# Step 1: aggregate transitions so each ribbon = ONE merged stream
flow_df <- flow_src %>%
  count(C4_bin, C1_bin, name = "n_TFs") %>%
  mutate(C4_bin    = factor(C4_bin, levels = star_lab),
         C1_bin    = factor(C1_bin, levels = star_lab),
         direction = case_when(as.integer(C1_bin) > as.integer(C4_bin) ~ "C1 > C4 (gained)",
                               as.integer(C1_bin) < as.integer(C4_bin) ~ "C1 < C4 (lost)",
                               TRUE                                    ~ "C1 = C4 (no change)"))

# Step 2: pre-compute stratum label positions (OUTSIDE the bars)
make_strat_labels <- function(bin_col, axis_x) {
  flow_src %>%
    count(bin = .data[[bin_col]]) %>%
    mutate(bin = factor(bin, levels = star_lab)) %>%
    complete(bin, fill = list(n = 0)) %>%
    arrange(desc(bin)) %>%                 # **** first → bottom of plot
    mutate(cum_top = cumsum(n),
           y       = cum_top - n / 2,
           x       = axis_x) %>%
    filter(n > 0)
}
strat_lab <- bind_rows(make_strat_labels("C4_bin", 1) %>% mutate(hjust = 1.15),
                       make_strat_labels("C1_bin", 2) %>% mutate(hjust = -0.15))


# Step 3: pre-compute the y-midpoint of every ribbon for per-flow labels.
strat_y_top <- function(bin_col) {
  flow_src %>%
    count(bin = .data[[bin_col]]) %>%
    mutate(bin = factor(bin, levels = star_lab)) %>%
    complete(bin, fill = list(n = 0)) %>%
    arrange(desc(bin)) %>%
    mutate(strat_top = cumsum(n))
}
sC4 <- strat_y_top("C4_bin")
sC1 <- strat_y_top("C1_bin")

left_pos <- flow_df %>%
  arrange(desc(C4_bin), C1_bin) %>%
  group_by(C4_bin) %>%
  mutate(off = cumsum(n_TFs) - n_TFs / 2) %>%
  ungroup() %>%
  left_join(sC4 %>% select(bin, strat_top), by = c("C4_bin" = "bin")) %>%
  mutate(y_left = strat_top - off) %>%
  select(C4_bin, C1_bin, y_left)

right_pos <- flow_df %>%
  arrange(desc(C1_bin), C4_bin) %>%
  group_by(C1_bin) %>%
  mutate(off = cumsum(n_TFs) - n_TFs / 2) %>%
  ungroup() %>%
  left_join(sC1 %>% select(bin, strat_top), by = c("C1_bin" = "bin")) %>%
  mutate(y_right = strat_top - off) %>%
  select(C4_bin, C1_bin, y_right)

flow_lbl <- flow_df %>%
  left_join(left_pos,  by = c("C4_bin", "C1_bin")) %>%
  left_join(right_pos, by = c("C4_bin", "C1_bin")) %>%
  mutate(y_mid = (y_left + y_right) / 2, x_mid = 1.5)

pC2 <- ggplot(flow_df,
             aes(axis1 = C4_bin, axis2 = C1_bin, y = n_TFs)) +
  geom_alluvium(aes(fill = direction), alpha = 0.78,
                width = 0.18, knot.pos = 0.40,
                color = "white", linewidth = 0.3) +
  geom_stratum(width = 0.18, fill = "grey92", color = "grey40",
               linewidth = 0.4) +
  # External labels (bin + count) — placed via x/y, NOT stat="stratum"
  geom_text(data        = strat_lab,
            aes(x = x, y = y,
                label = sprintf("%s (n=%d)", bin, n),
                hjust = hjust),
            size = 2.7, inherit.aes = FALSE) +
  # Per-ribbon count labels — what you asked for
  geom_label(data        = flow_lbl,
             aes(x = x_mid, y = y_mid, label = n_TFs),
             size          = 2.3,
             fontface      = "bold",
             color         = "grey10",
             fill          = scales::alpha("white", 0.85),
             label.padding = unit(0.08, "lines"),
             label.r       = unit(0.05, "lines"),
             label.size    = 0,
             inherit.aes   = FALSE) +
  scale_x_continuous(breaks = c(1, 2),
                     labels = c("C4\n(control)", "C1\n(designed)"),
                     limits = c(0.45, 2.55),
                     expand = c(0, 0)) +
  scale_fill_manual(values = c("C1 > C4 (gained)"    = "#d62728",
                               "C1 < C4 (lost)"      = "#1f77b4",
                               "C1 = C4 (no change)" = "#bdbdbd"),
                    name = NULL,
                    breaks = c("C1 > C4 (gained)",
                               "C1 = C4 (no change)",
                               "C1 < C4 (lost)")) +
  labs(title    = "C. Shift in CiiDER bin (C4 → C1)",
       subtitle = sprintf("%d / %d designed TFs gained signal  Wilcoxon p = %.1e",
                          sum(flow_src$`Enriched_C1>C4`),
                          nrow(flow_src), wsr$p.value),
       x = NULL, y = NULL) +
  theme_classic(base_size = 10) +
  theme(legend.position    = "bottom",
        legend.direction   = "vertical",
        legend.text        = element_text(size = 8),
        legend.key.size    = unit(0.45, "cm"),
        legend.margin      = margin(t = 2, b = 0),
        plot.title         = element_text(face = "bold", size = 11, margin = margin(b = 2)),
        plot.subtitle      = element_text(size = 8.5, color = "grey25", margin = margin(b = 6)),
        axis.line          = element_blank(),
        axis.ticks         = element_blank(),
        axis.text.y        = element_blank(),
        axis.text.x        = element_text(face = "bold", size = 9, margin = margin(t = 4)))
pC2

# Save the plot
ggsave(file.path(plots_dir, "catTFRE_design_CiiDER_C1_C4_new.pdf"), pC2, width = 3.6, height = 5.5)
```


## D. Heatmap of every designed TF
```{r panelD, fig.width=14, fig.height=4}
hm <- ud %>%
  arrange(desc(C1_ord), desc(delta)) %>%
  mutate(TF = factor(TF, levels = TF)) %>%
  dplyr::select(TF, C1_ord, C4_ord) %>%
  pivot_longer(c(C1_ord, C4_ord), names_to = "clone", values_to = "ord") %>%
  mutate(clone   = factor(ifelse(clone == "C1_ord","C1","C4"), levels = c("C1","C4")),
         bin_lab = star_lab[ord + 1L])

pD <- ggplot(hm, aes(TF, clone, fill = ord)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = bin_lab,
                color = ifelse(ord >= 2, "white", "black")), size = 1.9) +
  scale_color_identity() +
  scale_fill_gradientn(colors = c("#f7fbff","#bcd2e8","#7ba9d6","#3a7cbe","#0d3b66"),
                       breaks = 0:4, #labels = star_lab, 
                       name = "Bin") +
  labs(title = "D. CiiDER bin per designed TF (sorted by C1 strength, then Δ)",
       x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 5))
pD
ggsave(pD, file=file.path(plots_dir, "catTFRE_design_CiiDER_C1_C4_heatmap.pdf"), height = 2, width = 15) 

```

## E. Confusion matrix (Designed × CiiDER C1 significance)

```{r panelE}
mat_to_long <- function(mat, clone_label) {
  as.data.frame.table(mat, responseName = "n") %>%
    setNames(c("Designed", "Sig", "n")) %>%
    mutate(Designed = factor(Designed,
                             levels = c("Designed (unique)", "Not designed")),
           Sig      = factor(Sig,
                             levels = c("Detected (≥*)", "Not detected (NA)")),
           Clone    = clone_label)
}

plot_df <- bind_rows(mat_to_long(tab_C4, "C4"),
                     mat_to_long(tab_C1, "C1"))

# Two annotation labels per facet so the units are unambiguous.
ann_df <- bind_rows(
  tibble(Clone = "C1",
         label = sprintf("Recall = %d / %d = %.0f%%\nFisher OR = %.1f, p = %.1g",
                         n_des_C1_sig,
                         n_des_C1_sig + n_des_C1_notsig,
                         recall_C1 * 100,
                         fish_C1$estimate, fish_C1$p.value)),
  tibble(Clone = "C4",
         label = sprintf("Recall = %d / %d = %.0f%%\nFisher OR = %.1f, p = %.1g",
                         n_des_C4_sig,
                         n_des_C4_sig + n_des_C4_notsig,
                         recall_C4 * 100,
                         fish_C4$estimate, fish_C4$p.value)))

pE <- ggplot(plot_df, aes(Sig, Designed, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = format(n, big.mark = ","),
                color = ifelse(n > max(n) / 2, "white", "black")),
            size = 4.5, fontface = "bold") +
  geom_text(data = ann_df, aes(x = 1.5, y = 0.42, label = label),
            inherit.aes = FALSE, size = 2.5, color = "grey25",
            lineheight = 0.9) +
  scale_color_identity() +
  scale_fill_gradient(low = "#fff5eb", high = "darkgreen", guide = "none") +
  facet_wrap(~ Clone) +
  labs(title    = "Designed TFs vs CiiDER significance (TF-level counts)",
       subtitle = sprintf("Designed denominator = %d unique TFs; non-designed denominator = %d CiiDER motifs",
                          nrow(ud_TF), nrow(ud),
                          n_nd_C1_sig + n_nd_C1_notsig),
       x = NULL, y = NULL) +
  theme_bw(base_size = 11) +
  theme(strip.text    = element_text(face = "bold"),
        panel.grid    = element_blank(),
        plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(size = 8.5, color = "grey25"))
pE
# save the plot
ggsave(file.path(plots_dir, "catTFRE_design_CiiDER_combined.pdf"), pE, height = 3.6, width = 7.2)
```


# Export comparison workbook

```{r export}
wb <- createWorkbook()
hdr <- createStyle(fontColour = "white", fgFill = "#1F4E78",
                   textDecoration = "bold", halign = "center")
ok  <- createStyle(fgFill = "#C6EFCE")
bad <- createStyle(fgFill = "#FFC7CE")

addWorksheet(wb, "Summary"); writeData(wb, "Summary", stats_tbl, headerStyle = hdr)
setColWidths(wb, "Summary", 1:2, c(48, 22))

addWorksheet(wb, "Designed_TFs")
ud_out <- ud %>%
  dplyr::select(Designed_TF = TF, Start, End, Sequence = `Seq.`,
         CiiDER_match = best_match, CiiDER = CiiDER,
         C1_bin, C4_bin, `Delta(C1-C4_ord)` = delta,
         Detected_C4, Detected_C1, Strong_C4, Strong_C1, `Enriched_C1>C4`)
writeData(wb, "Designed_TFs", ud_out, headerStyle = hdr)
freezePane(wb, "Designed_TFs", firstRow = TRUE)
setColWidths(wb, "Designed_TFs", 1:ncol(ud_out), 16)

addWorksheet(wb, "All_CiiDER")
ci_out <- CiiDER_results %>%
  dplyr::select(TF, Clone_C1, Clone_C4, C1_ord, C4_ord, delta, Designed) %>%
  arrange(desc(Designed), desc(C1_ord), desc(delta))
writeData(wb, "All_CiiDER", ci_out, headerStyle = hdr)
freezePane(wb, "All_CiiDER", firstRow = TRUE)
setColWidths(wb, "All_CiiDER", 1:ncol(ci_out), 14)

addWorksheet(wb, "Not_Detected")
miss <- ud %>%
  dplyr::filter(!Detected_C1 & Detected_C4) %>%
  dplyr::select(Designed_TF = TF, Start, End, Sequence = `Seq.`,
         CiiDER_table = CiiDER, Best_match = best_match)
writeData(wb, "Not_Detected", miss, headerStyle = hdr)
setColWidths(wb, "Not_Detected", 1:ncol(miss), 18)
# Write the excel sheet
saveWorkbook(wb, file.path(plots_dir, "Design_vs_CiiDER_comparison.xlsx"), overwrite = TRUE)
```

# Session info
```{r session}
sessionInfo()
```
