---
title: "MS_Osteosarcoma_FinalRound_Emilydata_C1C4"
author: "XuanThangNguyen"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---
### Summary 
This script implements a simple overview of Mass Spectrometry (MS) data analysis with **DEP package**. 
The data including 3 cell lines (HOS, U2OS and MG63), with 4 biological/technical replicates, that make total 36 single run

The main steps of analysis:  
- Exploring data before the differential expression analysis  
- Importing **MS counts** into **DEP package** list    
- Performing the analysis with **DESeq2**  
- Exporting and exploring the results  
The results are exported into a text file and visualized using ggplot heatmaps and volcano-plots. 

Detailed tutorial for **DEP package** package could be accessed here:  
- https://bioconductor.org/packages/devel/bioc/vignettes/DEP/inst/doc/DEP.html

**1. Load package**

```{r setup, include=FALSE}
# Importing required packages
library("dplyr")
library("DEP")
library("DOSE")
library("openxlsx")
library("ggplot2")
library("ggrepel")
library("ggtext")
library("SummarizedExperiment")
library("ComplexHeatmap")
library("clusterProfiler")
library("org.Hs.eg.db")
library("AnnotationHub")
library("circlize")
library("data.table")
library("fgsea")
library("msigdbr")
set.seed(1234)

# Set directory paths
root_dir <- rprojroot::find_root(rprojroot::has_dir("catTFRE_MS_FinalRound"))
analysis_dir <- file.path(root_dir, "catTFRE_MS_FinalRound")
plots_dir <- file.path(analysis_dir, "DEP_Plots_C1C4")
# Create plots_dir if it doesn't exist
if (!dir.exists(plots_dir)) {dir.create(plots_dir, recursive = TRUE)}


```

**2. Import and manipulate MS data from Emily and Transcription factor list**
Data were ready generated by Emily and XXX that saved in .csv file with rowname is gene/protein list and colname is sample names. Loading by using open.xlsx package
```{r}
# 1.1 Import the data
# Load the data
# Load the data
catTFRE_MS_Round4 <- read.xlsx(file.path(analysis_dir,"catTFRE_Round4_HOS_MG63_U2OS_Thang.xlsx"), rowNames = F)
colnames(catTFRE_MS_Round4)

# Load the CRAPome data base, select top 75% or 50% detect in all database
U2OS_CRAPome <- read.xlsx(file.path(analysis_dir,"CrapDB_U2OS_StrepHA.xlsx"), rowNames = T)
U2OS_CRAPome <- U2OS_CRAPome %>% dplyr::filter(NUM_EXPT > 1.5 & AVE_SC > 3) # 3 U2OS samples

# Filter only C1 Mass Spectrmetry data
catTRFE_MS_C1C4 <- catTFRE_MS_Round4[ , grepl("C1|C4|Protein_Uniprot|Protein_Symbol|Gene_Symbol", colnames(catTFRE_MS_Round4))] %>% 
  dplyr::filter(!Gene_Symbol %in% rownames(U2OS_CRAPome))
colnames(catTRFE_MS_C1C4)
dim(catTRFE_MS_C1C4)


# Human TFs list of Transcription Factor (TFs) were generated by merging the list TFs between: 
# Human TFs list of Human Protein Atlas database: https://www.proteinatlas.org/search/transcription+factor
# Human TFs list of AnimalTFDB: # http://bioinfo.life.hust.edu.cn/AnimalTFDB/#!/download
# Human TFs list of Human Transcription Factors network 2018: # https://doi.org/10.1016/j.cell.2018.01.029
# Human TFs list of Uniprot database: https://www.uniprot.org/uniprotkb?facets=model_organism%3A9606%2Cexistence%3A2&query=transcription%20factor 
# Human TFs list of GeneCards database: https://www.genecards.org/Search/Keyword?queryString=%22human%20transcription%20factor%22
# Human Epifactor gene list: https://epifactors.autosome.org/description

# Import Transcription factor list
TFs_list <- read.xlsx(file.path(analysis_dir,"TFs_list_Cell2018.xlsx"))
dim(TFs_list)
# Human epifactor and histone gene
Epifactor_list <- read.xlsx(file.path(analysis_dir,"Epifactor_Histone_list.xlsx"))
dim(Epifactor_list)

table(rownames(U2OS_CRAPome) %in% TFs_list$Gene_Symbol)
table(rownames(U2OS_CRAPome) %in% Epifactor_list$HGNC_symbol)



```

**3. Generate the SummarizedExperiment data**

We can generate the SummarizedExperiment object from our data using DEP package. We need a list of expression data with unique gene name and protein ID, and colname of data set. Futhermore, the experimental design must contain ‘label’, ‘condition’ and ‘replicate’ columns. The ‘label’ column contains the identifiers of the different samples and they should correspond to the column names containing the assay data. The ‘condition’ and ‘replicate’ columns contain the annotation of these samples as defined by the user.

**3.1 Preparation the SummarizedExperiment data**
```{r}

# Generate a SummarizedExperiment object using an experimental design
# colnames(catTFRE_MS_Round4)
# The dataset has unique Uniprot identifiers, however those are not immediately informative. The associated gene names are informative, however these are not always unique.
# Check if any duplicated gene symbol or ID
catTRFE_MS_C1C4$Gene_Symbol %>% 
  duplicated() %>% 
  any()
# List the duplicate if TRUE
catTRFE_MS_C1C4 %>% 
   group_by(Gene_Symbol) %>% 
   summarize(frequency = n()) %>% 
   arrange(desc(frequency)) %>% 
   dplyr::filter(frequency > 1) %>% head()

# For further analysis these proteins must get unique names. Additionally, some proteins do not have an annotated gene name and for those we will use the Uniprot ID.
## Make it unique incase duplicate exist, and add two column of ID and gene for the SE  step
catTFRE_MS_unique <- make_unique(catTRFE_MS_C1C4, "Gene_Symbol", "Protein_Symbol", delim = ";")
# Check the list is unique for further steps?
catTFRE_MS_unique$name %>% duplicated() %>% any()
catTFRE_MS_unique$Gene_Symbol %>% duplicated() %>% any()
# then make list of unique gene name and protein ID for further steps
catTFRE_MS_unique$Gene.names <- catTFRE_MS_unique$Gene_Symbol
catTFRE_MS_unique$Protein.IDs <- catTFRE_MS_unique$Protein_Symbol

## Prepare the column number (should be string vector)
catTFRE_columns <- grep("C1|C4", colnames(catTRFE_MS_C1C4))

## design the experiment
# Experiment design for Total with only 1 replication will not consider to put number in replication column. If you set up with any replication with only one, make_se function will not works
experimental_design <- data.frame(label = colnames(catTRFE_MS_C1C4)[grep("C1|C4", colnames(catTRFE_MS_C1C4))] ,
                                  condition = c(rep("HOS_C1C4", 8), rep("MG63_C1C4", 8), rep("U2OS_C1C4", 8)),
                                  classification = c(rep("High aggressive", 8), rep("Low aggressive", 8),  rep("High aggressive", 8)),
                                  cell = c(rep("HOS", 8), rep("MG63", 8), rep("U2OS", 8)),
                                  replicate = paste0("rep_", c(1:8)))
# kable(experimental_design)
write.xlsx(experimental_design, file=file.path(plots_dir, "MS_experiment_design_Round4.xlsx"))

# The wrapper function performs the full analysis
#' The _make_se_ and _make_se_parse_ functions generate a `r Biocpkg("SummarizedExperiment") ` object that has a couple of specifications.
#' The assay data is log2-transformed and its rownames depict the protein names.
#' The rowData contains, amongst others, the 'name' and 'ID' columns that were generated by _make_unique_.
#' The colData contains the experimental design and thereby the sample annotation.
#' Thereby the colData includes the 'label', 'condition' and 'replicate' columns as well as a newly generated 'ID' column.
#' The log2-transformed assay data and the specified rowData and colData columns are prerequisites for the subsequent analysis steps.
catTFRE_MS <- make_se(catTFRE_MS_unique, catTFRE_columns, experimental_design)

# Let's have a look at the SummarizedExperiment object
catTFRE_MS


```


#### 3.3 Filter on missing values, Normalization, Differential enrichment analysis (DEA) based on linear models and empherical Bayes statistics
Filter options: look more at https://bioconductor.org/packages/release/bioc/vignettes/DEP/inst/doc/MissingValues.html
Filter for proteins based on stringent filtering: The dataset contains proteins which are not quantified in all replicates. Some proteins are even only quantified in a single replicate so need to filter on missing values
```{r, Filtering}
# set.seed(1234)
# Filter for proteins that are identified in all replicates of at least one condition
# data_filt <- filter_missval(catTFRE_MS, thr = 0) # Most stringent
# Filter for proteins that are identified in 4 out of 8 replicates of at least one condition
dim(catTFRE_MS)
data_filt <- filter_missval(catTFRE_MS, thr = 3)
dim(data_filt)

# Plot a barplot of the number of identified proteins per samples
# plot_numbers(catTFRE_MS)
# plot_numbers(data_filt)

# Plot intensity distributions and cumulative fraction of proteins with and without missing values
# plot_detect(data_filt)

# The data is background corrected and normalized by variance stabilizing transformation (vsn). Then the remaining missing values in the dataset need to be imputed before analyse the differential enrichment analysis (DEA)

# Normalize the data
data_norm <- normalize_vsn(data_filt)

# Visualize normalization by boxplots for all samples before and after normalization
p1 <- meanSdPlot(data_norm, plot = FALSE)
p1$gg + ggplot2::theme_bw()
p2 <- meanSdPlot(data_norm, plot = FALSE)
p2$gg + ggplot2::theme_bw()

# Visualize normalization by boxplots for all samples before and after normalization
# plot_normalization(data_filt, data_norm)

# Impute data for missing values
# Plot a heatmap of proteins with missing values
# plot_missval(data_filt)
# plot_missval(catTFRE_MS)

# Impute missing data using random draws from a Gaussian distribution centered around a minimal value (for MNAR)
data_imp <- impute(data_norm, fun = "MinProb", q = 0.01)
library(tibble)
expr_mat <- assay(data_imp) %>% as.data.frame() %>% rownames_to_column("GeneName")
write.xlsx(expr_mat, file=file.path(plots_dir, "MS_data_imputed_gene.xlsx"))


```


```{r, Filtering}
# Plot the data after normalization
# plot_numbers(data_norm)
# plot_numbers(data_imp)

# Plot intensity distributions and cumulative fraction of proteins with and without missing values
# plot_detect(data_filt)

# Protein-wise linear models combined with empirical Bayes statistics are used for the differential enrichment analysis (or differential expression analysis). The test_diff function introduced here uses limma and automatically generates the contrasts to be tested.

# Different enrichment possible comparisons of samples
# data_diff <- test_diff(data_imp, type = "all") # Test all_contrasts

# Set up manually defined comparisons
data_diff <- test_diff(data_imp, type = "manual",
                       test = c("HOS_C1C4_vs_MG63_C1C4", "U2OS_C1C4_vs_MG63_C1C4")) 

# Finally, significant proteins are defined by user-defined cutoffs using add_rejections.
# Denote significant proteins based on user defined cutoffs
dep <- add_rejections(data_diff, alpha = 0.05, lfc=1)

# How many genes are need to be imputed?
table(dep@elementMetadata@listData[["imputed"]])

# How many significant gene in MS data comparison
table(dep@elementMetadata@listData[["U2OS_C1C4_vs_MG63_C1C4_significant"]])
table(dep@elementMetadata@listData[["HOS_C1C4_vs_MG63_C1C4_significant"]])

# Prepare a volcano plot for the contrast "HOS_C1C4_vs_MG63_C1C4""
# Extract results for your contrast as data.frame

# Now you can plot using your custom function or ggplot
# Use ggplot2 to plot and customize
plot_volcano(dep, 
             contrast = "HOS_C1C4_vs_MG63_C1C4", 
             label_size = 2,
             add_names = TRUE, 
             plot = FALSE, adjusted = FALSE) -> HOS_MG63_C1C4
# Prepare a volcano plot for the contrast "U2OS_C1C4_vs_MG63_C1C4""
plot_volcano(dep, 
             contrast = "U2OS_C1C4_vs_MG63_C1C4", 
             label_size = 2, 
             add_names = TRUE, 
             plot = FALSE, adjusted = FALSE) -> U2OS_MG63_C1C4
```

**4. Visualization of the results**
The results from the previous analysis can be easily visualized by a number of functions. These visualizations assist in the determination of the optimal cutoffs to be used, highlight the most interesting samples and contrasts, and pinpoint differentially enriched/expressed proteins.

**4.1 Plot the principal components analysis and correlation**
```{r}
# plot PCA of total catTFRE MS protein
pdf(file = file.path(plots_dir, "PCA_sumC1C4_MS_OS.pdf") , height = 5, width = 6)
plot_pca(dep, x = 1, y = 2, n = 1000, point_size = 4)
dev.off()

# Correlation matrix
# Plot the Pearson correlation matrix of all protein
pdf(file = file.path(plots_dir, "Correlation_heatmap_All_MS_OS.pdf"), height = 6, width = 6)
plot_cor(dep, significant = TRUE, lower = 0, upper = 1, pal = "Reds")
dev.off()

```

**4.2 Heatmap of all significant proteins**
```{r, heatmap}
pdf(file = file.path(plots_dir,"Heatmap_OS_MS_C1C4_catTFRE-All_rownames.pdf"), height = 10) # starts writing a PDF to file
plot_heatmap(dep, 
             type = "centered", kmeans = TRUE, 
             k = 6, col_limit = 4, show_row_names = FALSE, #TRUE
             indicate = c("condition", "replicate", "classification")) -> dep_heatmap
dev.off() 

pdf(file = file.path(plots_dir,"Heatmap_OS_MS_C1C4-All_rownames.pdf"), height = 50) # starts writing a PDF to file
plot_heatmap(dep, 
             type = "centered", kmeans = TRUE, 
             k = 6, col_limit = 4, show_row_names = TRUE, #TRUE
             indicate = c("condition", "replicate", "classification"))
dev.off() 


## Extract gene from heatmap cluster of interest
library(ComplexHeatmap)
heatmap_matrix <- dep_heatmap@ht_list[["log2 Centered intensity"]]@matrix #= matrix with RNAseq
# lapply(row_cluster_list, function(x) length(x))  #check/confirm size clusters
# loop to extract genes for each cluster.
gene_cluster <- lapply(names(row_order(dep_heatmap)), function(i){
  out <- data.frame(GeneID = rownames(heatmap_matrix[row_order(dep_heatmap)[[i]],]),
                    Cluster = paste0("cluster", sep="_", i),
                    stringsAsFactors = FALSE)
  return(out)}) %>%  #pipe (forward) the output 'out' to the function rbind to create 'gene_cluster'
  do.call(rbind, .)

gene_cluster <- gene_cluster %>% mutate(TF = if_else(GeneID %in% TFs_list$Name, "YES", "NO"))
# write the gene cluster to excell file
write.xlsx(gene_cluster, file.path(plots_dir,"Heatmap_MS_OS_C1C4_gene_cluster.xlsx"))

#Plot
ggplot(gene_cluster, aes(x = Cluster, fill = TF)) +
  geom_bar(position = "stack") +
  # add counts
  geom_text(stat = "count",
            aes(label = ..count..),
            position = position_stack(vjust = 0.5),  # center inside each stacked segment
            color = "white",
            size = 3.5) +
  labs(title = "Distribution of TFs across clusters",
       x = "Cluster", y = "Number of genes") +
  scale_fill_manual(values = c("YES" = "#E53935", "NO" = "#9E9E9E"),
                    name = "TF status") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(file = file.path(plots_dir,"Heatmap_Cluster_gene_TF.pdf"), height = 4, width = 4)
# Call the gene name of cluster of interest
```


**Alternative Volcano plot based on ggplot2 package, with option integrated volcano plot**
**4.3 Volcano plots of specific contrasts**
Volcano plots can be used to visualize a specific contrast (comparison between two samples). This allows to inspect the enrichment of proteins between the two samples (x axis) and their corresponding adjusted p value (y axis). This function is based on DEP package.

```{r}
# Prepare a volcano plot for the contrast "MG63_C1C4_vs_MG63_NE""
HOS_U2OS_MG63_C1C4_result <- get_results(dep)
colnames(HOS_U2OS_MG63_C1C4_result)

# ===== Step 1: Classify expression status =====
prepare_data <- function(res, contrast, tf_list) {
  p_col <- paste0(contrast, "_p.val")
  padj_col <- paste0(contrast, "_p.adj")
  ratio_col <- paste0(contrast, "_ratio")
  # Prepare the result table
  res %>%
    mutate(protein = name,
           log2_fold_change = .data[[ratio_col]],
           p_value = .data[[p_col]],
           p_adj = .data[[padj_col]],
           TF = if_else(protein %in% tf_list, "YES", "NO"),
           Expression = case_when(p_value <= 0.05 & log2_fold_change >= 1 & TF == "YES" ~ "TF_Upregulated",
                                  p_value <= 0.05 & log2_fold_change <= -1 & TF == "YES" ~ "TF_Downregulated",
                                  p_value <= 0.05 & log2_fold_change >= 1 ~ "Upregulated",
                                  p_value <= 0.05 & log2_fold_change <= -1 ~ "Downregulated",
                                  TRUE ~ "Not_significant"))
}
# ===== Step 2: Classify TF expression status =====
prepare_labels <- function(df, genes_of_interest) {
  df %>%
    dplyr::filter(protein %in% genes_of_interest) %>%
    mutate(label_color = case_when(Expression == "TF_Upregulated" ~ "red",
                                   Expression == "TF_Downregulated" ~ "blue",
                                   Expression == "Upregulated" ~ "red",
                                   Expression == "Downregulated" ~ "blue",
                                   TRUE ~ "grey"))
}

# ===== Step 3: Plot volcano =====
plot_volcano <- function(res_df, label_df, title, group_left = "NE", group_right = "C1C4") {
  # Count DEPs (including TFs)
  num_up <- sum(res_df$Expression %in% c("Upregulated", "TF_Upregulated"))
  num_down <- sum(res_df$Expression %in% c("Downregulated", "TF_Downregulated"))
  tf_up <- sum(res_df$Expression == "TF_Upregulated")
  tf_down <- sum(res_df$Expression == "TF_Downregulated")
  subtitle_text <- paste0(
    "<span style='color:#3182bd;'>Enriched in ", group_left, ": ", num_down, "</span>",
    " | <span style='color:#fc9272;'>Enriched in ", group_right, ": ", num_up, "</span><br>",
    "<span style='color:blue;'>TF in ", group_left, ": ", tf_down, "</span>",
    " | <span style='color:red;'>TF in ", group_right, ": ", tf_up, "</span>")
  # Add color to lables
  label_sig <- label_df %>% dplyr::filter(label_color %in% c("red", "blue"))
  # Plot TF points on top by ordering
  res_df <- res_df %>%
    mutate(Expression = factor(Expression,
                               levels = c("Not_significant", "Upregulated", "Downregulated", 
                                          "TF_Upregulated", "TF_Downregulated"))) %>%
    arrange(Expression)  # TFs plotted last = on top
  # Plot the volcano
  ggplot(res_df, aes(x = log2_fold_change, y = -log10(p_value))) +
    geom_point(aes(color = Expression), size = 1.5, alpha = 0.6) +
    scale_color_manual(values = c("Upregulated" = "#fcbba1",
                                  "Downregulated" = "#3182bd",
                                  "Not_significant" = "grey",
                                  "TF_Upregulated" = "red",
                                  "TF_Downregulated" = "blue"),
                       labels = c("Upregulated" = "Up",
                                  "Downregulated" = "Down",
                                  "Not_significant" = "NS",
                                  "TF_Upregulated" = "TF Up",
                                  "TF_Downregulated" = "TF Down")) +
    geom_label_repel(data = label_sig,
                     aes(label = protein, color = Expression),
                     fill = "white", size = 3, box.padding = 0.5,
                     point.padding = 0, min.segment.length = 0,
                     segment.color = "gray50", segment.size = 0.4,
                     max.overlaps = 20, force = 5, force_pull = 0.3,
                     direction = "both", show.legend = FALSE) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
    theme_minimal() +
    labs(title = title, subtitle = subtitle_text,
         x = expression(log[2]~Fold~Change),
         y = expression(-log[10]~(p_value)),
         color = "Expression") +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          plot.subtitle = element_markdown(hjust = 0.5))
}



# Multiple contrasts from same dep result
# Set genes to label
genes_to_label <- c("NFIB", "NFIX", "JUNB", "CXXC1", "CREB5",
                    "RUNX2","FOXO1", "MYBL2", "KMT2A", "HMGA2")

res_HOS_MG63 <- prepare_data(HOS_U2OS_MG63_C1C4_result, "HOS_C1C4_vs_MG63_C1C4", TFs_list$Name)
labels_HOS_MG63 <- prepare_labels(res_HOS_MG63, genes_to_label)


write.xlsx(res_HOS_MG63, file = file.path(plots_dir, "Diffenrential_Enrichment_Protein_HOS_vs_MG63.xlsx"))
# Plot and save to PDF
plot_HOS_MG63 <-plot_volcano(res_HOS_MG63, labels_HOS_MG63, group_left = "MG63_C1C4", group_right = "HOS_C1C4",
                             title= "Diffenrential Enrichment Protein in MG63 vs HOS")
plot_HOS_MG63_raster <- ggrastr::rasterise(plot_HOS_MG63, layers = "Point", dpi = 600)
ggsave(plot_HOS_MG63_raster, file = file.path(plots_dir, "volcano_plot_DEP_HOS_vs_MG63_new.pdf"), width = 7, height = 5)

# Prepare each result
# Set genes to label
genes_to_label <- c("NFIB", "NFIX", "JUNB", "CREB5","CXXC1",
                    "RUNX1","RUNX2", "MYBL2", "TGIF2LX", "NR2C2", "HMGA2")
res_U2OS_MG63 <- prepare_data(HOS_U2OS_MG63_C1C4_result, "U2OS_C1C4_vs_MG63_C1C4", TFs_list$Name)
labels_U2OS_MG63 <- prepare_labels(res_U2OS_MG63, genes_to_label)
# Write the table
write.xlsx(res_U2OS_MG63, file = file.path(plots_dir, "Diffenrential_Enrichment_Protein_U2OS_vs_MG63.xlsx"))
# Plot and save to PDF
plot_U2OS_MG63 <- plot_volcano(res_U2OS_MG63, labels_U2OS_MG63, group_left = "MG63_C1C4", group_right = "U2OS_C1C4",
                   title="Diffenrential Enrichment Protein in MG63 vs U2OS")
plot_U2OS_MG63_raster <- ggrastr::rasterise(plot_U2OS_MG63, layers = "Point", dpi = 600)
ggsave(plot_U2OS_MG63_raster, file = file.path(plots_dir, "volcano_plot_DEP_U2OS_vs_MG63_new.pdf"), width = 7, height = 5)

```




```{r, Volcano Plot}
# Prepare a volcano plot for the contrast "MG63_C1C4_vs_MG63_NE""
# ===== Step 1: Classify expression status =====
prepare_data <- function(res) {
  res_df <- as.data.frame(res) %>%
    mutate(Expression = case_when(
      `p_value_-log10` >= 1.3 & log2_fold_change >= 1 ~ "Upregulated",
        `p_value_-log10` >= 1.3 & log2_fold_change <= -1 ~ "Downregulated",
          TRUE ~ "Not_significant"))
  return(res_df)}

# ===== Step 2: Create label dataframe for specific genes =====
prepare_labels <- function(df, genes_of_interest) {
  # Combine all genes into a single regex OR pattern (case-insensitive)
  pattern <- paste(genes_of_interest, collapse = "|") #Include all type of family
  #pattern <- paste0("(^|::)(", paste(genes_to_label, collapse = "|"), ")($|::)", collapse = "") #Exactly string name
  df %>% dplyr::filter(grepl(pattern, protein, ignore.case = TRUE)) %>%
    mutate(label_color = case_when(
      Expression == "Upregulated" ~ "red",
      Expression == "Downregulated" ~ "blue",
      TRUE ~ "grey"))}

# ===== Step 3: Plot volcano with custom gene labels & DEG count =====
plot_volcano <- function(res_df, label_df, title) {
  # Count number of DEGs
  num_up <- sum(res_df$Expression == "Upregulated")
  num_down <- sum(res_df$Expression == "Downregulated")
  # Subtitle with colored text using HTML
  subtitle_text <- paste0(
    "<span style='color:blue;'> Enrichment in MG63", ": ", num_down, "</span>",
    " | <span style='color:red;'> Enrichment in HOS", ": ", num_up, "</span>" )
  ggplot(res_df, aes(x = log2_fold_change, y = `p_value_-log10`)) +
    geom_point(aes(color = Expression), size = 1.5, alpha = 0.6) +
    scale_color_manual(values = c("Upregulated" = "red",
                                  "Downregulated" = "blue",
                                  "Not_significant" = "grey")) +
    geom_label_repel(data = label_df %>% dplyr::filter(label_color %in% c("red", "blue")),
                     aes(label = protein),
                     color = label_df %>% dplyr::filter(label_color %in% c("red", "blue")) %>% pull(label_color),
                     fill = "white", size = 3, box.padding = 0.5, point.padding = 0.0, min.segment.length = 0,
                     segment.color = "gray50", segment.size = 0.4, max.overlaps = Inf, force = 5,
                     force_pull = 0.3, direction = "both", nudge_y = 0.05, show.legend = FALSE) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
    theme_minimal() +
    labs(title = title,
         subtitle = subtitle_text,
         x = "Log2FoldChange",
         y = "-log10(padj)",
         color = "Expression") +
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_markdown(hjust = 0.5))
}


# Set genes to label
genes_to_label <- c("NFIB", "NFIX", "JUNB", "CXXC1", "CREB5",
                    "RUNX2","FOXO1", "MYBL2", "KMT2A", "HMGA2")
# Prepare each result
res1 <- prepare_data(HOS_MG63_C1C4)
labels1 <- prepare_labels(res1, genes_to_label)


#write.csv(res1, file = file.path(plots_dir, "volcano_TOBIAS_HOS_vs_MG63.csv"))
write.xlsx(res1, file = file.path(plots_dir, "Diffenrential_Enrichment_Protein_HOS_vs_MG63.xlsx"))
# Plot and save to PDF
p1 <- plot_volcano(res1, labels1, "Diffenrential Enrichment Protein in MG63 vs HOS")
p1_raster <- ggrastr::rasterise(p1, layers = "Point", dpi = 600)
ggsave(p1_raster, file = file.path(plots_dir, "volcano_plot_DEP_HOS_vs_MG63.pdf"), width = 7, height = 5)

# Set genes to label
genes_to_label <- c("NFIB", "NFIX", "JUNB", "CREB5","CXXC1",
                    "RUNX1","RUNX2", "MYBL2", "TGIF2LX", "NR2C2", "HMGA2")
# Prepare each result
res2 <- prepare_data(U2OS_MG63_C1C4)
labels2 <- prepare_labels(res2, genes_to_label)
library(openxlsx)
#write.csv(res1, file = file.path(plots_dir, "volcano_TOBIAS_HOS_vs_MG63.csv"))
write.xlsx(res2, file = file.path(plots_dir, "Diffenrential_Enrichment_Protein_U2OS_vs_MG63.xlsx"))
# Plot and save to PDF
p2 <- plot_volcano(res2, labels2, "Diffenrential Enrichment Protein in MG63 vs U2OS")
p2_raster <- ggrastr::rasterise(p2, layers = "Point", dpi = 600)
ggsave(p2_raster, file = file.path(plots_dir, "volcano_plot_DEP_U2OS_vs_MG63.pdf"), width = 7, height = 5)

```


```{r, Integrate the volcano plot}
# ===== Step 1: Prepare data (add Design_C1 here, not in plot) =====
prepare_integration_data <- function(dep_results, contrast1, contrast2, tf_list, designed_list = NULL) {
  ratio1 <- paste0(contrast1, "_ratio")
  ratio2 <- paste0(contrast2, "_ratio")
  pval1 <- paste0(contrast1, "_p.val")
  pval2 <- paste0(contrast2, "_p.val")
  
  dep_results %>%
    mutate(protein = name,
           lfc_1 = .data[[ratio1]],
           lfc_2 = .data[[ratio2]],
           pval_1 = .data[[pval1]],
           pval_2 = .data[[pval2]],
           TF = if_else(protein %in% tf_list, "YES", "NO"),
           Design_C1 = if_else(!is.null(designed_list) & protein %in% designed_list, "YES", "NO"),
           sig_1 = pval_1 <= 0.05 & abs(lfc_1) >= 1,
           sig_2 = pval_2 <= 0.05 & abs(lfc_2) >= 1,
           hit = sig_1 & sig_2,
           point_group = case_when(TF == "YES" & hit ~ "TF (significant both)",
                              TF == "YES" & (sig_1 | sig_2) ~ "TF (significant one)",
                              TF == "YES" ~ "TF (not significant)",
                              hit ~ "Significant both",
                              sig_1 | sig_2 ~ "Significant one",
                              TRUE ~ "Not significant"),
           quadrant = case_when(sig_1 & lfc_1 < -1 & sig_2 & lfc_2 > 1 ~ "Down_Up",
                              sig_1 & lfc_1 > 1 & sig_2 & lfc_2 > 1 ~ "Up_Up",
                              sig_1 & lfc_1 < -1 & sig_2 & lfc_2 < -1 ~ "Down_Down",
                              sig_1 & lfc_1 > 1 & sig_2 & lfc_2 < -1 ~ "Up_Down",
                              TRUE ~ "NS"))
}

# ===== Step 2: Plot integration =====
plot_integration <- function(df, label_genes, title, xlab_name = "HOS/MG63", ylab_name = "U2OS/MG63") {
  quad_counts <- df %>% dplyr::filter(quadrant != "NS") %>% dplyr::count(quadrant)
  label_text <- paste0(quad_counts$quadrant, ": ", quad_counts$n, collapse = "\n")
  
  df <- df %>%
    mutate(point_group = factor(point_group,
      levels = c("Not significant", "Significant one", "Significant both",
                 "TF (not significant)", "TF (significant one)", "TF (significant both)"))) %>%
    arrange(point_group)
  
  label_df <- df %>% dplyr::filter(protein %in% label_genes)
  # plot the integration
  ggplot(df, aes(x = lfc_1, y = lfc_2)) +
    geom_point(aes(color = point_group, alpha = point_group, shape = Design_C1), size = 1.5) +
    scale_color_manual(values = c("Not significant" = "grey80",
                                  "Significant one" = "grey80",
                                  "Significant both" = "pink",
                                  "TF (not significant)" = "grey80",
                                  "TF (significant one)" = "grey80",
                                  "TF (significant both)" = "red")) +
    scale_alpha_manual(values = c("Not significant" = 0.3,
                                  "Significant one" = 0.5,
                                  "Significant both" = 0.7,
                                  "TF (not significant)" = 0.5,
                                  "TF (significant one)" = 0.7,
                                  "TF (significant both)" = 1)) +
    scale_shape_manual(values = c("YES" = 17, "NO" = 16)) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    geom_label_repel(data = label_df,
                     aes(label = protein, color = point_group),
                     fill = "white", size = 3, box.padding = 0.5,
                     point.padding = 0, min.segment.length = 0,
                     segment.color = "gray50", segment.size = 0.4,
                     max.overlaps = 20, force = 5, force_pull = 0.3,
                     show.legend = FALSE) +
    annotate("label", x = Inf, y = -Inf, label = label_text,
             hjust = 1.1, vjust = -0.1, size = 3.5, fill = "white") +
    theme_minimal() +
    labs(title = title,
         x = bquote(log[2]~FC~(.(xlab_name))),
         y = bquote(log[2]~FC~(.(ylab_name))),
         color = "Category", alpha = "Category", shape = "Designed TF") +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "right") +
    guides(alpha = "none")
}

# Read the designed TF list
designed_TFs <- readxl::read_excel(file.path(analysis_dir, "catTFRE_designed_TFs_126_TFs.xlsx"))

# Plot the integration plot
int_df <- prepare_integration_data(HOS_U2OS_MG63_C1C4_result,
                                   contrast1 = "HOS_C1C4_vs_MG63_C1C4",
                                   contrast2 = "U2OS_C1C4_vs_MG63_C1C4",
                                   designed_list = designed_TFs$TF_design,
                                   tf_list = TFs_list$Name)
write.xlsx(int_df, file = file.path(plots_dir, "HOS_U2OS_MG63_C1C4_result.xlsx"))
# add gene to labeling
genes_to_label <- c("RUNX2", "HMGA2", "NR2C2", "ADNP", "UBP1",
                     "JUNB", "NFIX", "NFIB", "PITX1", "JUND")
# ggplot
plot_integration(int_df, genes_to_label,
                 title = "Integration HOS/MG63 vs U2OS/MG63 C1C4",
                 xlab_name = "HOS/MG63",
                 ylab_name = "U2OS/MG63")
ggsave(file = file.path(plots_dir, "HOS_U2OS_MG63_C1C4_integrate.pdf"), width=6, height=4)

```


#### 4.5. Visualization of functional enrichment result using clusterProfile R package
```{r}
# Get significant genes from each contrast
sig_HOS <- HOS_U2OS_MG63_C1C4_result %>%
  dplyr::filter(HOS_C1C4_vs_MG63_C1C4_p.val <= 0.05 & abs(HOS_C1C4_vs_MG63_C1C4_ratio) >= 1) %>%
  pull(name)
sig_U2OS <- HOS_U2OS_MG63_C1C4_result %>%
  dplyr::filter(U2OS_C1C4_vs_MG63_C1C4_p.val <= 0.05 & abs(U2OS_C1C4_vs_MG63_C1C4_ratio) >= 1) %>%
  pull(name)

# Union for heatmap
gene_plot_heatmap <- unique(c(sig_HOS, sig_U2OS))
# Build heatmap matrix
heatmap_matrix <- assay(data_imp)[rownames(assay(data_imp)) %in% gene_plot_heatmap, ]
heatmap_z <- t(scale(t(heatmap_matrix)))
# TF annotation
tf_annot <- rownames(heatmap_z) %in% TFs_list$Name
row_ha <- rowAnnotation(TF = tf_annot,
                        col = list(TF = c("TRUE" = "darkgreen", "FALSE" = "grey90")))
# Counts
tf_up_HOS <- sum(rownames(heatmap_z) %in% sig_HOS & 
                  HOS_U2OS_MG63_C1C4_result$HOS_C1C4_vs_MG63_C1C4_ratio[match(rownames(heatmap_z), HOS_U2OS_MG63_C1C4_result$name)] > 1 &
                  rownames(heatmap_z) %in% TFs_list$Name)

# Simpler approach for counts — annotate the result table once
dep_annot <- HOS_U2OS_MG63_C1C4_result %>%
  mutate(TF = if_else(name %in% TFs_list$Name, "YES", "NO"),
         HOS_expr = case_when(HOS_C1C4_vs_MG63_C1C4_p.adj <= 0.05 & HOS_C1C4_vs_MG63_C1C4_ratio >= 1 ~ "Up",
                              HOS_C1C4_vs_MG63_C1C4_p.adj <= 0.05 & HOS_C1C4_vs_MG63_C1C4_ratio <= -1 ~ "Down",
                              TRUE ~ "NS"),
         U2OS_expr = case_when(U2OS_C1C4_vs_MG63_C1C4_p.adj <= 0.05 & U2OS_C1C4_vs_MG63_C1C4_ratio >= 1 ~ "Up",
                               U2OS_C1C4_vs_MG63_C1C4_p.adj <= 0.05 & U2OS_C1C4_vs_MG63_C1C4_ratio <= -1 ~ "Down",
                               TRUE ~ "NS"))

column_title <- paste0("HOS: Up=", sum(dep_annot$HOS_expr == "Up"),
                       " (TF=", sum(dep_annot$HOS_expr == "Up" & dep_annot$TF == "YES"), ")",
                       " Down=", sum(dep_annot$HOS_expr == "Down"),
                       " (TF=", sum(dep_annot$HOS_expr == "Down" & dep_annot$TF == "YES"), ")",
                       "\nU2OS: Up=", sum(dep_annot$U2OS_expr == "Up"),
                       " (TF=", sum(dep_annot$U2OS_expr == "Up" & dep_annot$TF == "YES"), ")",
                       " Down=", sum(dep_annot$U2OS_expr == "Down"),
                       " (TF=", sum(dep_annot$U2OS_expr == "Down" & dep_annot$TF == "YES"), ")")
library(ComplexHeatmap)
library(circlize)

col_fun <- colorRamp2(c(-2, 0, 2),
                      c("skyblue", "white", "purple"))

# column order
col_order <- c(grep("MG63", colnames(heatmap_z), value = TRUE),
               grep("U2OS", colnames(heatmap_z), value = TRUE),
               grep("HOS",  colnames(heatmap_z), value = TRUE))

heatmap_z_ordered <- heatmap_z[, col_order]
# group labeling
group_labels <- gsub("_C1C4_rep_\\d+$", "", colnames(heatmap_z_ordered))

col_annot <- HeatmapAnnotation(Group = group_labels,
                               col = list(Group = c("MG63" = "grey","U2OS" = "orange","HOS"  = "steelblue")))
# Heatmap
HOS_MG63_C1C4_heatmap <- Heatmap(heatmap_z_ordered,
                                 name = "Z-score",
                                 column_title = column_title,
                                 top_annotation = col_annot,
                                 right_annotation = row_ha,
                                 col = col_fun,
                                 show_row_names = FALSE,
                                 cluster_rows = TRUE,
                                 cluster_columns = FALSE,
                                 row_split = 5,
                                 row_gap = unit(2, "mm"))

pdf(file.path(plots_dir,"HOS_MG63_C1C4_catTFRE_complexHeatmap.pdf"),height = 10,width = 4.5)
draw(HOS_MG63_C1C4_heatmap)
dev.off()

```


**4.5. Visualization of functional enrichment result using clusterProfile R package**
```{r}
library(dplyr)
library(msigdbr)
library(clusterProfiler)

# ===== Step 1: Prepare ranked gene list from DEP results =====
prepare_genelist <- function(dep_results, contrast) {
  ratio_col <- paste0(contrast, "_ratio")
  dep_results %>%
    dplyr::filter(!is.na(.data[[ratio_col]])) %>%
    arrange(desc(.data[[ratio_col]])) %>%
    {setNames(sort(.[[ratio_col]], decreasing = TRUE),
              .$ID[order(-.[[ratio_col]])])}
}

# ===== Step 2: Run GSEA for a gene set category =====
run_gsea <- function(geneList, category = "H", subcategory = NULL, pval = 0.05) {
  gene_sets <- msigdbr(species = "Homo sapiens", category = category, subcategory = subcategory) %>%
    dplyr::select(gs_name, gene_symbol)
  
  GSEA(geneList = geneList,
       TERM2GENE = gene_sets,
       pvalueCutoff = pval,
       verbose = FALSE)
}

# ===== Step 3: Subset top N up and down =====
subset_gsea_top <- function(gsea_res, n = 5) {
  top_up <- gsea_res@result %>%
    dplyr::filter(NES > 0) %>%
    arrange(p.adjust) %>%
    slice_head(n = n) %>%
    pull(ID)
  
  top_down <- gsea_res@result %>%
    dplyr::filter(NES < 0) %>%
    arrange(p.adjust) %>%
    slice_head(n = n) %>%
    pull(ID)
  
  gsea_sub <- gsea_res
  gsea_sub@result <- gsea_res@result %>%
    dplyr::filter(ID %in% c(top_up, top_down))
  return(gsea_sub)
}

# ===== Step 4: Plot ridge + dot =====
plot_gsea <- function(gsea_res, contrast, plots_dir,
                      label_activated = NULL, label_suppressed = NULL,
                      top_n = 5, height_ridge = 6, height_dot = 6, width = 8) {
  gsea_sub <- subset_gsea_top(gsea_res, n = top_n)
  # Auto-generate labels from contrast if not provided
  if (is.null(label_activated)) {
    parts <- strsplit(contrast, "_vs_")[[1]]
    label_activated <- parts[1]
    label_suppressed <- parts[2]
  }
  # Ridge plot
  p_ridge <- ridgeplot(gsea_sub, showCategory = top_n * 2)
  ggsave(file.path(plots_dir, paste0(contrast, "_ridgeplot.pdf")),
         p_ridge, height = height_ridge, width = width)
  # Dot plot
  p_dot <- dotplot(gsea_sub, showCategory = top_n * 2, split = ".sign") +
    facet_grid(~ .sign, labeller = labeller(
      .sign = c("activated" = label_activated, "suppressed" = label_suppressed)))
  ggsave(file.path(plots_dir, paste0(contrast, "_dotplot.pdf")),
         p_dot, height = height_dot, width = width)
  
  return(list(ridge = p_ridge, dot = p_dot, gsea_sub = gsea_sub))
}



# ===== Run everything for multiple contrasts =====
contrasts <- c("HOS_C1C4_vs_MG63_C1C4", "U2OS_C1C4_vs_MG63_C1C4")
results <- list()
for (contrast in contrasts) {
  # Prepare gene list
  geneList <- prepare_genelist(HOS_U2OS_MG63_C1C4_result, contrast)
  # Hallmark
  gsea_hallmark <- run_gsea(geneList, category = "H")
  p_hallmark <- ridgeplot(gsea_hallmark)
  ggsave(file.path(plots_dir, paste0(contrast, "_hallmark_ridgeplot.pdf")),
         p_hallmark, height = 4, width = 8)
  # GO:BP
  gsea_gobp <- run_gsea(geneList, category = "C5", subcategory = "GO:BP")
  plots <- plot_gsea(gsea_gobp, contrast, plots_dir)
  # Store results
  results[[contrast]] <- list(geneList = geneList,
                              hallmark = gsea_hallmark,
                              gobp = gsea_gobp,
                              plots = plots)
}

```



```{r, GSEA analysis}
# generate the gene_annotation file using fetchAnnotation function in gsfisher (or using biomaRt)
if(!file.exists("annotation.rds")){
gene_annotation <- fetchAnnotation(species = "hs",
                 ensembl_version = NULL,ensembl_host = NULL)
saveRDS(gene_annotation, "annotation.rds") 
}
# Load gene_annotation file
gene_annotation <- readRDS("annotation.rds")

# Generate GSEA analysis by using unique expression matrix (no gene names duplicate)
catTFRE_MS_df <- catTFRE_MS_unique[, 1:24]
catTFRE_MS_df$Entrez_ID <- gene_annotation$entrez_id[match(catTFRE_MS_unique$Gene_Symbol, gene_annotation$gene_name)]
# remove duplicate and NA Entrez ID
catTFRE_MS_df <- catTFRE_MS_df[!duplicated(catTFRE_MS_df$Entrez_ID), ]
catTFRE_MS_df <- catTFRE_MS_df[!is.na(catTFRE_MS_df$Entrez_ID), ]
# Finalize the expression dataset
rownames(catTFRE_MS_df) <- catTFRE_MS_df$Entrez_ID
heatmap_data <- catTFRE_MS_df[, 1:24]
head(heatmap_data)
# Sort the matrix by rowmean()
heatmap_data <- heatmap_data[order(rowMeans(heatmap_data), decreasing=TRUE), ]


# Data from DEG will using for GSEA analysis
msigdbr_collections()
# Extract the hallmark pathway
pathwaysDF <- as.data.frame(msigdbr("hs", category="H")) #subcategory = "")
# pathwaysDF <- as.data.frame(msigdbr("human", category="C2", subcategory = "CP:KEGG"))
# pathwaysDF <- as.data.frame(msigdbr("human", category="C5", subcategory = "GO:BP"))

# Extract the gene list following pathway
pathways <- split(as.character(pathwaysDF$entrez_gene), pathwaysDF$gs_name)

# Run the GSEA analysis
set.seed(1234)
fgsea_MS <- geseca(pathways, heatmap_data, minSize = 10, maxSize = 500)
# Plot the pathway
pdf(file = "OS_catTFRE_C1C4_GSEA_hallmark_top10.pdf", height = 5.5, width = 14)
plotGesecaTable(fgsea_MS |> head(10), pathways, E=heatmap_data, titles = colnames(heatmap_data))
dev.off()
```


```{r}
sessionInfo()
```


