---
title: "MS_Osteosarcoma_Statistic_FinalRound_Emilydata"
author: "XuanThangNguyen"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---
# Summary 
This script implements a simple overview of Mass Spectrometry (MS) data analysis with **DEP package**. 
The data including 3 cell lines (HOS, U2OS and MG63), wiht 4 biological/technical replicates, that make total 36 single run

The main steps of analysis:  
- Exploring data before the differential expression analysis  
- Importing **MS counts** into **DEP package** list    
- Performing the analysis with **DESeq2**  
- Exporting and exploring the results  

The results are exported into a text file and visualized using ggplot heatmaps and volcano-plots. 

Detailed tutorial for **DEP package** package could be accessed here:  
- https://bioconductor.org/packages/devel/bioc/vignettes/DEP/inst/doc/DEP.html

#### 1. Start section by loading R packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Importing required packages
library("openxlsx")
library("dplyr")
library("readxl")
library("tidyverse")
library("DEP")
library("ggplot2")
library("forcats")
# Note that DEP package install MacOS-M1 sometime have trouble that can be resolved by installing gcc lower version.
# Time stamp
Sys.time()
# Note that DEP package install MacOS-M1 sometime have trouble that can be resolved by installing gcc lower version.
set.seed(1234)

# Set directory paths
root_dir <- rprojroot::find_root(rprojroot::has_dir("catTFRE_MS_FinalRound"))
analysis_dir <- file.path(root_dir, "catTFRE_MS_FinalRound/DEP_Plots_ALL")
plots_dir <- file.path(analysis_dir, "DEP_Plots")
# Create plots_dir if it doesn't exist
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}

```


```{r, ALL SUM}
## Import the data
catTRFE_MS_Round4 <- read.xlsx(file.path(analysis_dir, "catTRFR_MS_MG63_U2OS_HOS_Round4_sum_extend.xlsx"), rowNames = F)
# colnames(catTRFE_MS_Round4)

## Data sample Overview
# Experiment design for Total with only 1 replication will not consider to put number in replication column. 
# If you set up with any replication with only one, make_se function will not works
experimental_design <- data.frame(label = colnames(catTRFE_MS_Round4)[grep("#Spec.Sample",colnames(catTRFE_MS_Round4))],
                                  cell = c("All", rep("MG63",16), rep("U2OS",16), rep("HOS",16)),
                                  condition = c("All",
                                                "MG63_All", rep("MG63_C1",5),  rep("MG63_C4", 5), rep("MG63_NE", 5),
                                                "U2OS_All", rep("U2OS_C1",5),  rep("U2OS_C4", 5), rep("U2OS_NE", 5),
                                                "HOS_All", rep("HOS_C1",5),  rep("HOS_C4", 5), rep("HOS_NE", 5)),
                                  replicate = c("","","",c(1:4),"",c(1:4),"",c(1:4),
                                                   "","",c(1:4),"",c(1:4),"",c(1:4),
                                                   "","",c(1:4),"",c(1:4),"",c(1:4)))
# write.xlsx(experimental_design, file = "MS_experiment_design_Round4.xlsx")
sample_names <- experimental_design %>% 
  mutate(sample_names=ifelse(replicate=="", condition, paste(condition, replicate, sep = "_"))) %>%
  dplyr::select(label,sample_names)

#Data with sample names
MS_data_Round4 <- catTRFE_MS_Round4
# Change the sample name as real name
# colnames(catTRFE_MS_Round4)[1:49] <- sample_names$sample_names
colnames(MS_data_Round4) <- ifelse(colnames(MS_data_Round4) %in% sample_names$label, sample_names$sample_names, colnames(MS_data_Round4))

# Extract MS data with experiment
MS_data_raw <- MS_data_Round4 %>% 
  dplyr::select(Protein_Symbol, colnames(MS_data_Round4)[1:49]) %>%  #contains("#Spec.Sample")
  gather(sample, expression, -Protein_Symbol) %>%
  mutate(cell = sapply(strsplit(sample, "_"), "[", 1), 
         construct = sapply(strsplit(sample, "_"), "[", 2),
         experiment=paste(cell, construct, sep = "_"  ))  %>% 
  mutate(sample = fct_relevel(sample, unique(MS_data$sample)))
# Filtering data with out expression
MS_data <- MS_data_raw %>% dplyr::filter(expression > 0)

# See what data look like
dplyr::count(MS_data, sample)
dplyr::count(MS_data, experiment)
```

#### Overview of total protein Plots
```{r}
## Plot All data from catTFRE Mass spectrometry with expression > 0
num_Protein <- length(unique(MS_data$Protein_Symbol))
#Plot with ggplot2
ggplot(MS_data, aes(x=sample, fill=experiment)) +
  geom_bar() + 
  coord_flip() +
  geom_hline(yintercept = num_Protein, color = "coral1") +
  xlab("") +
  ylab("") +
    # Add number annotation next to the line
  annotate("text",
           x = 1, #length(unique(MS_data$sample)) + 0.5,  # place outside last sample row
           y = num_Protein + 2,  # slightly above the line
           label = num_Protein,color = "coral1",hjust = 0) +
  ggtitle("OS catTFRE Mass spectrometry TFs ONLY expression > 0 per samples")
ggsave(file.path(plots_dir, "OS_catTFRE_Mass_spectrometry_Allprotein.pdf"), width=4, height=10)

# Select ONLY replicate
MS_replicates <- MS_data %>%
  dplyr::filter(grepl("_[1-4]$", sample)) 
table(MS_replicates$sample)

# Plot
ggplot(MS_replicates, aes(x=sample, fill=experiment)) +
  geom_bar() + 
  coord_flip() +
  xlab("") +
  ylab("Number of protein") +
  ggtitle("OS catTFRE Mass spectrometry per samples")
ggsave(file.path(plots_dir, "OS_catTFRE_MS_Allprotein_rainbow.pdf"), width=4, height=10)

ggplot(MS_replicates, aes(x = sample, fill = experiment)) +
  geom_bar() +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "HOS_C1"  = "#8332FF", "HOS_C4" = "#A870FF", "HOS_NE" = "#C199FF",
      "U2OS_C1" = "#F66239", "U2OS_C4"= "#F88161", "U2OS_NE"= "#FAA188",
      "MG63_C1" = "#56B4E9", "MG63_C4"= "#78C3ED", "MG63_NE"= "#ABDAF4"))
#      "HOS_C1"  = "#8332FF", "HOS_C4" = "#9969C7", "HOS_NE" = "#AF8FE9",
#      "U2OS_C1" = "#fc8d59", "U2OS_C4"= "#fdae61", "U2OS_NE"= "#fee090",
#      "MG63_C1" = "#0570b0", "MG63_C4"= "#74add1", "MG63_NE"= "#abd9e9"))
ggsave(file.path(plots_dir, "OS_catTFRE_MS_Allprotein_colorManual.pdf"), width=4, height=10)

library(ggpattern)
ggplot(MS_replicates, aes(x = sample)) +
  geom_bar_pattern(
    aes(fill = cell, pattern = construct),
    position = "stack",
    color = NA,                 # no bar outline
    pattern_fill = "magenta",       # pattern color
    pattern_color = "magenta",      # match pattern stroke
    pattern_density = 0.05,
    pattern_spacing = 0.06,
    pattern_key_scale_factor = 0.5) +
  scale_fill_manual(values = c(MG63 = "#56B4E9", U2OS = "#F66239", HOS = "#8332FF"), name = "Cell") +
  scale_pattern_manual(values = c(C1 = "none", C4 = "stripe", NE = "crosshatch"), name = "Construct") +
  coord_flip() +
  labs(x = NULL, y = "Number of proteins",
       title = "OS catTFRE Mass spectrometry per sample",
       pattern = "Construct") +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) + 
  guides(fill    = guide_legend(override.aes = list(pattern = "none")),   # Cell legend: only colors
         pattern = guide_legend(override.aes = list(fill = "white")))      # Construct legend: only patterns
ggsave(file.path(plots_dir, "OS_catTFRE_MS_Allprotein_parttern.pdf"), width=4, height=10)
```


```{r, TF ONLY}
TFs_list <- read.xlsx("TFs_list_Final.xlsx")
# Select ONLY replicate
MS_TF_replicates <- MS_data %>%
  dplyr::filter(grepl("_[1-4]$", sample)) %>% 
  dplyr::filter(Protein_Symbol %in% TFs_list$Gene_Symbol)
table(MS_TF_replicates$sample)

# Plot
ggplot(MS_TF_replicates, aes(x=sample, fill=experiment)) +
  geom_bar() + 
  coord_flip() +
  xlab("") +
  ylab("Number of protein") +
  ggtitle("OS catTFRE Mass spectrometry per samples")
ggsave(file.path(plots_dir, "OS_catTFRE_MS_TFs_ONLY_rainbow.pdf"), width=4, height=10)

ggplot(MS_TF_replicates, aes(x = sample, fill = experiment)) +
  geom_bar() +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "HOS_C1"  = "#8332FF", "HOS_C4" = "#A870FF", "HOS_NE" = "#C199FF",
      "U2OS_C1" = "#F66239", "U2OS_C4"= "#F88161", "U2OS_NE"= "#FAA188",
      "MG63_C1" = "#56B4E9", "MG63_C4"= "#78C3ED", "MG63_NE"= "#ABDAF4"))
#      "HOS_C1"  = "#8332FF", "HOS_C4" = "#9969C7", "HOS_NE" = "#AF8FE9",
#      "U2OS_C1" = "#fc8d59", "U2OS_C4"= "#fdae61", "U2OS_NE"= "#fee090",
#      "MG63_C1" = "#0570b0", "MG63_C4"= "#74add1", "MG63_NE"= "#abd9e9"))
ggsave(file.path(plots_dir, "OS_catTFRE_MS_TFs_ONLY_colorManual.pdf"), width=4, height=10)

```


```{r}

library(dplyr)
library(tidyr)
library(stringr)

# Fill missing replicates with expression = 0
fill_missing_reps <- function(df, genes = NULL, reps = 1:4) {
  stopifnot(all(c("Protein_Symbol","sample","expression","cell","construct","experiment") %in% names(df)))
  
  df2 <- df
  if (!is.null(genes)) df2 <- df2 %>% filter(Protein_Symbol %in% genes)

  # Extract replicate number from sample like HOS_NE_3
  df2 <- df2 %>%
    mutate(rep = suppressWarnings(as.integer(stringr::str_extract(sample, "(?<=_)\\d+$"))))

  # Build all expected sample names for each gene/cell/construct
  template <- df2 %>%
    distinct(Protein_Symbol, cell, construct, experiment) %>%
    tidyr::expand(Protein_Symbol, cell, construct, experiment, rep = reps) %>%
    mutate(sample = paste(cell, construct, rep, sep = "_"))

  # Join with existing data, fill missing with 0
  out <- template %>%
    left_join(df2, by = c("Protein_Symbol","cell","construct","experiment","rep","sample")) %>%
    mutate(expression = ifelse(is.na(expression), 0, expression)) %>%
    arrange(Protein_Symbol, cell, construct, rep)

  out
}

# Look the Mass spec of single protein each replicate
# RUNX2 expression
MS_data_POI <- MS_replicates %>% filter((Protein_Symbol == "RUNX2" & cell == "HOS") |
                                          (Protein_Symbol == "PITX1" & cell == "MG63") |
                                          (Protein_Symbol == "FOXO1" & cell == "U2OS"))
MS_data_POI <- fill_missing_reps(MS_data_POI, genes = c("RUNX2","FOXO1","PITX1"), reps = 1:4)


ggplot(MS_data_POI, aes(x = sample, y = expression, fill = Protein_Symbol)) +
  geom_col() +
  facet_wrap(~ cell, scales = "free", strip.position = "top") +
  scale_fill_manual(values = c(RUNX2="#8332FF", FOXO1="#F66239", PITX1="#56B4E9")) +
  labs(x = "", y = "MS Spectral Counts",
       title = "MS spectral counts for protein") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text = element_text(face = "bold"),
        legend.title = element_blank())
ggsave(file.path(plots_dir, "Spectral_count_MS_selectGenes.pdf"), width=8, height=4) # starts writing a PDF to file
```




```{r}
# Vendiagram data between All protein of HOS/U2OS/MG63
library("ggVennDiagram")
dplyr::count(MS_data, sample)
OS_MS  <- list(HOS_C1 = MS_data[MS_data$sample=="HOS_C1", ]$Protein_Symbol,
               HOS_C4 = MS_data[MS_data$sample=="HOS_C4", ]$Protein_Symbol,
               HOS_NE = MS_data[MS_data$sample=="HOS_NE", ]$Protein_Symbol,
               U2OS_C1 = MS_data[MS_data$sample=="U2OS_C1", ]$Protein_Symbol,
               U2OS_C4 = MS_data[MS_data$sample=="U2OS_C4", ]$Protein_Symbol,
               U2OS_NE = MS_data[MS_data$sample=="U2OS_NE", ]$Protein_Symbol,
               MG63_C1 = MS_data[MS_data$sample=="MG63_C1", ]$Protein_Symbol,
               MG63_C4 = MS_data[MS_data$sample=="MG63_C4", ]$Protein_Symbol,
               MG63_NE = MS_data[MS_data$sample=="MG63_NE", ]$Protein_Symbol)


MS_All_TFs <- list(HOS_C1_TFs = MS_data[MS_data$sample=="HOS_C1", ]$Protein_Symbol,
                   U2OS_C1_TFs = MS_data[MS_data$sample=="U2OS_C1", ]$Protein_Symbol,
                   MG63_C1_TFs = MS_data[MS_data$sample=="MG63_C1", ]$Protein_Symbol,
                   HOS_C4_TFs = MS_data[MS_data$sample=="HOS_C4", ]$Protein_Symbol,
                   U2OS_C4_TFs = MS_data[MS_data$sample=="U2OS_C4", ]$Protein_Symbol,
                   MG63_C4_TFs = MS_data[MS_data$sample=="MG63_C4", ]$Protein_Symbol,
                   HOS_NE_TFs = MS_data[MS_data$sample=="HOS_NE", ]$Protein_Symbol,
                   U2OS_NE_TFs = MS_data[MS_data$sample=="U2OS_NE", ]$Protein_Symbol,
                   MG63_NE_TFs = MS_data[MS_data$sample=="MG63_NE", ]$Protein_Symbol)

library(ComplexUpset)
library(tidyr)
library(dplyr)
# Filter each set to keep only TFs
MS_TF <- lapply(OS_MS, function(x) x[x %in% TFs_list$Gene_Symbol])
# Convert list of sets into binary membership matrix
input_upset_TF <- UpSetR::fromList(MS_TF) %>% as.data.frame() %>% mutate(Protein = rownames(.))
# Add protein names back as a column
#input_upset_TF$Protein <- rownames(input_upset_TF)
my_order <- c("MG63_C1", "MG63_C4", "MG63_NE", "U2OS_C1", "U2OS_C4", "U2OS_NE", "HOS_C1", "HOS_C4", "HOS_NE")
# Plot overlap only for TFs
pdf(file.path(plots_dir,"catTFRE_All_TFs_ONLY_UpsetPlot.pdf"), width=12, height=6) # starts writing a PDF to file
ComplexUpset::upset(input_upset_TF,
                    intersect = my_order,
                    name = "TF Overlap",
                    min_size = 5, sort_sets = FALSE)
dev.off()

# 1. Convert list of sets into a binary membership matrix
input_upset <- UpSetR::fromList(OS_MS) %>% as.data.frame()
# 2. Add protein names as a column
input_upset$Protein <- rownames(input_upset)
# 3. Plot with ComplexUpset
pdf(file.path(plots_dir,"catTFRE_All_protein_UpsetPlot.pdf"), width=12, height=6) # starts writing a PDF to file
ComplexUpset::upset(input_upset,
                    intersect = my_order,
                    name = "Protein Overlap",
                    min_size = 64, sort_sets = FALSE)
dev.off()

```

#### 2.3. Load the Transcription factor list

Human TFs list of Transcription Factor (TFs) were generated by merging the list TFs between: 
Human TFs list of Human Protein Atlas database: https://www.proteinatlas.org/search/transcription+factor
Human TFs list of AnimalTFDB: # http://bioinfo.life.hust.edu.cn/AnimalTFDB/#!/download
Human TFs list of Human Transcription Factors network 2018: # https://doi.org/10.1016/j.cell.2018.01.029
Human TFs list of Uniprot database: https://www.uniprot.org/uniprotkb?facets=model_organism%3A9606%2Cexistence%3A2&query=transcription%20factor 
Human TFs list of GeneCards database: https://www.genecards.org/Search/Keyword?queryString=%22human%20transcription%20factor%22







