---
title: "MS_Osteosarcoma_FinalRound_Emilydata_C1"
author: "XuanThangNguyen"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---
### Summary 
This script implements a simple overview of Mass Spectrometry (MS) data analysis with **DEP package**. 
The data including 3 cell lines (HOS, U2OS and MG63), with 4 biological/technical replicates, that make total 36 single run

The main steps of analysis:  
- Exploring data before the differential expression analysis  
- Importing **MS counts** into **DEP package** list    
- Performing the analysis with **DESeq2**  
- Exporting and exploring the results  
The results are exported into a text file and visualized using ggplot heatmaps and volcano-plots. 

Detailed tutorial for **DEP package** package could be accessed here:  
- https://bioconductor.org/packages/devel/bioc/vignettes/DEP/inst/doc/DEP.html

**1. Load package**

```{r setup, include=FALSE}
# Importing required packages
library("dplyr")
library("DEP")
library("DOSE")
library("openxlsx")
library("ggplot2")
library("SummarizedExperiment")
library("ComplexHeatmap")
library("clusterProfiler")
library("org.Hs.eg.db")
library("AnnotationHub")
library("circlize")
library("gsfisher")
library("data.table")
library("fgsea")
library("msigdbr")

set.seed(1234)

# Set directory paths
root_dir <- rprojroot::find_root(rprojroot::has_dir("catTFRE_MS_FinalRound"))
analysis_dir <- file.path(root_dir, "catTFRE_MS_FinalRound")
plots_dir <- file.path(analysis_dir, "DEP_Plots_C1_threshold0")
# Create plots_dir if it doesn't exist
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}


```

**2. Import and manipulate MS data from Emily and Transcription factor list**
Data were ready generated by Emily and XXX that saved in .csv file with rowname is gene/protein list and colname is sample names. Loading by using open.xlsx package
```{r}
# 1.1 Import the data
# Load the data
# Load the data
catTFRE_MS_Round4 <- read.xlsx(file.path(analysis_dir,"catTFRE_Round4_HOS_MG63_U2OS_Thang.xlsx"), rowNames = F)
colnames(catTFRE_MS_Round4)

# Filter only C1 Mass Spectrmetry data
catTRFE_MS_C1 <- catTFRE_MS_Round4[ , c(2:5, 14:17, 26:29, 38:40)]
colnames(catTRFE_MS_C1)
dim(catTRFE_MS_C1)
# c(rep("HOS_C1", 4), rep("MG63_C1", 4), rep("U2OS_C1", 4))

# Human TFs list of Transcription Factor (TFs) were generated by merging the list TFs between: 
# Human TFs list of Human Protein Atlas database: https://www.proteinatlas.org/search/transcription+factor
# Human TFs list of AnimalTFDB: # http://bioinfo.life.hust.edu.cn/AnimalTFDB/#!/download
# Human TFs list of Human Transcription Factors network 2018: # https://doi.org/10.1016/j.cell.2018.01.029
# Human TFs list of Uniprot database: https://www.uniprot.org/uniprotkb?facets=model_organism%3A9606%2Cexistence%3A2&query=transcription%20factor 
# Human TFs list of GeneCards database: https://www.genecards.org/Search/Keyword?queryString=%22human%20transcription%20factor%22
# Human Epifactor gene list: https://epifactors.autosome.org/description

# Import Transcription factor list
TFs_list <- read.xlsx(file.path(analysis_dir,"TFs_list_Final.xlsx"))
dim(TFs_list)
# Human epifactor and histone gene
Epifactor_list <- read.xlsx(file.path(analysis_dir,"Epifactor_Histone_list.xlsx"))
dim(Epifactor_list)
```

**3. Generate the SummarizedExperiment data**

We can generate the SummarizedExperiment object from our data using DEP package. We need a list of expression data with unique gene name and protein ID, and colname of data set. Futhermore, the experimental design must contain ‘label’, ‘condition’ and ‘replicate’ columns. The ‘label’ column contains the identifiers of the different samples and they should correspond to the column names containing the assay data. The ‘condition’ and ‘replicate’ columns contain the annotation of these samples as defined by the user.

**3.1 Preparation the SummarizedExperiment data**
```{r}

# Generate a SummarizedExperiment object using an experimental design
# colnames(catTFRE_MS_Round4)
# The dataset has unique Uniprot identifiers, however those are not immediately informative. The associated gene names are informative, however these are not always unique.
# Check if any duplicated gene symbol or ID
catTRFE_MS_C1$Gene_Symbol %>% 
  duplicated() %>% 
  any()
# List the duplicate if TRUE
catTRFE_MS_C1 %>% 
   group_by(Gene_Symbol) %>% 
   summarize(frequency = n()) %>% 
   arrange(desc(frequency)) %>% 
   dplyr::filter(frequency > 1) %>% head()

# For further analysis these proteins must get unique names. Additionally, some proteins do not have an annotated gene name and for those we will use the Uniprot ID.
## Make it unique incase duplicate exist, and add two column of ID and gene for the SE  step
catTFRE_MS_unique <- make_unique(catTRFE_MS_C1, "Gene_Symbol", "Protein_Symbol", delim = ";")
# Check the list is unique for further steps?
catTFRE_MS_unique$name %>% duplicated() %>% any()
catTFRE_MS_unique$Gene_Symbol %>% duplicated() %>% any()
# then make list of unique gene name and protein ID for further steps
catTFRE_MS_unique$Gene.names <- catTFRE_MS_unique$Gene_Symbol
catTFRE_MS_unique$Protein.IDs <- catTFRE_MS_unique$Protein_Symbol

catTFRE_MS_rawdata <- catTFRE_MS_unique
colnames(catTFRE_MS_rawdata)[1:12] <- c("HOS_C1_1",  "HOS_C1_2",  "HOS_C1_3",  "HOS_C1_4",  "MG63_C1_1", "MG63_C1_2", "MG63_C1_3", "MG63_C1_4", "U2OS_C1_1", "U2OS_C1_2", "U2OS_C1_3", "U2OS_C1_4")
write.xlsx(catTFRE_MS_rawdata, file = file.path(plots_dir, "catTFRE_C1_raw.xlsx"))

## Prepare the column number (should be string vector)
catTFRE_columns <- grep("#Spec.Sample", colnames(catTRFE_MS_C1))

## design the experiment
# Experiment design for Total with only 1 replication will not consider to put number in replication column. If you set up with any replication with only one, make_se function will not works
experimental_design <- data.frame(label = colnames(catTRFE_MS_C1)[grep("#Spec.Sample", colnames(catTRFE_MS_C1))] ,
                                  condition = c(rep("HOS_C1", 4), rep("MG63_C1", 4), rep("U2OS_C1", 4)),
                                  classification = c(rep("High aggressive", 4), rep("Low aggressive", 4),  rep("High aggressive", 4)),
                                  cell = c(rep("HOS", 4), rep("MG63", 4), rep("U2OS", 4)),
                                  replicate = c(1:4))
# kable(experimental_design)
# write.xlsx(experimental_design, file = "MS_experiment_design_Round4.xlsx")

# The wrapper function performs the full analysis
#' The _make_se_ and _make_se_parse_ functions generate a `r Biocpkg("SummarizedExperiment") ` object that has a couple of specifications.
#' The assay data is log2-transformed and its rownames depict the protein names.
#' The rowData contains, amongst others, the 'name' and 'ID' columns that were generated by _make_unique_.
#' The colData contains the experimental design and thereby the sample annotation.
#' Thereby the colData includes the 'label', 'condition' and 'replicate' columns as well as a newly generated 'ID' column.
#' The log2-transformed assay data and the specified rowData and colData columns are prerequisites for the subsequent analysis steps.
catTFRE_MS <- make_se(catTFRE_MS_unique, catTFRE_columns, experimental_design)

# Let's have a look at the SummarizedExperiment object
catTFRE_MS

# Convert to data frame with rownames as a column
df <- as.data.frame(assay(catTFRE_MS)) %>% tibble::rownames_to_column("Gene")
df[is.na(df)] <- 0
write.xlsx(df, file = file.path(plots_dir, "catTFRE_C1_normalized.xlsx"))
```


#### 3.3 Filter on missing values, Normalization, Differential enrichment analysis (DEA) based on linear models and empherical Bayes statistics
Filter options: look more at https://bioconductor.org/packages/release/bioc/vignettes/DEP/inst/doc/MissingValues.html
Filter for proteins based on stringent filtering: The dataset contains proteins which are not quantified in all replicates. Some proteins are even only quantified in a single replicate so need to filter on missing values
```{r, Filtering}
set.seed(1012)
# Plot a barplot of the protein identification overlap between samples
# plot_frequency(catTFRE_MS)
# plot_frequency(catTFRE_MS_TFs)
# Filter for proteins that are identified in all replicates of at least one condition
# data_filt <- filter_missval(catTFRE_MS, thr = 0) # Most stringent
# Filter for proteins that are identified in 2 out of 4 replicates of at least one condition
data_filt <- filter_missval(catTFRE_MS, thr = 0) # more stringent for proteins filtering
dim(catTFRE_MS)
dim(data_filt)

# Plot a barplot of the number of identified proteins per samples
# plot_numbers(catTFRE_MS)
# plot_numbers(data_filt)

# Plot intensity distributions and cumulative fraction of proteins with and without missing values
# plot_detect(data_filt)

# The data is background corrected and normalized by variance stabilizing transformation (vsn). Then the remaining missing values in the dataset need to be imputed before analyse the differential enrichment analysis (DEA)

# Normalize the data
data_norm <- normalize_vsn(data_filt)

# Visualize normalization by boxplots for all samples before and after normalization
# p1 <- meanSdPlot(data_norm, plot = FALSE)
# p1$gg + ggplot2::theme_bw()
# p2 <- meanSdPlot(data_norm, plot = FALSE)
# p2$gg + ggplot2::theme_bw()

# Visualize normalization by boxplots for all samples before and after normalization
# plot_normalization(data_filt, data_norm)

# Impute data for missing values
# Plot a heatmap of proteins with missing values
# plot_missval(data_filt)
# plot_missval(catTFRE_MS)

# Impute missing data using random draws from a Gaussian distribution centered around a minimal value (for MNAR)
data_imp <- impute(data_norm, fun = "MinProb", q = 0.01)

# Plot the data after normalization
# plot_numbers(data_norm)
# plot_numbers(data_imp)

# Plot intensity distributions and cumulative fraction of proteins with and without missing values
# plot_detect(data_filt)

# Protein-wise linear models combined with empirical Bayes statistics are used for the differential enrichment analysis (or differential expression analysis). The test_diff function introduced here uses limma and automatically generates the contrasts to be tested.

# Different enrichment possible comparisons of samples
#data_diff <- test_diff(data_imp, type = "all") # Test all_contrasts

# Set up manually defined comparisons
data_diff <- test_diff(data_imp, type = "manual",
                       test = c("HOS_C1_vs_MG63_C1", "U2OS_C1_vs_MG63_C1")) 

# Finally, significant proteins are defined by user-defined cutoffs using add_rejections.
# Denote significant proteins based on user defined cutoffs
dep <- add_rejections(data_diff, alpha = 0.05, lfc=1)

save(dep, file = file.path(plots_dir, "catTFRE_DifferentialExpressionProtein_C1_threshold0_MS_OS.rds"))
```

**4. Visualization of the results**
The results from the previous analysis can be easily visualized by a number of functions. These visualizations assist in the determination of the optimal cutoffs to be used, highlight the most interesting samples and contrasts, and pinpoint differentially enriched/expressed proteins.

**4.1 Plot the principal components analysis and correlation**
```{r}
load(file.path(plots_dir, "catTFRE_DifferentialExpressionProtein_C1_threshold0_MS_OS.rds"))
# plot PCA of total catTFRE MS protein
# pdf(file = file.path(plots_dir, "PCA_C1_MS_OS.pdf") , height = 5, width = 6)
# plot_pca(dep, x = 1, y = 2, n = 1000, point_size = 4)
# dev.off()
# plot PCA of TFs catTFRE MS protein
#pdf(file = file.path(plots_dir, "PCA_C1_TFs_Histone_MS_OS.pdf"), height = 5, width = 6)
#plot_pca(dep[row.names(dep) %in% c(TFs_list$Gene_Symbol, Epifactor_list$HGNC_symbol), ], 
#         x = 1, y = 2, n = 500, point_size = 4)
#dev.off()

# 1) Extract normalized expression matrix (e.g. "scaled" or "centered")
expr <- assay(dep) %>% t()     # samples as rows
# 2) Run PCA
pca_res <- prcomp(expr, scale. = FALSE)
# 3) Build data frame with metadata
pca_df <- data.frame(pca_res$x[, 1:2]) %>%
  tibble::rownames_to_column("sample") %>%
  left_join(colData(dep) %>% as.data.frame() %>% rownames_to_column("sample"), by = "sample")
# 4) % variance explained
var_expl <- round(100 * (pca_res$sdev^2 / sum(pca_res$sdev^2)), 1)
# 5) Custom colors for cells
cell_colors <- c("MG63"="#56B4E9", "U2OS"="#F66239", "HOS"="#8332FF")
# 6) Plot with ggplot2
ggplot(pca_df, aes(x = PC1, y = PC2, color = cell, shape = factor(replicate))) +
  geom_point(size = 4, alpha = 0.9) +
  scale_color_manual(values = cell_colors) +
  theme_minimal(base_size = 14) +
  labs(title = "PCA of DEP expression",
       x = paste0("PC1 (", var_expl[1], "%)"),
       y = paste0("PC2 (", var_expl[2], "%)")) +
  theme(legend.position = "right",
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6))
ggsave(file.path(plots_dir, "PCA_top500_ALL_MS_OS.pdf"), height = 4, width = 5.5)


# 1) Extract normalized expression matrix (e.g. "scaled" or "centered")
expr <- assay(dep[row.names(dep) %in% TFs_list$Gene_Symbol, ]) %>% t()     # samples as rows
# 2) Run PCA
pca_res <- prcomp(expr, scale. = FALSE)
# 3) Build data frame with metadata
pca_df <- data.frame(pca_res$x[, 1:2]) %>%
  tibble::rownames_to_column("sample") %>%
  left_join(colData(dep) %>% as.data.frame() %>% rownames_to_column("sample"), by = "sample")
# 4) % variance explained
var_expl <- round(100 * (pca_res$sdev^2 / sum(pca_res$sdev^2)), 1)
# 5) Custom colors for cells
cell_colors <- c("MG63"="#56B4E9", "U2OS"="#F66239", "HOS"="#8332FF")
# 6) Plot with ggplot2
ggplot(pca_df, aes(x = PC1, y = PC2, color = cell, shape = factor(replicate))) +
  geom_point(size = 4, alpha = 0.9) +
  scale_color_manual(values = cell_colors) +
  theme_minimal(base_size = 14) +
  labs(
    title = "PCA of DEP TF ONLY expression",
    x = paste0("PC1 (", var_expl[1], "%)"),
    y = paste0("PC2 (", var_expl[2], "%)")) +
  theme(legend.position = "right",
        panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6))
ggsave(file.path(plots_dir, "PCA_top500_TFs_MS_OS.pdf"), height = 4, width = 5.5)


# Correlation matrix
# Plot the Pearson correlation matrix of all protein
pdf(file = file.path(plots_dir, "Correlation_heatmap_All_MS_OS.pdf"), height = 6, width = 6)
plot_cor(dep, significant = TRUE, lower = 0, upper = 1, pal = "Reds")
dev.off()
# Plot the Pearson correlation matrix of TFs list only
pdf(file = file.path(plots_dir,"Correlation_heatmap_TFs_Histone_MS_OS.pdf"), height = 6, width = 6)
plot_cor(dep[row.names(dep) %in% c(TFs_list$Gene_Symbol, Epifactor_list$HGNC_symbol), ],
         significant = TRUE, lower = 0, upper = 1, pal = "Reds")
dev.off()
```

**4.2 Heatmap of all significant proteins**
```{r, heatmap}

# Plot a heatmap of all significant proteins with the data centered per protein
pdf(file = file.path(plots_dir,"Heatmap_OS_MS_C1_catTFRE-All_rownames.pdf"), height = 10) # starts writing a PDF to file
plot_heatmap(dep, 
             type = "centered", kmeans = TRUE, 
             k = 6, col_limit = 4, show_row_names = FALSE, #TRUE
             indicate = c("condition", "replicate", "classification")) -> dep_heatmap
dev.off() 

pdf(file = file.path(plots_dir,"Heatmap_OS_MS_C1-All_rownames_extend.pdf"), height = 50) # starts writing a PDF to file
plot_heatmap(dep, 
             type = "centered", kmeans = TRUE, 
             k = 6, col_limit = 4, show_row_names = TRUE, #TRUE
             indicate = c("condition", "replicate", "classification"))
dev.off() 

## Extract gene from heatmap cluster of interest
library(ComplexHeatmap)
heatmap_matrix <- dep_heatmap@ht_list[["log2 Centered intensity"]]@matrix #= matrix with RNAseq
# lapply(row_cluster_list, function(x) length(x))  #check/confirm size clusters
# loop to extract genes for each cluster.
gene_cluster <- lapply(names(row_order(dep_heatmap)), function(i){
  out <- data.frame(GeneID = rownames(heatmap_matrix[row_order(dep_heatmap)[[i]],]),
                    Cluster = paste0("cluster", sep="_", i),
                    stringsAsFactors = FALSE)
  return(out)}) %>%  #pipe (forward) the output 'out' to the function rbind to create 'gene_cluster'
  do.call(rbind, .)

gene_cluster <- gene_cluster %>% mutate(TF = if_else(GeneID %in% TFs_list$Gene_Symbol, "YES", "NO"))
# write the gene cluster to excell file
write.xlsx(gene_cluster, file.path(plots_dir,"Heatmap_MS_OS_C1_gene_cluster.xlsx"))

# Class them into new dataframe
tf_summary <- gene_cluster %>%
  group_by(Cluster, TF) %>%
  summarise(n = n(), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = TF, values_from = n, values_fill = 0) %>%
  mutate(Total = YES + NO,
         TF_percent = 100 * YES / Total)
#Plot
ggplot(gene_cluster, aes(x = Cluster, fill = TF)) +
  geom_bar(position = "stack") +
  # add counts
  geom_text(stat = "count",
            aes(label = ..count..),
            position = position_stack(vjust = 0.5),  # center inside each stacked segment
            color = "white",
            size = 3.5) +
  labs(title = "Distribution of TFs across clusters",
       x = "Cluster", y = "Number of genes") +
  scale_fill_manual(values = c("YES" = "#E53935", "NO" = "#9E9E9E"),
                    name = "TF status") +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave(file = file.path(plots_dir,"Heatmap_Cluster_gene_TF.pdf"), height = 4, width = 4)
# Call the gene name of cluster of interest
gene_cluster_5 <- gene_cluster[gene_cluster$Cluster=="cluster_5", ]$GeneID

```


**Alternative Volcano plot based on ggplot2 package, with option integrated volcano plot**
**4.3 Volcano plots of specific contrasts**
Volcano plots can be used to visualize a specific contrast (comparison between two samples). This allows to inspect the enrichment of proteins between the two samples (x axis) and their corresponding adjusted p value (y axis). This function is based on DEP package.

```{r, Volcano Plot}
# Load required libraries
library(dplyr)
# ===== Step 1: Classify expression status =====
prepare_data <- function(res) {
  res_df <- as.data.frame(res) %>%
    mutate(Expression = case_when(
      `p_value_-log10` >= 1.3 & log2_fold_change >= 1 ~ "Upregulated",
        `p_value_-log10` >= 1.3 & log2_fold_change <= -1 ~ "Downregulated",
          TRUE ~ "Not_significant"))
  return(res_df)}

# ===== Step 2: Create label dataframe for specific genes =====
prepare_labels <- function(df, genes_of_interest) {
  # Combine all genes into a single regex OR pattern (case-insensitive)
  pattern <- paste(genes_of_interest, collapse = "|") #Include all type of family
  #pattern <- paste0("(^|::)(", paste(genes_to_label, collapse = "|"), ")($|::)", collapse = "") #Exactly string name
  df %>% dplyr::filter(grepl(pattern, protein, ignore.case = TRUE)) %>%
    mutate(label_color = case_when(
      Expression == "Upregulated" ~ "red",
      Expression == "Downregulated" ~ "blue",
      TRUE ~ "grey"))}

# ===== Step 3: Plot volcano with custom gene labels & DEG count =====
library(ggplot2)
library(ggrepel)
library(ggtext)
plot_volcano <- function(res_df, label_df, title) {
  # Count number of DEGs
  num_up <- sum(res_df$Expression == "Upregulated")
  num_down <- sum(res_df$Expression == "Downregulated")
  # Subtitle with colored text using HTML
  subtitle_text <- paste0(
    "<span style='color:blue;'> Higher expression in MG63", ": ", num_down, "</span>",
    " | <span style='color:red;'> Higher expression in HOS", ": ", num_up, "</span>" )
  
  ggplot(res_df, aes(x = log2_fold_change, y = `p_value_-log10`)) +
    geom_point(aes(color = Expression), size = 1.5, alpha = 0.6) +
    scale_color_manual(values = c("Upregulated" = "red",
                                  "Downregulated" = "blue",
                                  "Not_significant" = "grey")) +
    # Only label significant genes (exclude grey labels)
    geom_label_repel(
      data = label_df %>% filter(label_color %in% c("red", "blue")),
      aes(label = protein),
      color = label_df %>% filter(label_color %in% c("red", "blue")) %>% pull(label_color),
      fill = "white",
      size = 3,
      box.padding = 0.5,
      point.padding = 0.0,
      min.segment.length = 0,
      segment.color = "gray50",
      segment.size = 0.4,
      max.overlaps = Inf,
      force = 5,
      force_pull = 0.3,
      direction = "both",
      nudge_y = 0.05,
      show.legend = FALSE) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
    theme_minimal() +
    labs(title = title,
         subtitle = subtitle_text,
         x = "Log2FoldChange",
         y = "-log10(padj)",
         color = "Expression") +
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_markdown(hjust = 0.5)  # Enable color in subtitle
    )
}

# Extract the DEP of comparision
HOS_MG63_C1 <- DEP::plot_volcano(dep,
                                 contrast = "HOS_C1_vs_MG63_C1",
                                 plot = FALSE, adjusted = FALSE)
# Prepare a volcano plot for the contrast "U2OS_C1_vs_MG63_C1""
U2OS_MG63_C1 <- DEP::plot_volcano(dep,
                                  contrast = "U2OS_C1_vs_MG63_C1",
                                  plot = FALSE, adjusted = FALSE)
# Set genes to label
genes_to_label <- c("RUNX", "SNAI", "NFI", "JUN", "FOX", "KRT6", "HMGA")

# Prepare each result
results_HOSMG63 <- prepare_data(HOS_MG63_C1)
labels1 <- prepare_labels(results_HOSMG63, genes_to_label)
library(openxlsx)
#write.csv(results_HOSMG63, file = file.path(plots_dir, "volcano_TOBIAS_HOS_vs_MG63.csv"))
write.xlsx(results_HOSMG63, file = file.path(plots_dir, "Diffenrential_Protein_Expression_HOS_vs_MG63_C1.xlsx"))
write.csv(results_HOSMG63, file = file.path(plots_dir, "Diffenrential_Protein_Expression_HOS_vs_MG63_C1.csv"))
# Plot and save to PDF
pdf(file = file.path(plots_dir, "volcano_plot_DEP_HOS_vs_MG63_C1.pdf"), width = 7, height = 5)
print(plot_volcano(results_HOSMG63, labels1, "Diffenrential Protein Expression in MG63 vs HOS"))
dev.off()

# Prepare each result
results_U2OSMG63 <- prepare_data(U2OS_MG63_C1)
labels2 <- prepare_labels(results_U2OSMG63, genes_to_label)
library(openxlsx)
#write.csv(results_HOSMG63, file = file.path(plots_dir, "volcano_TOBIAS_HOS_vs_MG63.csv"))
write.xlsx(results_U2OSMG63, file = file.path(plots_dir, "Diffenrential_Protein_Expression_U2OS_vs_MG63_C1.xlsx"))
write.csv(results_U2OSMG63, file = file.path(plots_dir, "Diffenrential_Protein_Expression_U2OS_vs_MG63_C1.csv"))
# Plot and save to PDF
pdf(file = file.path(plots_dir, "volcano_plot_DEP_U2OS_vs_MG63_C1.pdf"), width = 7, height = 5)
print(plot_volcano(results_U2OSMG63, labels2, "Diffenrential Protein Expression in MG63 vs U2OS"))
dev.off()

```


```{r, Integrate the volcano plot}
# Merge by protein
HOS_U2OS_C1 <- merge(x = results_HOSMG63, y = results_U2OSMG63, by = "protein", all = TRUE)
#result_HOS_U2OS_MG63 <- inner_join(res_HOSMG63, res_U2OSMG63,   by = c("output_prefix", "name", "motif_id", "cluster"), suffix = c("_HOS", "_U2OS")) %>%
#  select(output_prefix, matches("TOBIASscore|padj|Expression|change|name"))
# Rename columns for clarity
colnames(HOS_U2OS_C1) <- gsub("\\.x", "_HOS", colnames(HOS_U2OS_C1))
colnames(HOS_U2OS_C1) <- gsub("\\.y", "_U2OS", colnames(HOS_U2OS_C1))

# Create hit and label columns
HOS_U2OS_C1 <- HOS_U2OS_C1 %>%
  mutate(hit = ifelse((Expression_HOS %in% c("Upregulated", "Downregulated")) &
                        (Expression_U2OS %in% c("Upregulated", "Downregulated")), "Yes", "No"),
         labels = ifelse(hit == "Yes", protein, ""))


# Define your genes of interest
genes_to_label <- c("RUNX2", "SNAI2", "SIX4", "FOXO1", "CHD7", "HMGA2","COL11A1", "MYH14", "KRT80",
                    "CHD2","MYH11",
                    "JUNB", "COL14A1","NFIX", "NFIB", "PITX1", "NFIA", "JUND", 
                    "ZMYM3", "EPHA2")
# Filter for those genes
selected_genes <- HOS_U2OS_C1 %>%
  dplyr::filter(protein %in% genes_to_label) %>%
  mutate(label_color = case_when(
    log2_fold_change_HOS >= 1 | log2_fold_change_U2OS >= 1 ~ "red",       # Upregulated
    log2_fold_change_HOS <= -1 | log2_fold_change_U2OS <= -1 ~ "blue",    # Downregulated
    TRUE ~ "grey"  # not significant (won't be plotted)
  ))
# count table of expression
cluster_coords <- HOS_U2OS_C1 %>%
  dplyr::filter(Expression_HOS != "Not_significant",
         Expression_U2OS != "Not_significant") %>%   # exclude both axes
  group_by(Expression_HOS, Expression_U2OS) %>%
  summarise(x = max(log2_fold_change_HOS, na.rm = TRUE),
            y = max(log2_fold_change_U2OS, na.rm = TRUE),
            n = n(),
            .groups = "drop")

# Plot the HOS U2OS integrate
ggplot(HOS_U2OS_C1, aes(x = log2_fold_change_HOS, y = log2_fold_change_U2OS,
                        size = `p_value_-log10_HOS`, alpha = hit)) +
  geom_point(aes(colour = `p_value_-log10_U2OS`)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_vline(xintercept = 0, linetype = 2) +
  scale_colour_gradientn(colors = c( "white", "blue")) +
#  scale_colour_gradient( low = "white", high = "green", name = expression(-log[10]*"(pValueU2OS/MG63)")) +
  ggrepel::geom_label_repel(
    data = selected_genes %>% dplyr::filter(label_color %in% c("red", "blue")),
    aes(label = protein),
    color = selected_genes %>% dplyr::filter(label_color %in% c("red", "blue")) %>% pull(label_color),
    fill = "white",
    size = 3,
    box.padding = 0.5,
    point.padding = 0.6,
    min.segment.length = 0,
    segment.color = "gray50",
    segment.size = 0.4,
    max.overlaps = Inf,
    force = 5,
    force_pull = 0.3,
    direction = "both",
    nudge_y = 0.05,
    show.legend = FALSE) +
  geom_label(data = cluster_coords,
             aes(x = x, y = y, label = paste0("n =", n)),
             color = "black", fill = "white", size = 4, inherit.aes = FALSE) +
  labs(title = "Integration log2Foldchange of HOS and U2OS versus MG63 (Selected Genes)",
       x = expression("Log"[2]*"(Fold change HOS/MG63)"),
       y = expression("Log"[2]*"(Fold change U2OS/MG63)"),
       colour = expression(-log[10]*"(pValueU2OS/MG63)"),
       size = expression(-log[10]*"(pValueHOS/MG63)")) + 
  theme_classic()
ggsave(filename = file.path(plots_dir,"Integration_log2Foldchange_MS_OS_C1.pdf"), height = 6, width = 8)

```



```{r}
# Define your TF + Epifactor symbols
target_genes <- unique(c(TFs_list$Gene_Symbol, Epifactor_list$HGNC_symbol))

# Extract downregulated proteins that are TFs/Epifactors
results_HOSMG63_down <- results_HOSMG63 %>%
  filter(Expression == "Downregulated", protein %in% target_genes) %>%
  pull(protein)
results_U2OSMG63_down <- results_U2OSMG63 %>%
  filter(Expression == "Downregulated", protein %in% target_genes) %>%
  pull(protein)
# Find the intersection (common proteins)
common_downregulated <- intersect(results_HOSMG63_down, results_U2OSMG63_down)
# View result
common_downregulated

# Extract downregulated proteins that are TFs/Epifactors
results_HOSMG63_up <- results_HOSMG63 %>%
  filter(Expression == "Upregulated", protein %in% target_genes) %>%
  pull(protein)
results_U2OSMG63_up <- results_U2OSMG63 %>%
  filter(Expression == "Upregulated", protein %in% target_genes) %>%
  pull(protein)
# Find the intersection (common proteins)
common_upregulated <- intersect(results_HOSMG63_up, results_U2OSMG63_up)
# View result
common_upregulated
```


**4.5. Visualization of functional enrichment result using clusterProfile R package**
```{r, GSEA analysis}
# Load packages
collection <- as.data.frame(msigdbr_collections())
# 1. Load Hallmark gene sets from MSigDB #, subcollection = "KEGG_LEGACY"
library(clusterProfiler)
library(org.Hs.eg.db)
library(msigdbr)
library(dplyr)
library(enrichplot)

## 1. Load MSigDB collections ----

# Hallmark (H)
msig_h <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, entrez_gene)
# GO Biological Process (C5:BP)
msig_go_bp <- msigdbr(species = "Homo sapiens",
                      category = "C5",
                      subcollection = "GO:BP") %>%
  dplyr::select(gs_name, entrez_gene)
# KEGG pathways (C2:CP:KEGG)
msig_kegg <- msigdbr(species = "Homo sapiens",
                     category = "C2",
                     subcollection = "CP:KEGG_LEGACY") %>%
  dplyr::select(gs_name, entrez_gene)
# Reactome pathways (C2:CP:REACTOME)
msig_reactome <- msigdbr(species = "Homo sapiens",
                         category = "C2",
                         subcollection = "CP:REACTOME") %>%
  dplyr::select(gs_name, entrez_gene)

## 2. Prepare ranked gene list ----
# Assume you have DE results (protein = SYMBOL, log2FC)
results_HOS_MG63_df <- results_HOSMG63 %>%
  dplyr::filter(Expression != "Not_significant") %>%
  distinct(protein, .keep_all = TRUE)
# Map SYMBOL -> ENTREZ
gene_map <- bitr(results_HOS_MG63_df$protein,
                 fromType = "SYMBOL",
                 toType   = "ENTREZID",
                 OrgDb    = org.Hs.eg.db)
# Keep only successfully mapped
results_HOS_MG63_df <- results_HOS_MG63_df %>%
  dplyr::filter(protein %in% gene_map$SYMBOL)
# Named vector: log2FC with ENTREZID names
ranked_genes <- results_HOS_MG63_df$log2_fold_change
names(ranked_genes) <- gene_map$ENTREZID[match(results_HOS_MG63_df$protein,
                                               gene_map$SYMBOL)]
ranked_genes <- sort(ranked_genes, decreasing = TRUE)


## 3. Run GSEA ----
gsea_hallmark <- GSEA(geneList      = ranked_genes, TERM2GENE     = msig_h,
                      minGSSize     = 10, maxGSSize     = 500, pvalueCutoff  = 0.25,
                      verbose       = FALSE)
gsea_go_bp <- GSEA(geneList = ranked_genes, TERM2GENE = msig_go_bp, verbose = FALSE)
gsea_kegg  <- GSEA(geneList = ranked_genes, TERM2GENE = msig_kegg, verbose = FALSE)
gsea_reactome <- GSEA(geneList = ranked_genes, TERM2GENE = msig_reactome, verbose = FALSE)


## 4. Plot results ----
# Dotplot: Hallmark
dotplot(gsea_hallmark, showCategory = 10, split = ".sign") + 
  facet_wrap(~.sign)
dotplot(gsea_kegg, showCategory = 10, split = ".sign") + 
  facet_wrap(~.sign)
dotplot(gsea_go_bp, showCategory = 10, split = ".sign") + 
  facet_wrap(~.sign)
dotplot(gsea_reactome, showCategory = 10, split = ".sign") + 
  facet_wrap(~.sign)
# Ridgeplot (another nice visualization)
ridgeplot(gsea_hallmark, showCategory = 15)

# Enrichment plot for a specific pathway
gseaplot2(gsea_hallmark, geneSetID = "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION")

```



```{r, MSigDb analysis}
## MSigDb analysis
library(msigdbr)
MSigDb_list <- msigdbr(species = "Homo sapiens", category = "C5",
                      subcollection = "GO:BP") %>% # Human C6, oncogenic gene sets
  dplyr::select(gs_name, entrez_gene)
head(MSigDb_list)
MSigDb_HOS_MG63_C1 <- enricher(names(geneList_HOS_MG63_C1), TERM2GENE=MSigDb_list)
head(MSigDb_HOS_MG63_C1)
MSigDb_U2OS_MG63_C1 <- enricher(names(geneList_U2OS_MG63_C1), TERM2GENE=MSigDb_list)
head(MSigDb_U2OS_MG63_C1)

## Run the GSEA analysis for the list of protein expression 
HOS_MG63_C1_DO <- gseDO(geneList_HOS_MG63_C1)
head(HOS_MG63_C1_DO)
U2OS_MG63_C1_DO <- gseDO(geneList_U2OS_MG63_C1)
head(U2OS_MG63_C1_DO)
# running score and preranked list of GSEA result
# p1 <- gseaplot(U2OS_MG63_C1_DO, geneSetID = 1, by = "runningScore", title = U2OS_MG63_C1_DO$Description[1])
# p2 <- gseaplot(U2OS_MG63_C1_DO, geneSetID = 1, by = "preranked", title = U2OS_MG63_C1_DO$Description[1])
# p3 <- gseaplot(U2OS_MG63_C1_DO, geneSetID = 1, title = U2OS_MG63_C1_DO$Description[1])
# cowplot::plot_grid(p1, p2, p3, ncol=1, labels=LETTERS[1:3])
library(enrichplot)
# gseaplot2(U2OS_MG63_C1_DO, geneSetID = 1, title = U2OS_MG63_C1_DO$Description[1])
# gseaplot2(U2OS_MG63_C1_DO, geneSetID = 1:1, pvalue_table = TRUE)
## Running the Gene Set Enrichment Analysis (GSEA) based on gene list that extracted from heatmap

#### 
# https://bioinformatics-core-shared-training.github.io/cruk-summer-school-2021/RNAseq/Markdowns/12_Gene_set_testing.html
## Gene Set Enrichment Analysis (GSEA): require logFoldchange of gene list. then can filtering the LFC names(geneList)[abs(geneList) > 1.5]
# U2OS_MG63_C1_GO <- gseGO(geneList   = names(geneList_U2OS_MG63_C1),
#                         OrgDb      = org.Hs.eg.db, ont = "BP", #"CC" or "MF"
#                         minGSSize    = 100, maxGSSize    = 500,
#                         pvalueCutoff = 0.05, verbose      = FALSE )
# head(U2OS_MG63_C1_GO)
# Visualize the enrichment plot with one or multiple gene set
# gseaplot2(U2OS_MG63_C1_GO, geneSetID=1, title = U2OS_MG63_C1_GO$Description[1])
# gseaplot2(U2OS_MG63_C1_GO, geneSetID = 1:3, pvalue_table = TRUE,color = c("#E495A5", "#86B875", "#7DB0DD"))

# https://github.com/junjunlab/GseaVis/wiki/GseaVis-%E4%B8%BA%E4%BD%A0%E7%9A%84%E9%80%9A%E8%B7%AF%E5%9F%BA%E5%9B%A0%E6%B7%BB%E5%8A%A0%E7%83%AD%E5%9B%BE


## Barplots of a protein of interest
# Plot a barplot for RUNX2, FOXO1 and LRRFIP2
plot_single(dep, proteins = c("RUNX2", "FOXO1", "CHD7", "HMGA2")) #
# Plot a barplot for the protein USP15 with the data centered
plot_single(dep, proteins = "RUNX2", type = "centered")
plot_single(dep, proteins = "FOXO1", type = "centered")
plot_single(dep, proteins = "CHD7", type = "centered")
plot_single(dep, proteins = "HMGA2", type = "centered")

## Frequency plot of significant proteins and overlap of conditions
# Plot a frequency plot of significant proteins for the different conditions
# plot_cond(dep)
```

# 4.6 Results table
```{r}
## 
# Generate a results table
data_results <- get_results(dep)
# write.xlsx(data_results, file = "MS_Data_Results_Round4_sumC1C4_thr3.xlsx")

data_results_TFs <- get_results(dep[row.names(dep) %in% TFs_list$Gene_Symbol, ])
# write.xlsx(data_results_TFs, file = "MS_Data_Results_Round4_sumC1C4_TFsONLY_thr3.xlsx")

# Number of significant proteins
data_results_TFs %>% dplyr::filter(HOS_C1_vs_MG63_C1_significant) %>% nrow() # 28
data_results_TFs %>% dplyr::filter(U2OS_C1_vs_MG63_C1_significant) %>% nrow() # 22

# intersect((data_results_TFs %>% dplyr::filter(HOS_C1_vs_MG63_C1_significant))$name,
#           (data_results_TFs %>% dplyr::filter(U2OS_C1_vs_MG63_C1_significant))$name)


# Generate a data.frame from the resulting SummarizedExperiment object
# You might want to obtain an ordinary data.frame of the results. For this purpose, the package provides functions to convert SummarizedExperiment objects to data.frames. get_df_wide will generate a wide table, whereas get_df_long will generate a long table.

# Generate a wide data.frame
df_wide <- get_df_wide(dep)

# Generate a long data.frame
df_long <- get_df_long(dep)
```


```{r}
sessionInfo()
```


