---
title: "Osteosarcoma_TFnetworks using PANDAR (netZooR)"
author: "NguyenXuanThang"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
vignette: >
  %\VignetteIndexEntry{pandaR}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
## Introduction
The fundamental concepts behind the PANDA approach is to model the regulatory network as a bipartite network and estimate edge weights based on the evidence that information from a particular transcription factor *i* is successfully being passed to a particular gene *j*. 
 
### 1. Load packages
An example dataset derived from a subset of stress-induced Yeast is available by running
```{r setup, include=FALSE}
# devtools::install_github("netZoo/netZooR", build_vignettes = TRUE)
library("netZooR")     # load panda, condor, alpaca, lioness
library("data.table") # To load data 
library("openxlsx")
library("topGO")
library("dplyr")
library('fgsea')       # for enrichment analysis
library('ggplot2')     # for plotting
library('reshape2')    # to resize data frames
library('limma')       # to compute differential targeting
library('viridisLite') # plot communities
library('visNetwork')  # for network visualization
library('RCy3')
library("tibble")

root_dir <- rprojroot::find_root(rprojroot::has_dir("PANDA_sumC1C4"))
analysis_dir <- file.path(root_dir, "PANDA_sumC1C4")
plots_dir <- file.path(analysis_dir, "PANDA_plots")
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}
if (!dir.exists(analysis_dir)) {
  dir.create(analysis_dir, recursive = TRUE)
}
```

## 2. Data preparation 
The HOS function in pandaR is called with
1. prior A "prior" regulatory network of putative TF-gene interactions. Here, we will be working with a motif prior that we previously built by scanning known TF motifs to promoter regions in the human genome. In this file, the first column indicates the TF, the second column the target gene, and the third column a "1" indicating a putative interaction. Download at: https://sites.google.com/a/channing.harvard.edu/kimberlyglass/tools/resources
2. exp Gene expression data. Gene expression data and gene annotations were retrieved from MS data. This dataset contains gene expression data obtained from microarrays that were used to profile HOS and low_aggMG63rent brain cancer. These data were downloaded with the R package GEOquery and then normalized and log transformed. Rows indicate probe IDs and columns sample names.
3. ppi Potential protein-protein interactions (PPI) between TFs (optional). PPI data is optional in PANDA. Here, we downloaded data from StringDb, which contains scores in the range of [0,1000] indicating potential PPIs. We filtered these for TFs present in the motif prior and transformed scores to scores [0,1]. The first column indicates a TF, the second the TF it binds, and the third the interaction score.

```{r}
## Expression data: Rows indicate probe IDs and columns sample names.
# exp <- read.xlsx("catTRFR_MS_HOS_Round4_sumC1C4.xlsx", rowNames = FALSE, colNames = TRUE)
# To begin, we will calculate the standard deviation of the rows, and connect that to the original expression matrix:
# rowsd <- apply(exp[, 1:12], 1, sd)
# exp <- cbind(rowsd, exp)
# rm(rowsd)
# Let's sort the expression data based on the standard deviation of each probe, so that the most variable entries will be on the top.
# exp <- exp[order(exp$rowsd, decreasing=T),]
# In R, row names must be unique. Therefore, we need to remove duplicated genes from our expression matrix. we will do this by, for each gene, choosing to keep the probe ID that has the highest variance across samples. Now, continue remove duplicate gene symbols with low variance (gene name with low variance sorted bottom of list that duplicated will be removed)
# exp <- exp[!duplicated(exp$Gene_Symbol), ]
# Set Gene_Symbol as rownames
# rownames(exp) <- exp[, "Gene_Symbol"]
# nrow(exp)
# Remove colname not to be used
# exp <- exp %>% select(-rowsd, -Protein_Uniprot, -Protein_Symbol, -Gene_Symbol) # exp <- exp[, -c(1,13:15)]
# Change the colnames
# colnames(exp) <- sub("#Spec.Sample.", "",colnames(exp))
# dim(exp)
# Determine the number of non-NA/non-zero rows in the expression data. This is to be able to ensure that PANDA will have enough values in the vectors to calculate pearson correlations between gene expression profiles in the construction of the gene co-expression prior.
# zero_na_counts <- apply(exp, MARGIN = 1, FUN = function(x) length(x[(!is.na(x) & x!=0) ]))
# dim(exp[zero_na_counts > 0,])
# exp <- exp[zero_na_counts > 0,]
# write.table(exp, file = "protein_MS_spectral.txt" , sep = "\t", row.names = TRUE, col.names = TRUE)
# Load the expression data
# exp <- read.delim("protein_MS_spectral.txt", sep = "\t", header = TRUE)
# length(unique(rownames(exp)))


## motif dataset prior: the first column indicates the TF, the second column as the target gene, and the third column indicating a putative interaction.
# motif <- read.delim("prior_motif_hg38_refseq.txt", stringsAsFactors=FALSE, header=FALSE)
# Merge motif info and motif name
# motif_info <- read.delim("prior_motifinfo_Homo_sapiens.txt", stringsAsFactors=FALSE, header=TRUE)
# motif_info[, 1] <- sapply(strsplit(as.character(motif_info[, 1]), "\\_"), "[", 1)
# motif_info[, 2] <- sapply(strsplit(as.character(motif_info[, 2]), "\\_"), "[", 1)
# motif_info[, 2] <- gsub("\\*|\\(|\\)", "", motif_info[, 2])
# motif <- merge(motif_info, motif, by.x="MotifID", by.y="V1",  all.y =TRUE)
# motif <- motif[, -1]
# table(duplicated(motif[,1])) #dplyr::count() for column in dataframe, table for logical TRUE/FALSE
# write.table(motif, file = "prior_motif_hg38data.txt" , sep = "\t", row.names = FALSE, col.names = TRUE)
# motif <- read.delim("prior_motif_hg38data.txt", stringsAsFactors=FALSE, header=TRUE)


## Protein protein interaction data: The first column indicates a TF, the second are the TF it binds, and the third as the interaction score.
# https://string-db.org/cgi/download?sessionId=bPWghZV6jLBS&species_text=Homo+sapiens
# ppi <- read.delim("9606.protein.links.v12.0.txt", sep = "", stringsAsFactors = F)
# ppi_list <- read.delim("9606.protein.info.v12.0.txt", sep = "\t", stringsAsFactors = F)
# Add protein name
# ppi <- merge(ppi, ppi_list[, 1:2], by.x="protein1", by.y="X.string_protein_id", all=TRUE)
# ppi <- merge(ppi, ppi_list[, 1:2], by.x="protein2", by.y="X.string_protein_id", all=TRUE)
# select based on column name data
# ppi <- ppi %>% select(preferred_name.x, preferred_name.y, combined_score)
# colnames(ppi) <- c("protein1", "protein2", "score")
# ppi <- ppi[!is.na(ppi[, 1]), ]
# ppi <- ppi[!is.na(ppi[ ,2]), ]
# ppi[, 1] <- sapply(strsplit(as.character(ppi[, 1]), "\\_"), "[", 1)
# ppi[, 2] <- sapply(strsplit(as.character(ppi[, 2]), "\\_"), "[", 1)
# transformed scores [0:1000] to scores [0,1]
# ppi[ , 3] <- ppi[, 3]/1000
# How many unique TFs are present in the PPI data?
# length(unique(c(ppi[ , 1], ppi[ , 2])))
# write.table(ppi, file = "protein_interaction_STRING_All.txt" , sep = "\t", row.names = FALSE, col.names = TRUE)
# ppi <- read.delim("protein_interaction_STRING_All.txt" , sep = "\t", header = T, stringsAsFactors = F)

## In PANDA, row names from the gene expression dataset should match names of gene targets in the motif prior (second row in the motif prior). Here, we will pre-filter data before running PANDA, we subset our data to only include information common to both the motif prior and the gene expression data (same set of genes) and the motif prior and PPI (same set of TFs).
## Filtering the network that only included protein in the expression data
## How many genes are in common between the motif prior and expression data?
# length(intersect(motif[ , 1], rownames(exp)))
# length(intersect(motif[ , 2], rownames(exp)))
# motif <- motif[motif[ , 1]  %in% rownames(exp), ]
# motif <- motif[which(motif[, 2] %in% rownames(exp)),]
# length(unique(motif[, 1])) # unique TFs in motif data
# length(unique(motif[, 2])) # unique genes in motif data

## To do the matching we first find the gene names that are common to both the motif prior and expression data and then filter both to only contain information for these genes.
# exp <- exp[which(rownames(exp) %in% motif[ , 2]),]
# length(unique(rownames(exp)))

## Protein-protein interaction data
## Filtering the network that only included protein in the expression data
# ppi <- ppi[ppi[ , 1]  %in% rownames(exp), ]
# ppi <- ppi[ppi[ , 2]  %in% rownames(exp), ]
# dim(ppi)
# write.table(ppi, file = "protein_interaction_MassSpec_filtering.txt" , sep = "\t", row.names = FALSE, col.names = TRUE)

## Now do the same for TFs in the motif prior and PPI data. How many TFs are in common? remember that ppi data is TFs-TFs, so that both column 1 and 2 will be TFs
# length(intersect(unique(motif[, 1]), unique(c(ppi[, 1], ppi[, 2])))) # 258
# matching TFs
# commonTFs <- intersect(motif[, 1], c(ppi[, 1], ppi[, 2]))
# motif <- motif[which(motif[, 1] %in% commonTFs),]
# ppi <- ppi[which(ppi[, 1] %in% commonTFs & ppi[, 2] %in% commonTFs),]
# length(unique(ppi[, 1]))
# length(unique(ppi[, 2]))
# write.table(ppi, file = "protein_interaction_MassSpec_ppi_filtering.txt" , sep = "\t", row.names = FALSE, col.names = TRUE)

ppi <- read.delim(file.path(analysis_dir, "protein_interaction_MassSpec_ppi_filtering.txt") , sep = "\t", header = T, stringsAsFactors = F)
# ppi <- read.delim("protein_interaction_MassSpec_filtering.txt" , sep = "\t", header = T, stringsAsFactors = F)


# write.table(motif, file = "prior_motif_MassSpec_filtering.txt" , sep = "\t", row.names = FALSE, col.names = TRUE)
motif <- read.delim(file.path(analysis_dir, "prior_motif_MassSpec_filtering.txt"), stringsAsFactors=FALSE, header=TRUE)

# write.table(exp, file = "protein_MS_spectral_filtering.txt" , sep = "\t", row.names = TRUE, col.names = TRUE)
# exp <- read.delim("protein_MS_spectral_filtering.txt", sep = "\t", header = TRUE)
# Then, we log transform the protein expression
# exp <- log2(exp+1)
# write.table(exp, file = "protein_MS_spectral_filtering_log2.txt" , sep = "\t", row.names = TRUE, col.names = TRUE)
exp <- read.delim(file.path(analysis_dir, "protein_MS_spectral_filtering_log2.txt"), sep = "\t", header = TRUE)

# exp[rownames(exp)=="RUNX2",]

```

## 3. Running PANDA on OS cell type HOS, U2OS and MG63
In the previous module, we modeled one aggregate regulatory network on the entire expression dataset. Here, we will focus on comparative network analysis between (1) two aggregate networks; and (2) single-sample networks. To do so, we need to start looking at the clinical data. Clinical features associated with our expression dataset can be found in the clin object.

#### 3.1. Data classification
```{r PANDA, include=FALSE}

# extract column of expression data order by cell type
exp_HOS <- exp[ , grepl("HOS", colnames(exp))] 
zero_na_counts <- apply(exp, MARGIN = 1, FUN = function(x) length(x[(!is.na(x) & x!=0) ]))
exp_HOS <- exp_HOS[zero_na_counts >= 4,]
motif_HOS <- motif[motif[ , 1]  %in% rownames(exp_HOS), ]
motif_HOS <- motif_HOS[which(motif_HOS[, 2] %in% rownames(exp_HOS)),]
## Filtering the network that only included protein in the expression data
ppi_HOS <- ppi[ppi[ , 1]  %in% rownames(exp_HOS), ]
ppi_HOS <- ppi_HOS[ppi_HOS[ , 2]  %in% rownames(exp_HOS), ]
## Then run PANDA on both datasets with default parameters. 
net_HOS <- panda(motif_HOS, exp_HOS, ppi_HOS, progress = T) #39 steps, ran PANDA on 4265 Genes and 219 TFs, with time elapsed:269.36seconds
RegNet_HOS <- slot(net_HOS, "regNet") # Successfully ran PANDA on 50 steps, 5891 Genes and 5875 TFs with time elapsed:17874.27seconds 16296.24seconds. (5 hours). 

## U2OS
exp_U2OS <- exp[ , grepl("U2OS", colnames(exp))]
exp_U2OS <- exp_U2OS[zero_na_counts >= 4,]
motif_U2OS <- motif[motif[ , 1]  %in% rownames(exp_U2OS), ]
motif_U2OS <- motif_U2OS[which(motif_U2OS[, 2] %in% rownames(exp_U2OS)),]
## Filtering the network that only included protein in the expression data
ppi_U2OS <- ppi[ppi[ , 1]  %in% rownames(exp_U2OS), ]
ppi_U2OS <- ppi_U2OS[ppi_U2OS[ , 2]  %in% rownames(exp_U2OS), ]
## Then run PANDA on both datasets with default parameters. 
net_U2OS <- panda(motif_U2OS, exp_U2OS, ppi_U2OS, progress = T) #39 steps, ran PANDA on 4985 Genes and 200 TFs, with time elapsed:269.36seconds
RegNet_U2OS <- slot(net_U2OS, "regNet") 

## U2OS
exp_MG63 <- exp[ , grepl("MG63", colnames(exp))] 
exp_MG63 <- exp_MG63[zero_na_counts >= 4,]
motif_MG63 <- motif[motif[ , 1]  %in% rownames(exp_MG63), ]
motif_MG63 <- motif_MG63[which(motif_MG63[, 2] %in% rownames(exp_MG63)),]
## Filtering the network that only included protein in the expression data
ppi_MG63 <- ppi[ppi[ , 1]  %in% rownames(exp_MG63), ]
ppi_MG63 <- ppi_MG63[ppi_MG63[ , 2]  %in% rownames(exp_MG63), ]
## Then run PANDA on both datasets with default parameters. 
net_MG63 <- panda(motif_MG63, exp_MG63, ppi_MG63, progress = T) #39 steps, ran PANDA on 4627 Genes and 208 TFs, with time elapsed:433.48seconds.
RegNet_MG63 <- slot(net_MG63, "regNet") 

# Save an object to a file: save() with rda extension will contain multiple file, saveRDS with rds/RData extension will be single one
save(net_HOS, RegNet_HOS, net_U2OS, RegNet_U2OS, net_MG63, RegNet_MG63, ppi, motif, exp, file = file.path(analysis_dir, "PANDA_eachcelltype_filtering_celltype_new.rda"))
```

#### 3.2 Comparing two PANDA networks
Now we are ready to compare the two PANDA networks. The panda() function in netZooR outputs each network of a matrix, with TFs in rows and their target genes in columns. 
```{r COMPARE}
# Restore the object
load(file.path(analysis_dir, "PANDA_eachcelltype_filtering_celltype_new.rda"))
# To easily visualize differential edges and the TFs and genes connected to them, we will transform this into an edgelist using the code below. 
tfnamerep_HOS <- rep(row.names(RegNet_HOS), ncol(RegNet_HOS)) 
genenamerep_HOS <- rep(colnames(RegNet_HOS), each=nrow(RegNet_HOS)) 
# We first combine the TF and gene names in the same order as the c() function in R unlists matrices, and convert that into a data.frame.
edgelist_HOS <- cbind(tfnamerep_HOS, genenamerep_HOS) 
edgelist_HOS <- as.data.frame(edgelist_HOS)
# Now, we are ready to attach the edges in the HOS and low_aggMG63rent networks to this edge list. 
edgelist <- cbind(edgelist_HOS, c(RegNet_HOS), c(RegNet_U2OS), c(RegNet_MG63)) 
colnames(edgelist) <- c("TF","Gene","HOS", "U2OS","MG63")

# then we will select edges that are both likely to be "real" in a particular network (i.e. have a high edge weight) as well as "different" between the two networks. Because the weights of PANDA edges can be thought of as z-score units, edgeweights can be easily converted into probability values.
p_high_aggHOS <- pnorm(RegNet_HOS) 
p_high_aggU2OS <- pnorm(RegNet_U2OS) 
p_low_aggMG63 <- pnorm(RegNet_MG63) 
# then select edges based on this combined probability score. In our example, we use the PANDA output matrices to select edges that have at least a 75% chance of being both real and different:
high_aggHOS_spec <- p_high_aggHOS*pnorm(RegNet_HOS - RegNet_MG63) > 0.75
high_aggU2OS_spec <- p_high_aggU2OS*pnorm(RegNet_U2OS - RegNet_MG63) > 0.75

low_aggMG63_spec1 <- p_low_aggMG63*pnorm(RegNet_MG63 - RegNet_HOS) > 0.75
low_aggMG63_spec2 <- p_low_aggMG63*pnorm(RegNet_MG63 - RegNet_U2OS) > 0.75
# We can then port these to our edgelist as follows: 
high_aggHOS_spec_edges <- edgelist[which(c(high_aggHOS_spec)) , ] 
low_aggMG63_spec_edges1 <- edgelist[which(c(low_aggMG63_spec1)) , ]

high_aggU2OS_spec_edges <- edgelist[ which(c(high_aggU2OS_spec)) ,] 
low_aggMG63_spec_edges2 <- edgelist[which(c(low_aggMG63_spec2)) ,]


# Visualize these tumor type-specific edges on top of the scatterplot you made in Q6. Hint: use point() to overlay points of different colors on the scatterplot. 
png(file=file.path(plots_dir, "DEGs_HOS_versus_MG63_cell_type_edges.png"), height = 1600, width = 1600, res = 300)
plot(edgelist[,c(3,5)], pch = 16) + 
  points(high_aggHOS_spec_edges[ , c(3,5)], pch = 16, col="#8332FF") +
  points(low_aggMG63_spec_edges1[ , c(3,5)], pch = 16, col="#56B4E9") +
  title(main="Cell type-specific edges on top of HOS versus MG63")
  legend("topright", 
       legend = c(paste0("U2OS-specific: ", length(high_aggHOS_spec_edges$HOS)),
                  paste0("MG63-specific: ", length(low_aggMG63_spec_edges1$MG63))),
       col = c("#8332FF", "#56B4E9"), pch = 16, bty = "o")
dev.off()
# How many edges are specific to the HOS network? And to the MG63 network?
length(unique(edgelist$TF)) #258
length(unique(edgelist$Gene)) #5891

# Visualize these tumor type-specific edges on top of the scatterplot, use point() to overlay points of different colors on the scatterplot. 
png(file=file.path(plots_dir, "DEGs_U2OS_versus_MG63_cell_type_edges.png"), height = 1600, width = 1600, res = 300)
plot(edgelist[,4:5], pch = 16) + 
  points(high_aggU2OS_spec_edges[ , 4:5], pch = 16, col="#F66239") +
  points(low_aggMG63_spec_edges2[ , 4:5], pch = 16, col="#56B4E9") +
  title(main="Cell type-specific edges on top of U2OS versus MG63")
  legend("topright", 
       legend = c(paste0("U2OS-specific: ", length(high_aggU2OS_spec_edges$U2OS)),
                  paste0("MG63-specific: ", length(low_aggMG63_spec_edges2$MG63))),
       col = c("#F66239", "#56B4E9"), pch = 16, bty = "o")
dev.off()
# Let's take a look at the top differential edges.
length(unique(edgelist$TF)) #258
length(unique(edgelist$Gene)) #5891



# What is the TF that makes most specific interactions in each of these networks? 
# table(high_aggHOS_spec_edges$TF)[order(table(high_aggHOS_spec_edges$TF), decreasing = TRUE)] # MBD2/MECP2 makes most interactions, 1793/1787
# table(low_aggMG63_spec_edges1$TF)[order(table(low_aggMG63_spec_edges1$TF), decreasing = TRUE)] # SP3/ZBTB7B makes most interactions, 2020/2007

# table(high_aggU2OS_spec_edges$TF)[order(table(high_aggU2OS_spec_edges$TF), decreasing = TRUE)] # EGR1 makes most interactions, 2231
# table(low_aggMG63_spec_edges2$TF)[order(table(low_aggMG63_spec_edges2$TF), decreasing = TRUE)] # MBD2 makes most interactions, 1788

# How many genes does it target? Do you recognize any of these genes?
# RUNX2_HOS_target <- edgelist[which(c(high_aggHOS_spec)),][which(edgelist[which(c(high_aggHOS_spec)),]$TF=="RUNX2"),]$Gene
# RUNX2_MG63_target1 <- edgelist[which(c(low_aggMG63_spec1)),][which(edgelist[which(c(low_aggMG63_spec1)),]$TF=="RUNX2"),]$Gene

# RUNX2_U2OS_target <- edgelist[which(c(high_aggU2OS_spec)),][which(edgelist[which(c(high_aggU2OS_spec)),]$TF=="RUNX2"),]$Gene
# RUNX2_MG63_target2 <- edgelist[which(c(low_aggMG63_spec2)),][which(edgelist[which(c(low_aggMG63_spec2)),]$TF=="RUNX2"),]$Gene

```


#### 4 Comparing agressive TFs between two PANDA networks
#### 4.1 Preparing data to plot
```{r}
load(file.path(analysis_dir, "PANDA_eachcelltype_filtering_celltype_new.rda"))
# We will use the function vis.panda.in.cytoscape to plot a set of nodes and edges on Cytoscape. The input for this function is a data.frame of edges to plot with 4 columns: "tf", "gene", "motif" (TF motif present or not on gene promoter), "force" (edge weight calculated by PANDA).
HOS_vis <- reshape2::melt(net_HOS@regNet)
U2OS_vis <- reshape2::melt(net_U2OS@regNet)
MG63_vis  <- reshape2::melt(net_MG63@regNet)

HOS_vis <- data.frame("TF"=as.character(HOS_vis[,1]),
                      "Gene"=as.character(HOS_vis[,2]),
                      "Score"=as.numeric(HOS_vis[,3]),
                      "Motif"=NA, stringsAsFactors = FALSE)
U2OS_vis <- data.frame("TF"=as.character(U2OS_vis[,1]),
                      "Gene"=as.character(U2OS_vis[,2]),
                      "Score"=as.numeric(U2OS_vis[,3]),
                      "Motif"=NA, stringsAsFactors = FALSE)
MG63_vis <- data.frame("TF"=as.character(MG63_vis[,1]),
                       "Gene"=as.character(MG63_vis[,2]),
                       "Score"=as.numeric(MG63_vis[,3]),
                       "Motif"=NA, stringsAsFactors = FALSE)
head(HOS_vis)
head(U2OS_vis)
head(MG63_vis)

# Load all TF and Epigenetics database
TF_list <- read.xlsx(file.path(analysis_dir, "TFs_list_Final.xlsx"))
Epi_list <- read.csv(file.path(analysis_dir, "Epifactor_Histone_list.csv"))
# DEP_gene_list <- read.table("All_diff_gene_cluster_DEP_C1C4.txt", header = TRUE)
DEP_gene_list <- read.xlsx(file.path(analysis_dir, "Heatmap_MS_OS_C1_C4_NE_gene_cluster.xlsx"))


# Filtering the data of TFs and Epigenetic factors
# HOS_vis_TF <- HOS_vis[HOS_vis$TF %in% DEP_gene_list$GeneID, ]
HOS_vis_TF <- HOS_vis %>% filter(TF %in% DEP_gene_list$GeneID, 
                                 Gene %in% DEP_gene_list$GeneID )

#HOS_vis_TF <- HOS_vis %>% filter(TF %in% c(TF_list$Gene_Symbol), 
#                                 Gene %in% c(TF_list$Gene_Symbol) )
table(HOS_vis_TF$TF)
table(HOS_vis_TF$Gene)
#U2OS_vis_TF <- U2OS_vis[U2OS_vis$TF %in% DEP_gene_list$GeneID, ]
U2OS_vis_TF <- U2OS_vis %>% filter(TF %in% DEP_gene_list$GeneID, 
                                 Gene %in% DEP_gene_list$GeneID )

#U2OS_vis_TF <- U2OS_vis %>% filter(TF %in% c(TF_list$Gene_Symbol), 
#                                 Gene %in% c(TF_list$Gene_Symbol) )
table(U2OS_vis_TF$TF)
table(U2OS_vis_TF$Gene)
table(HOS_vis_TF$Gene)
# MG63_vis_TF <- MG63_vis[MG63_vis$TF %in% DEP_gene_list$GeneID, ]
MG63_vis_TF <- MG63_vis %>% filter(TF %in% DEP_gene_list$GeneID, 
                                 Gene %in% DEP_gene_list$GeneID )

#MG63_vis_TF <- MG63_vis %>% filter(TF %in% c(TF_list$Gene_Symbol), 
#                                 Gene %in% c(TF_list$Gene_Symbol) )
table(MG63_vis_TF$TF)
table(MG63_vis_TF$Gene)

```


#### 4.2. Plot the highest edge weights each cell types
```{r}
createPandaStyle <- function(style_name = "PandaStyle",
                             weight_col = "weight_transformed",
                             group_col  = "group",
                             interact_col = "interaction") {
  # If style already exists, delete to avoid stale/partial mappings
  if (style_name %in% getVisualStyleNames()) {
    deleteVisualStyle(style_name)
  }
  # --- Node mappings (expects node table has 'group' with values TF/Gene) ---
  nodeShape <- mapVisualProperty('NODE_SHAPE', group_col, 'd', c("TF","Gene"), c("ELLIPSE","RECTANGLE"))
  nodeFill <- mapVisualProperty('NODE_FILL_COLOR', group_col, 'd', c("TF","Gene"), c('#FD7622', '#499df3'))
  nodeBorder <- mapVisualProperty('NODE_BORDER_PAINT', group_col, 'd', c("TF","Gene"), c('#FD7622', '#499df3'))
  nodeLabel <- mapVisualProperty('NODE_LABEL', 'id', 'p')
  # --- Edge mappings (expects edge table has 'interaction' and numeric weight_col) ---
  # Keep the type consistent: use strings "1","0" for both mappings
  edgeLine <- mapVisualProperty('EDGE_LINE_TYPE', interact_col, 'd', c("1","0"), c("SOLID","LONG_DASH"))
  edgeArrow <- mapVisualProperty('EDGE_TARGET_ARROW_SHAPE', interact_col, 'd', c("1","0"), c("DELTA","NONE"))  # Or DELTA/DELTA if you want arrows on both
  # Continuous mappings on weight
  edgeColor <- mapVisualProperty('EDGE_STROKE_UNSELECTED_PAINT', weight_col, 'c', c(0, 10000), c('#d3d3d3', '#000000'))
  edgeWidth <- mapVisualProperty('EDGE_WIDTH', weight_col, 'c', c(0, 10000), c(0.5, 7))
  defaults <- list(NODE_SIZE           = 50,
                   EDGE_TRANSPARENCY   = 255,
                   NODE_LABEL_POSITION = "center")
  createVisualStyle(style.name = style_name,
                    defaults   = defaults,
                    mappings   = list(nodeShape, nodeFill, nodeBorder, nodeLabel, edgeLine, edgeArrow, edgeColor, edgeWidth))
  # Apply to current network
  setVisualStyle(style_name)
}
```

```{r}

# Plot the 1000 highest edge weights
n = 500 # number of edges to plot 
#HOS
top_HOS <- order(HOS_vis$Score,decreasing=T)[1:n]
HOS_vis_top <- HOS_vis[top_HOS,]
#U2OS
top_U2OS <- order(U2OS_vis$Score,decreasing=T)[1:n]
U2OS_vis_top <- U2OS_vis[top_U2OS,]
# MG63
top_MG63 <- order(MG63_vis$Score,decreasing=T)[1:n]
MG63_vis_top <- MG63_vis[top_MG63,]
## Plot in cytoscape (open Cytoscape before running this command)
cytoscapePing()
cytoscapeVersionInfo()
visPandaInCytoscape(HOS_vis_top, network_name="HOS_500")
visPandaInCytoscape(U2OS_vis_top, network_name="U2OS_500")
visPandaInCytoscape(HOS_vis_top, network_name="MG63_500")


nrow(HOS_vis_TF) #1573
nrow(U2OS_vis_TF) #1573
nrow(MG63_vis_TF)#1573
unique(HOS_vis_TF$TF)


n = 500 # number of edges to plot 
#HOS
top_HOS <- order(HOS_vis_TF$Score,decreasing=T)[1:n]
HOS_vis_top <- HOS_vis[top_HOS,]
#U2OS
top_U2OS <- order(U2OS_vis_TF$Score,decreasing=T)[1:n]
U2OS_vis_top <- U2OS_vis[top_U2OS,]
# MG63
top_MG63 <- order(MG63_vis_TF$Score,decreasing=T)[1:n]
MG63_vis_top <- MG63_vis[top_MG63,]
## Plot in cytoscape (open Cytoscape before running this command)
cytoscapePing()
cytoscapeVersionInfo()
visPandaInCytoscape(HOS_vis_top, network_name="HOS_TFs_500")
visPandaInCytoscape(U2OS_vis_top, network_name="U2OS_TFs_500")
visPandaInCytoscape(MG63_vis_top, network_name="MG63_TFs_500")

```

#### Copy visDiffPandaInCytoscape() and createDiffPandaStyle() function from github.
```{r, include=FALSE}
# https://github.com/netZoo/netZooR/blob/master/R/visDiffPandaInCytoscape.R
visDiffPandaInCytoscape <- function(merged_panda, condition_name = "cond.1", network_name="diff.PANDA"){
  cytoscapePing()
  cytoscapeVersionInfo()
  #Plot
  colnames(merged_panda) <- c("source","target","motif","weight",condition_name)
  merged_panda$weight_transformed <- rank(merged_panda$weight)/length(merged_panda$weight)*10000
  
  if(nrow(merged_panda)>6666){
    message("Cytoscape can get slow for networks >20k objects. Consider reducing network size.")
  }
  panda_nodes <- data.frame(
    id    = c(merged_panda$source, merged_panda$target),
    group = rep(c("TF","Gene"), each=nrow(merged_panda)),
    stringsAsFactors=FALSE)
  # Create the network
  createNetworkFromDataFrames(nodes=panda_nodes, edges=merged_panda,
                              title=network_name, collection="DataFrame Example")
  # 🔑 Apply your custom style (must exist)
  style_name <- "Diff.PandaStyle"
  if (!(style_name %in% getVisualStyleNames())) {
    createDiffPandaStyle(style_name = style_name, condition_name = condition_name)
  }
  setVisualStyle(style_name)
}
# Plot the single network https://github.com/netZoo/netZooR/blob/master/R/createPandaStyle.R
# Plot the differential network https://github.com/netZoo/netZooR/blob/master/R/createDiffPandaStyle.R
createDiffPandaStyle <- function(style_name="Diff.PandaStyle", condition_name="cond.1"){
  # node properties
  nodeShape <- mapVisualProperty('node shape','group','d',
                                 c("TF","Gene"),
                                 c("ELLIPSE","RECTANGLE"))
  nodeColor <- mapVisualProperty('node fill color', 'group','d', 
                                 c("TF","Gene"), 
                                 c('#FD7622', '#499df3'))
  nodeBorderColor <- mapVisualProperty("Node Border Paint", 'group', 'd', 
                                       c("TF","Gene"), 
                                       c('#FD7622', '#499df3'))
  nodeLabel <- mapVisualProperty("node label","id",'p')
  # edge properties
  edgeLineType <- mapVisualProperty('edge line type','motif','d',c(1,0), 
                                    c("SOLID","LONG_DASH"))
  edgeTargetArrowShape <- mapVisualProperty("edge target arrow shape",'motif','d',c(1,0),c("DELTA","DELTA"))
  edgeTargetArrowUnSelectedPaint <- mapVisualProperty("Edge Target Arrow Unselected Paint",
                                                      condition_name,'d',c("T","F"), 
                                                      c('#92278f', '#60bf71'))
  # edge width and edge color shade both represent the edge weight.
  edgeWidth <- mapVisualProperty("edge width", "weight_transformed",'c', c(0,10000), c(0.5,7))
  edgeStrokeUnselectedPaint <- mapVisualProperty('Edge Stroke Unselected Paint',
                                                 condition_name,'d',c("T","F"), 
                                                 c('#92278f', '#60bf71'))
  # default setting
  defaults <- list(NODE_SIZE=50,
                   EDGE_TRANSPARENCY=255,
                   NODE_LABEL_POSITION="center")
  # create visual style working for PANDA network
  createVisualStyle(style_name,defaults,list(nodeShape, nodeColor, nodeBorderColor,nodeLabel,
                                             edgeLineType, edgeTargetArrowShape, edgeTargetArrowUnSelectedPaint,
                                             edgeWidth, edgeStrokeUnselectedPaint))
  # ✅ Automatically apply it to the current network
  setVisualStyle(style_name)
}


```


#### 4.3. Plot the highest differential networks betweeen TF High Agressive (HOS/U2OS) versus Low Agressive (MG63)
```{r}
## Plot differential networks betweeen HOS (High Agressive) versus MG63 (Low Agressive)
# Select the top differential edge weights betweeen HOS and MG63
diffRes_HOS_MG63_TF <- pandaDiffEdges(HOS_vis_TF, MG63_vis_TF, threshold=0.8, condition_name="High_Agressive")
## Plot differential networks betweeen U2OS (High Agressive) versus MG63 (Low Agressive)
diffRes_U2OS_MG63_TF <- pandaDiffEdges(U2OS_vis_TF, MG63_vis_TF, threshold=0.8, condition_name="High_Agressive")

# Number of differential edges is:
nrow(diffRes_HOS_MG63_TF) # 19623 #108
nrow(diffRes_U2OS_MG63_TF) # 20022  #123

#  Apply the style to the network. Purple edges indicate higher edge weight in the defined "condition_name" parameter (HOS in our example), and green edges indicate higher edge weight in the other condition (MG63 in our example).
# createDiffPandaStyle(style_name="Diff.PandaStyle", condition_name="High_Agressive")

# Select the top differential edges higher in HOS to plot in Cytoscape
n=200 # number of edges to select from each condition
diffResHOS_TF <- diffRes_HOS_MG63_TF[diffRes_HOS_MG63_TF$High_Agressive=="T",]
diffResHOS_TF <- diffResHOS_TF[order(diffResHOS_TF$Score,decreasing=TRUE),][1:n,]
# Select the top differential edges higher in MG63 to plot in Cytoscape
diffResMG63_TF1 <- diffRes_HOS_MG63_TF[diffRes_HOS_MG63_TF$High_Agressive=="F",]
diffResMG63_TF1 <- diffResMG63_TF1[order(diffResMG63_TF1$Score,decreasing=TRUE),][1:n,]
unique(diffResHOS_TF$TF)
unique(diffResMG63_TF1$TF)
# Combine top differential edges in HOS and MG63 to plot in Cytoscape
diffRes_vis_TF1 <- rbind(diffResHOS_TF, diffResMG63_TF1)
unique(diffRes_vis_TF1$TF)
# Plot the network
visDiffPandaInCytoscape(diffRes_vis_TF1, condition_name = "High_Agressive", network_name="HOS_MG63_sumC1C4diff_200_TF.PANDA")


# Select the top differential edges higher in HOS to plot in Cytoscape
n=200 # number of edges to select from each condition
diffResU2OS_TF <- diffRes_U2OS_MG63_TF[diffRes_U2OS_MG63_TF$High_Agressive=="T",]
diffResU2OS_TF <- diffResU2OS_TF[order(diffResU2OS_TF$Score,decreasing=TRUE),][1:n,]
# Select the top differential edges higher in MG63 to plot in Cytoscape
diffResMG63_TF2 <- diffRes_U2OS_MG63_TF[diffRes_U2OS_MG63_TF$High_Agressive=="F",]
diffResMG63_TF2 <- diffResMG63_TF2[order(diffResMG63_TF2$Score,decreasing=TRUE),][1:n,]
unique(diffResU2OS_TF$TF)
unique(diffResMG63_TF2$TF)
# Combine top differential edges in HOS and MG63 to plot in Cytoscape
diffRes_vis_TF2 <- rbind(diffResU2OS_TF, diffResMG63_TF2)
unique(diffRes_vis_TF2$TF)
# Plot the network
visDiffPandaInCytoscape(diffRes_vis_TF2, condition_name = "High_Agressive", network_name="U2OS_MG63_sumC1C4diff_200_Filtering.PANDA")

# Combine all HOS/MG63 and U2OS/MG63
diffRes_vis_ALL <- rbind(diffResU2OS_TF, diffResHOS_TF, diffResMG63_TF2)
unique(diffRes_vis_ALL$TF)
diffRes_vis_ALL <- diffRes_vis_ALL[!duplicated(c("TF", "Gene", "High_Agressive")), ]
# Plot the network
visDiffPandaInCytoscape(diffRes_vis_ALL, condition_name = "High_Low_Agressive", network_name="HOS_U2OS_MG63_sumC1C4diff_400_Filtering.PANDA")
```



#### 4.3. Plot the highest differential networks betweeen High Agressive (HOS/U2OS) versus Low Agressive (MG63)
```{r}
## Plot differential networks betweeen HOS (High Agressive) versus MG63 (Low Agressive)
# Select the top differential edge weights betweeen HOS and MG63
diffRes_HOS_MG63 <- pandaDiffEdges(HOS_vis, MG63_vis, threshold=0.8, condition_name="High_Agressive")
## Plot differential networks betweeen U2OS (High Agressive) versus MG63 (Low Agressive)
diffRes_U2OS_MG63 <- pandaDiffEdges(U2OS_vis, MG63_vis, threshold=0.8, condition_name="High_Agressive")

# Number of differential edges is:
nrow(diffRes_HOS_MG63) # 125104
nrow(diffRes_U2OS_MG63) # 127677

#  Apply the style to the network. Purple edges indicate higher edge weight in the defined "condition_name" parameter (HOS in our example), and green edges indicate higher edge weight in the other condition (MG63 in our example).
# createDiffPandaStyle(style_name="Diff.PandaStyle", condition_name="High_Agressive")

# Select the top differential edges higher in HOS to plot in Cytoscape
n=200 # number of edges to select from each condition
diffResHOS <- diffRes_HOS_MG63[diffRes_HOS_MG63$High_Agressive=="T",]
diffResHOS <- diffResHOS[order(diffResHOS$Score,decreasing=TRUE),][1:n,]
# Select the top differential edges higher in MG63 to plot in Cytoscape
diffResMG63_1 <- diffRes_HOS_MG63[diffRes_HOS_MG63$High_Agressive=="F",]
diffResMG63_1 <- diffResMG63_1[order(diffResMG63_1$Score,decreasing=TRUE),][1:n,]
# Combine top differential edges in HOS and MG63 to plot in Cytoscape
diffRes_vis_1 <- rbind(diffResHOS, diffResMG63_1)
# Plot the network
visDiffPandaInCytoscape(diffRes_vis_1, condition_name = "High_Agressive", network_name="HOS_MG63_sumC1C4diff_200_Filtering.PANDA")


# Select the top differential edges higher in HOS to plot in Cytoscape
n=200 # number of edges to select from each condition
diffResU2OS <- diffRes_U2OS_MG63[diffRes_U2OS_MG63$High_Agressive=="T",]
diffResU2OS <- diffResU2OS[order(diffResU2OS$Score,decreasing=TRUE),][1:n,]
# Select the top differential edges higher in MG63 to plot in Cytoscape
diffResMG63_2 <- diffRes_U2OS_MG63[diffRes_U2OS_MG63$High_Agressive=="F",]
diffResMG63_2 <- diffResMG63_2[order(diffResMG63_2$Score,decreasing=TRUE),][1:n,]
# Combine top differential edges in HOS and MG63 to plot in Cytoscape
diffRes_vis_2 <- rbind(diffResU2OS, diffResMG63_2)
# Plot the network
visDiffPandaInCytoscape(diffRes_vis_2, condition_name = "High_Agressive", network_name="U2OS_MG63_sumC1C4diff_200_Filtering.PANDA")

```



####  Legend
```{r}
pdf(file.path(plots_dir, "HOS_U2OS_legend_only.pdf"), width = 3, height = 2)
par(mar = c(0,0,0,0)); plot.new()
legend("center",
       legend = c("HOS", "MG63"),
       col    = c("#92278f",       "#60bf71"),
       lty = 1, lwd = 3, bty = "o", cex = 0.9)
dev.off()

pdf(file.path(plots_dir, "U2OS_MG63_legend_only.pdf"), width = 3, height = 2)
par(mar = c(0,0,0,0)); plot.new()
legend("center",
       legend = c("U2OS", "MG63"),
       col    = c( "#92278f",        "#60bf71"),
       lty = 1, lwd = 3, bty = "o", cex = 0.9)
dev.off()
```




##References
http://netbooks.networkmedicine.org/user/thangnx1012/notebooks/netZooR/panda_gtex_tutorial_server.ipynb

