---
title: "RNAseq_Osteosarcoma"
author: "Xuan-Thang Nguyen"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---
# Summary 
This script implements a simple overview of Mass Spectrometry (MS) data analysis with **DESeq2 package**. 
The data including 3 cell lines (HOS, U2OS and MG63), with 4 biological/technical replicates, that make total 12 single run

The main steps of analysis:  
- Exploring data before the differential expression analysis  
- Importing **RNAseq HTS counts** into **DESeq2 package** list    
- Performing the analysis with **DESeq2**  
- Exporting and exploring the results  
The results are exported into a text file and visualized using ggplot heatmaps and volcano-plots. 

Detailed tutorial for **DESeq2 package** package could be accessed here:  
http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html
*[r4rnaseq-workshop.Rmd] (https://github.com/nih-cfde/training-rstudio-binder/blob/data/GTEx/r4rnaseq-workshop.Rmd) was last modified 14 March, 2022.*


1. R package and data preparation

After installing packages, we need to load the functions and tools we want to use from the package with the `library()` command. Let's load the following packages: `ggplot2, tidyr, dplyr, readr, and tibble` by copying and pasting or typing these commands into your new script.

```{r setup, include=FALSE,  warning=F}
library("RColorBrewer")
# library("pals")
library("pheatmap")
library("ggrepel")
library("ggplot2")
library("tidyr")
library("dplyr")
library("tibble")
library("cowplot")
library("scales")
library("forcats")
library("stringr")
library("DESeq2")
library("openxlsx")
library("clusterProfiler")
library("ComplexHeatmap")
library("circlize")

options(width=999)

# Set directory paths
#/Users/xuantn/Library/CloudStorage/OneDrive-UniversitetetiOslo/Bioinformatics/LPS_WGS_Longread/Circos_Plot
# Load required package
library(rprojroot)
# Define paths
root_dir <- rprojroot::find_root(rprojroot::has_dir("OS_RNAseq_Emily"))
analysis_dir <- file.path(root_dir, "OS_RNAseq_Emily")
plots_dir <- file.path(root_dir, "OS_RNAseq_Emily", "Results_FINAL_RAW")

# Create directories if they don't exist
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir, recursive = TRUE)
}
if (!dir.exists(analysis_dir)) {
  dir.create(analysis_dir, recursive = TRUE)
}

```

2. Overview of RNAseq data
Data can be imported using packages from base R or the tidyverse. To begin with, `read.csv()` replaces spaces and dashes periods in column names, and it also preserves row.names. On the other hand, we will use DESeqDataSetFromHTSeqCount function from DESeq2 to import data from HTSeq count, which means we may do not need to be imported single file with `read.table()` or  `read_tsv()`.
2.1. Design experiment and input data

```{r import-1, appendix = TRUE}
# Import HTSeq file folder
directory <- file.path(analysis_dir, 'RNAseq_RAW') 
# Collect sample from HTseq file
sample_files <- grep("_pergene_counts.txt", list.files(directory), value=TRUE)
sample_files

# Generate experimental design
experiment_design <- data.frame(sample = gsub("_pergene_counts.txt", "\\1", sample_files))
experiment_design$sample_files <- sample_files
experiment_design$cell <- gsub("_[0-9]_pergene_counts.txt", "\\1", sample_files)
experiment_design$replicate <- sapply(strsplit(experiment_design$sample, split = "_"), "[", 2)
experiment_design$conditions <- ifelse(experiment_design$cell == "MG63",  "Low_Aggressive", "High_Aggressive")
# Prints the sample / condition list
head(experiment_design)
# is.factor(experiment_design$cell)
# relevel the factor as comparision status: R will choose a reference level for factors based on alphabetical order. you can either explicitly tell results which comparison to make using the contrast argument after DESeq2 analysis
# experiment_design$cell <- factor(experiment_design$cell, levels = c("HOS","MG63", "U2OS"))
experiment_design$cell <- relevel(factor(experiment_design$cell), ref = "MG63")
is.factor(experiment_design$cell)
experiment_design$conditions <- relevel(factor(experiment_design$conditions), ref = "Low_Aggressive")
is.factor(experiment_design$conditions) 

# Import Transcription factor list
TFs_list <- read.xlsx(file.path(analysis_dir, "TFs_list_Final.xlsx"))
dim(TFs_list)
# Human Epifactor gene list: https://epifactors.autosome.org/description
Epifactor_list <- read.xlsx(file.path(analysis_dir, "Epifactor_Histone_list.xlsx"))
dim(Epifactor_list)


# 2.2. SummarizedExperiment input and filtering samples

## Build a DESeq2DataSet with condition with HTseq
ddsHTSeq <- DESeqDataSetFromHTSeqCount(sampleTable = experiment_design,
                                       directory = directory,
                                       design= ~ cell ) # + conditions cell
# Look at DEseq2 dataset
# ddsHTSeq
dim(ddsHTSeq)


# Pre-filtering by removing the lower expression data
# keep <- rowSums(counts(ddsHTSeq)) >= 10 # remove gene with low/not expression level detection
keep <- rowSums(counts(ddsHTSeq) >= 10) >= 4 # at least 4 samples (smallestGroupSize) with a count of 10 or higher
table(keep)
# Filtering
ddsHTSeq <- ddsHTSeq[keep,] 
rm(keep)

# Check what different of experiment between fitering and raw as half of gene names with low expression were removed (46k versus 23k genes)
# ddsHTSeq
dim(ddsHTSeq)

## Differential expression analysis
OS_dds <- DESeq(ddsHTSeq)
# How the data look like
OS_dds
# Calculate the size factor of DESeq data
sizeFactors(OS_dds)
# Check the results
resultsNames(OS_dds) # compare with relevel factor MG63:"Intercept"         "cell_HOS_vs_MG63"  "cell_U2OS_vs_MG63"

# save(OS_dds, ddsHTSeq, TFs_list, Epifactor_list, experiment_design, file = file.path(plots_dir, "DESeq2DataSet_cell.rda"))
# load(file = file.path(plots_dir, "DESeq2DataSet_cell.rda"))

```

# 2.3. Visualization 
```{r, appendix = TRUE}
load(file = file.path(plots_dir, "DESeq2DataSet_cell.rda"))

## Perform a rlog transformation on count data (log2 scale) to assume a normal distribution 
# the VST and the rlog are to remove the dependence of the variance on the mean, particularly 
# the high variance of the logarithm of count data when the mean is low to remove experiment-wide trend. 
vsd <- vst(ddsHTSeq, blind=FALSE)
rld <- rlog(ddsHTSeq, blind=FALSE)
head(assay(vsd), 3)
# Data transformations and visualization
df <- bind_rows(as_tibble(log2(counts(OS_dds, normalized=TRUE)[, 1:2]+1)) %>% mutate(transformation = "log2(x + 1)"),
                as_tibble(assay(vsd)[, 1:2]) %>% mutate(transformation = "vst"),
                as_tibble(assay(rld)[, 1:2]) %>% mutate(transformation = "rlog"))
colnames(df)[1:2] <- c("x", "y")  
lvls <- c("log2(x + 1)", "vst", "rlog")
df$transformation <- factor(df$transformation, levels=lvls)
# Plot the transformation
ggplot(df, aes(x = x, y = y)) + geom_hex(bins = 80) +
  coord_fixed() + facet_grid( . ~ transformation)
ggsave(filename=file.path(plots_dir, "DESeq_df_transformation.pdf"))

## Plot sample correlation heatmap
# Computes pairwise correlations between experiment_design based on gene expression
Sample_correlation <- cor(as.matrix(assay(rld)))  
# Plot the sample correlation heatmap
# pdf(filename="Results_HighLow/DESeq_sampleCorr_HM.pdf", units = 'in', width = 8, height = 8, res = 250)
# Set color palette for figures
colors <- colorRampPalette(c("white", "darkgreen"))(5)
# Computes pairwise correlations between experiment_design based on gene expression
Sample_correlation <- cor(as.matrix(assay(rld)))  
# Sample correlation heatmap

pdf(file=file.path(plots_dir,"DESeq_sampleCorr_HM.pdf"), height = 4.5, width = 5)
pheatmap(Sample_correlation,
         col=colors,
         main="Sample Correlations")
dev.off()

## Plot distance between samples
# Set color palette for figures
colors <- colorRampPalette(rev(brewer.pal(9, "Blues")) )(255)
# To ensure we have a roughly equal contribution from all genes, we use it on the VST data
sampleDists <- dist(t(assay(vsd)))
# Sample distance heatmap
sampleDistMatrix <- as.matrix(sampleDists)
# rownames(sampleDistMatrix) <- paste(vsd$cell, vsd$replicate, sep="-")
# colnames(sampleDistMatrix) <- NULL
# heatmap of distance between samples
pdf(file=file.path(plots_dir, "DESeq_sampleDist_HM.pdf"))
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists, 
         clustering_distance_cols=sampleDists,
         col=colors,
         main="Sample to Sample Distances")
dev.off()

## Principal Component Analysis plot of the samples
# Separates experiment_design based on variation between sample's gene expression greater variation will affect separation to a greater degree
# plotted(vsd, intgroup=c("cell", "replicate"))
data <- plotPCA(vsd, intgroup=c("cell"), returnData=TRUE)
percentVar <- round(100 * attr(data, "percentVar"))
# plot PCA with ggplot2 package
ggplot(data, aes(PC1, PC2, color=cell, shape = replicate)) +
  geom_point(size=3.5) +
#  geom_text_repel(aes(label = paste0(cell, "_rep", replicate)),
#                  box.padding = 0.4, point.padding = 0.5, 
#                  min.segment.length = 0,) +
  scale_colour_manual(values = c("MG63"="#56B4E9", "U2OS"="#F66239", "HOS"="#8332FF")) +
  theme_bw() +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("")
ggsave(filename=file.path(plots_dir, "DESeq_PCA.pdf"), width = 4.5, height = 3.5, dpi = 600)

## Write normalized counts to file
normalized.counts <- as.data.frame(counts(OS_dds, normalized=TRUE ))
head(rownames(normalized.counts))
# rownames(normalized.counts) <- sub("\\.[0-9]*", "", rownames(normalized.counts))
head(normalized.counts)
# Dont know why write.xlsx isn't work with rowName=T
normalized.counts1 <- cbind("GeneName"=rownames(normalized.counts), normalized.counts) 
# write.xlsx(normalized.counts1, filename=file.path(plots_dir,"RNAseq_normalize.count_Osteosarcoma_Filtering.xlsx"))
```

# 3. Differential expression analysis and Visualization
```{r, appendix = TRUE}
resultsNames(OS_dds) # compare with relevel factor MG63: "Intercept"         "cell_HOS_vs_MG63"  "cell_U2OS_vs_MG63"

# Results HOS and MG63
results_HOS_MG63 <- results(OS_dds,contrast=c("cell", "HOS", "MG63"),alpha = 0.05)
results_HOS_U2OS <- results(OS_dds,contrast=c("cell", "HOS", "U2OS"),alpha = 0.05)
results_U2OS_MG63 <- results(OS_dds,contrast=c("cell", "U2OS", "MG63"),alpha = 0.05)
colnames(results_HOS_MG63)
# Filtering and Transform data results of High and Low expression comparative

results_HOS_MG63_df <- results_HOS_MG63 %>% data.frame() %>% dplyr::filter(padj <= 0.05, abs(log2FoldChange) >= 1) %>% rownames_to_column()
results_HOS_U2OS_df <- results_HOS_U2OS %>% data.frame() %>% filter(padj <= 0.05, abs(log2FoldChange) >= 1) %>% rownames_to_column()
results_U2OS_MG63_df <- results_U2OS_MG63 %>% data.frame() %>% dplyr::filter(padj <= 0.05, abs(log2FoldChange) >= 1) %>% rownames_to_column()
# how the dimension of result look like
dim(results_HOS_MG63_df)

## Heatmap plot
gene_plot_heatmap <- c(results_HOS_MG63_df$rowname,  results_U2OS_MG63_df$rowname) %>% unique() #, results_HOS_U2OS_df$rowname

# Heatmap data with only DEG in DESeq2
heatmap_data <- vsd[gene_plot_heatmap , ] %>%  assay()
# Calculate the Z-score matrix
# Zscore_matrix <- t(scale(t(heatmap_data), center = TRUE, scale = TRUE))
# Transform Expressiondata to Z-score dataframe
Zscore_matrix <- t(apply(heatmap_data, 1, function(x){ (x - mean(x)) / sd(x)} ))
colnames(Zscore_matrix)


library(ComplexHeatmap)
library(circlize)
library(grid)
library(dplyr)
# --------------------------
# Step 1: Reorder columns
# --------------------------
fixed_order <- c("MG63_1", "MG63_2", "MG63_3", "MG63_4",
                 "U2OS_1", "U2OS_2", "U2OS_3", "U2OS_4",
                 "HOS_1", "HOS_2", "HOS_3", "HOS_4")
# Reorder Zscore_matrix columns to match
Zscore_matrix <- Zscore_matrix[, fixed_order]
# --------------------------
# Step 2: Ensure sample_group matches Zscore_matrix columns
# --------------------------
sample_group <- experiment_design %>%
  mutate(sample_id = paste0(cell, "_", replicate)) %>%
  dplyr::filter(sample_id %in% fixed_order) %>%
  column_to_rownames("sample_id") %>%
  dplyr::select(cell, conditions, replicate) %>%
  .[fixed_order, ]
# Check if rownames match the heatmap matrix columns
stopifnot(all(rownames(sample_group) == colnames(Zscore_matrix)))

# --------------------------
# Step 3: hierarchical alignment
# --------------------------
# The hierarchical tree indication if there are outlayers among the samples.
sample_cluster <- hclust(as.dist(1-cor(Zscore_matrix, method="spearman")), 
                         method="complete") # Clusters columns by Spearman correlation.
# The distance data is used to plot the dendrogram.
# plot(as.dendrogram(sample_cluster, method="average"))

# Create the annotation of gene clusters
gene_cluster_hc = hclust(dist(Zscore_matrix))
## Extract gene from heatmap cluster of interest
gene_cluster_list <- as.data.frame(cutree(gene_cluster_hc, k = 5))
colnames(gene_cluster_list) <- "Gene_cluster"

# --------------------------
# Step 4: Create Annotations
# --------------------------
annotation_colour <- list(
  cell = c("HOS" = "#0073e6", "U2OS" = "skyblue", "MG63"= "white"),
  Gene_cluster = c("1"= "darkgreen", "2" = "#5ba300", "3" = "#00bf7d", "4"=  "#0073e6", "5" = "#5928ed", "6" = "pink"))

selected_genes <- c("RUNX2", "TEAD4", "FOXO1", "JUN", "SMAD7", "VIM", "SNAI2", "ZEB1", "ZEB2", "TGFB1", "TGFB2", "TWIST1", "SOX4", "SOX9", "HMGA2")
idx <- which(rownames(Zscore_matrix) %in% selected_genes)
labs <- rownames(Zscore_matrix)[idx]
right_row_anno <- rowAnnotation(genes = anno_mark(at = idx, labels = labs, labels_gp = grid::gpar(fontsize = 8)))


top_col_anno <- HeatmapAnnotation(df = sample_group,
                                  col = annotation_colour)

left_row_anno <- rowAnnotation(Gene_cluster = gene_cluster_list$Gene_cluster,
                               col = list(Gene_cluster = annotation_colour$Gene_cluster))
# --------------------------
# Step 5: Define Heatmap color palette
# --------------------------
heatmap_colors <- colorRampPalette(c("skyblue", "white", "purple"))(50)

# --------------------------
# Step 6: Draw heatmap
# --------------------------
pdf(file = file.path(plots_dir, "DESeq2_DEG_heatmap_OS_gene_name.pdf"), width = 7, height = 6)
Heatmap(
  Zscore_matrix,
  name = "Z-score",
  col = heatmap_colors,
  top_annotation = top_col_anno,
  left_annotation = left_row_anno,
  right_annotation = right_row_anno,
  show_row_names = FALSE,
  show_column_names = TRUE,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  row_split = 5,
  column_title = "RNA-seq_DEGs_OS")
dev.off()

```



```{r}
# loop to extract genes for each cluster.
# gene_cluster_list <- lapply(rownames(Zscore_matrix), function(i){
#  out <- data.frame(GeneID = rownames(Zscore_matrix[[i]], ),
#                    Cluster = paste0("cluster", sep="_", i),
#                    stringsAsFactors = FALSE)
#  return(out)}) %>%  #pipe (forward) the output 'out' to the function rbind to create 'gene_cluster'
#  do.call(rbind, .)
gene_cluster <- tibble::rownames_to_column(gene_cluster_list, "GeneID")
# Add cluster before cluster number
gene_cluster$Gene_cluster <- paste0("cluster_", gene_cluster$Gene_cluster)
head(gene_cluster)
# write the gene cluster to excell file
write.xlsx(gene_cluster, file=file.path(plots_dir,"gene_cluster_DEG_HOSMG63_U2OSMG63_HOSU2OS.xlsx"))

dplyr::count(gene_cluster, Gene_cluster)
## 0) Packages
# install.packages("BiocManager")
# BiocManager::install(c("clusterProfiler","org.Hs.eg.db","enrichplot"))
library(dplyr)
library(tidyr)
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(ggplot2)

## 1) Clean cluster labels and make an ordered factor
gene_cluster <- gene_cluster %>%
  mutate(Gene_cluster = gsub("^cluster[_ ]?", "cluster_", Gene_cluster),
         Gene_cluster = factor(Gene_cluster,
                               levels = paste0("cluster_", sort(unique(
                                 as.integer(gsub("\\D", "", Gene_cluster))
                               )))))

## 2) Split genes by cluster -> list
genes_by_cluster <- split(gene_cluster$GeneID, gene_cluster$Gene_cluster)

## 3) Define universe = all unique genes you tested (recommended)
universe_symbols <- unique(gene_cluster$GeneID)

## 4) Map SYMBOL -> ENTREZ (once)
sym2entrez_all <- bitr(universe_symbols,
                       fromType = "SYMBOL", toType = "ENTREZID",
                       OrgDb = org.Hs.eg.db)

## Helper: run GO BP for one cluster
run_go_cluster <- function(sym_vec, cluster_name) {
  # map symbols to Entrez
  df_map <- sym2entrez_all[sym2entrez_all$SYMBOL %in% sym_vec, , drop = FALSE]
  if (nrow(df_map) < 10) return(NULL)  # need enough genes

  # universe in ENTREZ
  uni_entrez <- sym2entrez_all$ENTREZID

  ego <- enrichGO(gene          = unique(df_map$ENTREZID),
                  universe      = unique(uni_entrez),
                  OrgDb         = org.Hs.eg.db,
                  keyType       = "ENTREZID",
                  ont           = "BP",
                  pAdjustMethod = "BH",
                  pvalueCutoff  = 0.05,
                  qvalueCutoff  = 0.2,
                  readable      = TRUE)
  if (is.null(ego) || nrow(as.data.frame(ego)) == 0) return(NULL)

  # add cluster label
  res <- as.data.frame(ego)
  res$Cluster <- cluster_name
  res
}

## 5) Run GO for all clusters
go_list <- lapply(names(genes_by_cluster), function(cl){
  run_go_cluster(genes_by_cluster[[cl]], cl)
})
names(go_list) <- names(genes_by_cluster)

## 6) Combine results
go_all <- bind_rows(go_list)
# Quick peek: top terms per cluster
go_top <- go_all %>%
  group_by(Cluster) %>%
  slice_min(order_by = p.adjust, n = 10, with_ties = FALSE) %>%
  ungroup()

## 7) Plot: dotplot faceted by cluster (top 10 each)
# clusterProfiler’s dotplot expects an enrichResult; we can facet using ggplot on go_top:
ggplot(go_top,
       aes(x = Cluster, y = reorder(Description, -p.adjust),
           size = Count, color = p.adjust)) +
  geom_point() +
  scale_color_continuous(low = "red", high = "blue", trans = "log10", name = "adj. p") +
  #facet_wrap(~ Cluster, scales = "free_y") +
  labs(x = NULL, y = NULL, title = "GO BP enrichment per cluster") +
  theme_minimal(base_size = 12)

## 8) Save tables
# Full
# write.csv(go_all, "GO_BP_all_clusters.csv", row.names = FALSE)
# Top per cluster
# write.csv(go_top, "GO_BP_top10_per_cluster.csv", row.names = FALSE)
```


```{r}
# https://scienceparkstudygroup.github.io/rna-seq-lesson/08-cluster-analysis/index.html
# alternatively it is also posible to set the number of cluster you want.
library(reshape2)
library(ggplot2)
library(gridExtra)

# Get genes from cluster 4
cluster_number <- 5
clusters <- cutree(gene_cluster_hc, k = cluster_number)
clust4 <- t(Zscore_matrix[clusters == cluster_number, ])

# Melt the data frame into long format for plotting
clust4Molten <- melt(clust4)
colnames(clust4Molten) <- c('sample', 'gene', 'value')

# Order sample names
clust4Molten$sample <- factor(clust4Molten$sample,
                              levels = c("MG63_1", "MG63_2", "MG63_3", "MG63_4",
                                         "U2OS_1", "U2OS_2", "U2OS_3", "U2OS_4",
                                         "HOS_1", "HOS_2", "HOS_3", "HOS_4"))
# Calculate the average expression for each sample
avg_expression <- clust4Molten %>%
  group_by(sample) %>%
  summarise(avg_value = mean(value, na.rm = TRUE))

# Plot the data for all genes in cluster 4 with baseline average expression
ggplot(clust4Molten, aes(x = sample, y = value)) +
  geom_line(aes(group = gene), color = "grey") +
  geom_point(aes(group = gene), color = "grey") +
  # Add baseline for average expression
  geom_line(data = avg_expression, aes(x = sample, y = avg_value, group = 1), color = "blue", size = 1.2) +
  xlab("Samples") +
  ylab("Expression") +
  labs(title = paste0("Cluster 4 consisting of ", ncol(clust4), " genes"), color = "Score") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
  # theme_classic()

##### All cluster in a picture
plot_list <- lapply(1:cluster_number, function(i) {
  clust <- t(Zscore_matrix[clusters == i, ])
  # Melt the data frame into long format for plotting
  clustMolten <- melt(clust)
  colnames(clustMolten) <- c('sample', 'gene', 'value')
  # Order sample names
  clustMolten$sample <- factor(clustMolten$sample, 
                               levels = c("MG63_1", "MG63_2", "MG63_3", "MG63_4",
                                          "U2OS_1", "U2OS_2", "U2OS_3", "U2OS_4",
                                          "HOS_1", "HOS_2", "HOS_3", "HOS_4"))
  # Calculate the average expression for each sample
  avg_expression <- clustMolten %>%
    group_by(sample) %>%
    summarise(avg_value = mean(value, na.rm = TRUE))
  # Create the plot
  p <- ggplot(clustMolten, aes(x = sample, y = value)) +
    geom_line(aes(group = gene), color = "grey") +
    geom_point(aes(group = gene), color = "grey") +
    geom_line(data = avg_expression, aes(x = sample, y = avg_value, group = 1), color = "blue", size = 0.5) +
    xlab("Samples") +
    ylab("Expression") +
    labs(title = paste0("Cluster ", i, " consisting of ", ncol(clust), " genes")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
    #theme_classic()
  return(p)
})

# Arrange the plots
grid.arrange(grobs = plot_list, ncol = 2)


```


```{r, Volcano Plot}
# Load required libraries
library(ggplot2)
library(dplyr)
library(ggrepel)

# ===== Step 1: Classify expression status =====
prepare_data <- function(res) {
  res_df <- as.data.frame(res) %>%
    rownames_to_column("Gene") %>%
    mutate(Expression = case_when(padj < 0.05 & log2FoldChange > 1 ~ "Upregulated",
                                  padj < 0.05 & log2FoldChange < -1 ~ "Downregulated",
                                  TRUE ~ "Not_significant"))
  return(res_df)
}

# ===== Step 2: Create label dataframe for specific genes =====
prepare_labels <- function(df, genes_of_interest) {
  df %>%
    dplyr::filter(Gene %in% genes_of_interest) %>%
    mutate(label_color = case_when(Expression == "Upregulated" ~ "red",
                                   Expression == "Downregulated" ~ "blue",
                                   TRUE ~ "grey"))
}

# ===== Step 3: Plot volcano with custom gene labels & DEG count =====
plot_volcano <- function(res_df, label_df, title) {
  # Keep only LABEL significant genes
  label_df <- label_df %>% dplyr::filter(Expression != "Not_significant")
  # Count DEGs
  num_up <- sum(res_df$Expression == "Upregulated")
  num_down <- sum(res_df$Expression == "Downregulated")
  subtitle_text <- paste0("Downregulated genes: ", num_down,
                          " | Upregulated genes: ", num_up )
  # Plot
  ggplot(res_df, aes(x = log2FoldChange, y = -log10(padj))) +
    geom_point(aes(color = Expression), size = 1.5, alpha = 0.6) +
    scale_color_manual(values = c("Upregulated" = "red", "Downregulated" = "blue")) +
    geom_label_repel(
      data = label_df,
      aes(x = log2FoldChange, y = -log10(padj), label = Gene, color = label_color),
      fill = "white",
      color = label_df$label_color,
      size = 3,
      box.padding = 0.5,
      point.padding = 0.6,
      min.segment.length = 0,
      segment.color = "gray50",
      segment.size = 0.4,
      max.overlaps = Inf,
      force = 5,
      force_pull = 0.3,
      direction = "both",
      nudge_y = 0.05,
      show.legend = FALSE,
      inherit.aes = FALSE) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
    theme_minimal() +
    labs(title = title,
         subtitle = subtitle_text,
         x = "log2(Fold Change)",
         y = "-log10(padj)",
         color = "Expression") +
    theme(plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5))
}


```


```{r, appendix = TRUE}
# ---- Step 4: Run the pipeline ----
# Set genes to label
genes_to_label <- c("NFIA", "ZEB1", "FOXC1", "NFIX", "JUND", "JUNB", "RUNX2",
                    "KDM5D", "IGF2BP1", "NFIB", "DTX4", "LAMA1", "KRT8")   # <- edit this list

# Check the results
resultsNames(OS_dds) # compare "condition_DDLPS_vs_CONTROL" "condition_WDLPS_vs_CONTROL"

# Results HOS and MG63
results_HOS_MG63 <- results(OS_dds,contrast = c("cell", "HOS", "MG63"), alpha = 0.05)
results_U2OS_MG63 <- results(OS_dds,contrast = c("cell", "U2OS", "MG63"), alpha = 0.05)
results_HOS_U2OS <- results(OS_dds,contrast = c("cell", "HOS", "U2OS"), alpha = 0.05)
# Prepare each result
res1 <- prepare_data(results_HOS_MG63)
labels1 <- prepare_labels(res1, genes_to_label)
write.xlsx(res1, file = file.path(plots_dir, "RNAseq_results_HOS_MG63_p0.05.xlsx"))

res2 <- prepare_data(results_U2OS_MG63)
labels2 <- prepare_labels(res2, genes_to_label)
write.xlsx(res2, file = file.path(plots_dir, "RNAseq_results_U2OS_MG63_p0.05.xlsx"))

# Plot and save to PDF
pdf(file = file.path(plots_dir, "volcano_HOS_vs_MG63..pdf"), width = 7, height = 5)
print(plot_volcano(res1, labels1, "MG63 vs HOS"))
dev.off()

pdf(file = file.path(plots_dir, "volcano_U2OS_vs_MG63..pdf"), width = 7, height = 5)
print(plot_volcano(res2, labels2, "MG63 vs U2OS"))
dev.off()



genes_to_label <- c("ZEB2", "FOXC1", "TGFBR2", "JUND", "LTBP2", "RUNX2",
                    "IGFBP4", "IGF2BP1", "IGFBP2", "DTX4", "CCND1", "KRT81")   # <- edit this list
res3 <- prepare_data(results_HOS_U2OS)
labels3 <- prepare_labels(res3, genes_to_label)
write.xlsx(res3, file = file.path(plots_dir, "RNAseq_results_HOS_U2OS_p0.05.xlsx"))
pdf(file = file.path(plots_dir, "volcano_HOS_vs_U2OS.pdf"), width = 7, height = 5)
print(plot_volcano(res3, labels3, "HOS vs U2OS"))
dev.off()
```

```{r}
library(dplyr)
# Prepare DESeq2 results → gene lists
res_HOS <- as.data.frame(results_HOS_MG63) %>%
  rownames_to_column("gene") %>%
  dplyr::filter(padj < 0.05, abs(log2FoldChange) > 1)

res_U2OS <- as.data.frame(results_U2OS_MG63) %>%
  rownames_to_column("gene") %>%
  dplyr::filter(padj < 0.05, abs(log2FoldChange) > 1)

# Convert gene symbols → Entrez IDs
library(clusterProfiler)
library(org.Hs.eg.db)
genes_HOS <- bitr(res_HOS$gene,
                  fromType = "SYMBOL",
                  toType   = "ENTREZID",
                  OrgDb    = org.Hs.eg.db)$ENTREZID

genes_U2OS <- bitr(res_U2OS$gene,
                  fromType = "SYMBOL",
                  toType   = "ENTREZID",
                  OrgDb    = org.Hs.eg.db)$ENTREZID

# Run GO enrichment (standard parameters!)
ego_HOS <- enrichGO(gene          = genes_HOS,
                    OrgDb         = org.Hs.eg.db,
                    ont           = "BP",      # BP / MF / CC
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.05,
                    qvalueCutoff  = 0.05,
                    readable      = TRUE)

ego_U2OS <- enrichGO(gene         = genes_U2OS,
                    OrgDb         = org.Hs.eg.db,
                    ont           = "BP",      # BP / MF / CC
                    pAdjustMethod = "BH",
                    pvalueCutoff  = 0.05,
                    qvalueCutoff  = 0.05,
                    readable      = TRUE)
# Combine GO results into one dataframe
go_HOS_df <- as.data.frame(ego_HOS) %>%
  mutate(Comparison = "HOS vs MG63")
go_U2OS_df <- as.data.frame(ego_U2OS) %>%
  mutate(Comparison = "U2OS vs MG63")
go_all <- bind_rows(go_HOS_df, go_U2OS_df)
# Select only common term of GO
common_terms <- intersect(go_HOS_df$Description, go_U2OS_df$Description)
go_common <- go_all %>%
  dplyr::filter(Description %in% common_terms)
# Rank the term
go_common_ranked <- go_common %>%
  group_by(Description) %>%
  summarise(
    mean_padj = mean(p.adjust, na.rm = TRUE),
    sum_logp  = sum(-log10(p.adjust), na.rm = TRUE),
    .groups = "drop") %>%
  arrange(mean_padj)
# Select top 10 GO term
top10_terms <- go_common_ranked %>%
  slice_head(n = 10) %>%
  pull(Description)
# filter the GO term plot
go_plot <- go_common %>%
  dplyr::filter(Description %in% top10_terms)
# Plot
library(ggplot2)
go_plot$Comparison <- factor(
  go_plot$Comparison,
  levels = c("HOS vs MG63", "U2OS vs MG63"))
# ggplot 
p <- ggplot(go_plot,
            aes(x = Count,
                y = reorder(Description, Count),
                color = p.adjust,
                size = Count)) +
  geom_point() +
  facet_wrap(~ Comparison, scales = "free_x") +
  scale_color_gradient(low = "blue", high = "purple") +
  labs(x = "Gene count",
       y = "GO Biological Process",
       color = "Adjusted p-value",
       size = "Gene count") +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"),
        axis.text.y = element_text(size = 9))
# ggplot to hide one text graph
p + theme(axis.text.y.right = element_blank(),
          axis.ticks.y.right = element_blank())
ggsave(filename = file.path(plots_dir, "GOBP_RNAseq_HOS_MG63_and_U2OS_MG63.pdf"), width = 7, height = 4)

```





